### 2020-03-02 submission/manuscript.Rmd
Initial commit

---
bibliography: references.bib
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: header.tex
csl: plos-computational-biology.csl
fontsize: 11pt
geometry: margin=1.0in
---

```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
options(tidyverse.quiet = TRUE)

library(tidyverse)
library(knitr)

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("message" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
  print(x)
  if(is.list(x)){
    x <- unlist(x)
  }
  if(is.numeric(x)){
    if(abs(x - round(x)) < .Machine$double.eps^0.5){
      paste(format(x,big.mark=',', digits=0, scientific=FALSE))
    } else {
      paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
    }
  } else {
      paste(x)
  }
}
knitr::knit_hooks$set(inline=inline_hook)
```

\vspace{10mm}

# Ten Simple Rules for running a Code Club to increasing computational skills among biologists

\vspace{35mm}

Patrick D. Schloss${^1}$${^\dagger}$

\vspace{40mm}

$\dagger$ To whom correspondence should be addressed: \href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109

\vspace{35mm}

### Ten Simple Rules

\newpage
\linenumbers

## Introduction

* We are awash in data
* Traditional undergraduate and graduate training is insufficient to develop data analysis skills needed to analyze one's own data
* Common to find exceptional bench scientists who are inexperienced at analyzing large datasets
* As new people join a research group there is a mix of coding experience in the group

### Our solution
* This is not unlike the situation researchers find themselves in regards to an ever growing literature
* Developed a Code Club to parallel the traditional Journal Club
* Have experimented with different strategies
* Overall goal is to enable intra-research group learning surrounding topics in data analysis

### Initial format
* Code review - project someone's script, go line by line through the code with each person in room stating the logic behind the code
* Pros
  - Emphasizes importance of readability
  - People get to see real, smelly code
* Cons
  - Can be hard to integrate logic across lines of code
  - Although beginners could contribute, experts dominated conversation
  - Could become adversarial, especially if the presenter didn't articulate their goal for the review
  - Not easy for presenter to take information back to modify their own code

### Current format
1. Help!
   - Clearly state problem the presenter would like to solve
   - Distribute code to pairs of lab members
   - Pairs work on problem
   - Diversity of solutions shared back to group and presenter
   - Ex: make code DRY, convert to tidyverse framework, go through issue tracker fixing code
2. Tutorial
   - Introduce new package or way of doing things
   - Start with brief presentation of package, workflow, book chapter
   - Break into pairs to work through exercises
   - Diversity of solutions shared back to group
   - Ex: base vs ggplot, implement make/snakemake workflows

## Rule 1: Reciprocate respect (Group)

1.	Sign up
*	Establish a code of conduct
4.	Check your attitude
	a.	Be humble
	b.	Be willing to incorporate feedback
	c.	Ask clarifying questions
	b.	Remember that this isn’t a competition

7.	Reciprocate respect
	a.	Facilitation
		i.	Lead the discussion
		ii.	 Have an action plan: Focus the conversation
		iii.	Balance expertise in groups
		iv.	Attempt to purposefully engage early learners
		v.	Avoid letting (senior) individuals dominate
	b.	Attitude:
		i.	Both:
			1.	 Be humble
			2.	 Be willing to incorporate feedback
			3.	 Ask clarifying questions and note why you need the clarification
		ii.	Participants:
			1.	Remember that this isn’t a competition
			2.	Focus on the speaker’s goals
			3.	Avoid distractions and engage with the presenter
			4.	Allow others to contribute
			5.	Give feedback & coding suggestions at a novice explanation level
			6.	Give compliments

## Rule 2: Set specific goals (Presenter)

Preparation
5.	Lead the discussion
	a.	Have an action plan
	c.	Focus the conversation
1.	Set specific goals for each code review - Try to be as explicit and clear as possible for what you are asking the participants to do
	a.	I have tried to be (what felt to me) painfully clear but I still was falling short making it clear to some people.
	b.	People will always tend to notice or ask about things that are not the focus of the exercise, being specific about the goal may help. Also setting up a simplified scenario can be helpful (although requires much more prep time)

## Rule 3: Keep it simple (Presenter)

2.	Practice exercise
	a.	When in doubt try it out! Run it by someone prior to lab meeting to make sure your request/goal is understandable
	b.	Running on different systems will inevitably run into issues. Send out basic script ahead of time to try the dependencies/functions of focus
3.	Create a *minimal working example* - Working directly with your own raw code can be challenging
	a.	It is likely filled with code specific to you/your project
	b.	It can take a lot of time to explain details and working of code
	c.	Code can have multiple steps that are needed to run to get to script focus
	d.	Using a whole script opens up the risk of getting off topic

5.	Keep It Simple Silly (KISS)
	a.	There is never enough time
		i.	I typically have multiple activities planned, but rarely get through them all
		ii.	Modularize the activity so that each can be its own and if accomplished you are able to move onto the next
	b.	One activity per code review is typically sufficient
		i.	This will take up most of the time
			1.	Intro - 10 min
			2.	Think/pair - 30 min
			3.	Share in group - 10 min
			4.	Debrief - 5 min
		ii.	I have not been successful in trying to do more than that in different ways because it is going to be taking time away from one of the other steps

## Rule 4: Give group time to prepare (Presenter)

3.	Communicate with the group in advance so they can prepare
	a.	Provide code chunk and necessary data
	b.	Provide instructions for appropriate dependencies
	c.	Provide use case (where available/appropriate)
	d.	Specify goals & expectations

## Rule 5: Prepare in advance (Audience)

1.	Prepare
	a.	Review the code in advance
	b.	Download datasets, install packages
	b.	Google what you don’t understand

Read chapter, paper, etc.

## Rule 6: Don't give busy work (Presenter)

2.	For Help! format, select your code
	a.	That you are willing to receive and incorporate feedback on
	b.	That is digestible in an hour (<100 lines)
	c.	Use “vanilla” R to check for dependencies on a “fresh” environment
	d.	Prepare a small example dataset

## Rule 7: Include all levels of learner (Presenter)

* Ask questions that are appropriate for different levels
* Be purposeful about how you pair people off
  - Use think/pair/share
  - Senior/Beginner
  - Senior/senior and Beginner/beginner
  - Self-selected
  - Navigator/driver with a switch midway through

## Rule 8: Engage with the material (Audience)

3.	Participate
	a.	Focus on the speaker’s goals
	b.	Avoid distractions and engage with the presenter
	c.	Allow others to contribute
	d.	Ask clarifying questions and note why you need the clarification
	e.	Give feedback & coding suggestions at a novice explanation level
	f.	Give compliments

## Rule 9: Let the material change you (Audience)

- Reflection
  - Use what you've learned to refactor your own code / improve new code

## Rule 10: Speak up (Group)

* Speak up!
  - Topics
		- check interest level of potential topics
		- check group is at sufficient skill level needed
		- need for specific skills/methods in group
		- May need to circle back to "beginner" stuff with natural cycle of research group development
	- What works and doesn't

## Conclusion

## Acknowledgements

\newpage

## References

<div id="refs"></div>

https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004526

\newpage

**Figure 1. XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX.**  XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX.

### 2020-03-02 submission/manuscript.tex
Initial commit

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font

%\usepackage{mathpazo} % Palatino font
%\renewcommand*\familydefault{\rmdefault} % Use the roman version of the font

\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\vspace{10mm}

\hypertarget{ten-simple-rules-for-running-a-code-club-to-increasing-computational-skills-among-biologists}{%
\section{Ten Simple Rules for running a Code Club to increasing
computational skills among
biologists}\label{ten-simple-rules-for-running-a-code-club-to-increasing-computational-skills-among-biologists}}

\vspace{35mm}

Patrick D. Schloss\({^1}\)\({^\dagger}\)

\vspace{40mm}

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

\(1\) Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI 48109

\vspace{35mm}

\hypertarget{ten-simple-rules}{%
\subsubsection{Ten Simple Rules}\label{ten-simple-rules}}

\newpage
\linenumbers

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\begin{itemize}
\tightlist
\item
  We are awash in data
\item
  Traditional undergraduate and graduate training is insufficient to
  develop data analysis skills needed to analyze one's own data
\item
  Common to find exceptional bench scientists who are inexperienced at
  analyzing large datasets
\item
  As new people join a research group there is a mix of coding
  experience in the group
\end{itemize}

\hypertarget{our-solution}{%
\subsubsection{Our solution}\label{our-solution}}

\begin{itemize}
\tightlist
\item
  This is not unlike the situation researchers find themselves in
  regards to an ever growing literature
\item
  Developed a Code Club to parallel the traditional Journal Club
\item
  Have experimented with different strategies
\item
  Overall goal is to enable intra-research group learning surrounding
  topics in data analysis
\end{itemize}

\hypertarget{initial-format}{%
\subsubsection{Initial format}\label{initial-format}}

\begin{itemize}
\tightlist
\item
  Code review - project someone's script, go line by line through the
  code with each person in room stating the logic behind the code
\item
  Pros

  \begin{itemize}
  \tightlist
  \item
    Emphasizes importance of readability
  \item
    People get to see real, smelly code
  \end{itemize}
\item
  Cons

  \begin{itemize}
  \tightlist
  \item
    Can be hard to integrate logic across lines of code
  \item
    Although beginners could contribute, experts dominated conversation
  \item
    Could become adversarial, especially if the presenter didn't
    articulate their goal for the review
  \item
    Not easy for presenter to take information back to modify their own
    code
  \end{itemize}
\end{itemize}

\hypertarget{current-format}{%
\subsubsection{Current format}\label{current-format}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Help!

  \begin{itemize}
  \tightlist
  \item
    Clearly state problem the presenter would like to solve
  \item
    Distribute code to pairs of lab members
  \item
    Pairs work on problem
  \item
    Diversity of solutions shared back to group and presenter
  \item
    Ex: make code DRY, convert to tidyverse framework, go through issue
    tracker fixing code
  \end{itemize}
\item
  Tutorial

  \begin{itemize}
  \tightlist
  \item
    Introduce new package or way of doing things
  \item
    Start with brief presentation of package, workflow, book chapter
  \item
    Break into pairs to work through exercises
  \item
    Diversity of solutions shared back to group
  \item
    Ex: base vs ggplot, implement make/snakemake workflows
  \end{itemize}
\end{enumerate}

\hypertarget{rule-1-reciprocate-respect-group}{%
\subsection{Rule 1: Reciprocate respect
(Group)}\label{rule-1-reciprocate-respect-group}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Sign up
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Establish a code of conduct
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  Check your attitude

  \begin{enumerate}
  \def\labelenumii{\alph{enumii}.}
  \tightlist
  \item
    Be humble
  \item
    Be willing to incorporate feedback
  \item
    Ask clarifying questions
  \item
    Remember that this isn't a competition
  \end{enumerate}
\item
  Reciprocate respect

  \begin{enumerate}
  \def\labelenumii{\alph{enumii}.}
  \tightlist
  \item
    Facilitation

    \begin{enumerate}
    \def\labelenumiii{\roman{enumiii}.}
    \tightlist
    \item
      Lead the discussion
    \item
      Have an action plan: Focus the conversation
    \item
      Balance expertise in groups
    \item
      Attempt to purposefully engage early learners
    \item
      Avoid letting (senior) individuals dominate
    \end{enumerate}
  \item
    Attitude:

    \begin{enumerate}
    \def\labelenumiii{\roman{enumiii}.}
    \tightlist
    \item
      Both:

      \begin{enumerate}
      \def\labelenumiv{\arabic{enumiv}.}
      \tightlist
      \item
        Be humble
      \item
        Be willing to incorporate feedback
      \item
        Ask clarifying questions and note why you need the clarification
      \end{enumerate}
    \item
      Participants:

      \begin{enumerate}
      \def\labelenumiv{\arabic{enumiv}.}
      \tightlist
      \item
        Remember that this isn't a competition
      \item
        Focus on the speaker's goals
      \item
        Avoid distractions and engage with the presenter
      \item
        Allow others to contribute
      \item
        Give feedback \& coding suggestions at a novice explanation
        level
      \item
        Give compliments
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}
\end{enumerate}

\hypertarget{rule-2-set-specific-goals-presenter}{%
\subsection{Rule 2: Set specific goals
(Presenter)}\label{rule-2-set-specific-goals-presenter}}

Preparation 5. Lead the discussion a. Have an action plan c.~Focus the
conversation 1. Set specific goals for each code review - Try to be as
explicit and clear as possible for what you are asking the participants
to do a. I have tried to be (what felt to me) painfully clear but I
still was falling short making it clear to some people. b. People will
always tend to notice or ask about things that are not the focus of the
exercise, being specific about the goal may help. Also setting up a
simplified scenario can be helpful (although requires much more prep
time)

\hypertarget{rule-3-keep-it-simple-presenter}{%
\subsection{Rule 3: Keep it simple
(Presenter)}\label{rule-3-keep-it-simple-presenter}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Practice exercise

  \begin{enumerate}
  \def\labelenumii{\alph{enumii}.}
  \tightlist
  \item
    When in doubt try it out! Run it by someone prior to lab meeting to
    make sure your request/goal is understandable
  \item
    Running on different systems will inevitably run into issues. Send
    out basic script ahead of time to try the dependencies/functions of
    focus
  \end{enumerate}
\item
  Create a \emph{minimal working example} - Working directly with your
  own raw code can be challenging

  \begin{enumerate}
  \def\labelenumii{\alph{enumii}.}
  \tightlist
  \item
    It is likely filled with code specific to you/your project
  \item
    It can take a lot of time to explain details and working of code
  \item
    Code can have multiple steps that are needed to run to get to script
    focus
  \item
    Using a whole script opens up the risk of getting off topic
  \end{enumerate}
\item
  Keep It Simple Silly (KISS)

  \begin{enumerate}
  \def\labelenumii{\alph{enumii}.}
  \tightlist
  \item
    There is never enough time

    \begin{enumerate}
    \def\labelenumiii{\roman{enumiii}.}
    \tightlist
    \item
      I typically have multiple activities planned, but rarely get
      through them all
    \item
      Modularize the activity so that each can be its own and if
      accomplished you are able to move onto the next
    \end{enumerate}
  \item
    One activity per code review is typically sufficient

    \begin{enumerate}
    \def\labelenumiii{\roman{enumiii}.}
    \tightlist
    \item
      This will take up most of the time

      \begin{enumerate}
      \def\labelenumiv{\arabic{enumiv}.}
      \tightlist
      \item
        Intro - 10 min
      \item
        Think/pair - 30 min
      \item
        Share in group - 10 min
      \item
        Debrief - 5 min
      \end{enumerate}
    \item
      I have not been successful in trying to do more than that in
      different ways because it is going to be taking time away from one
      of the other steps
    \end{enumerate}
  \end{enumerate}
\end{enumerate}

\hypertarget{rule-4-give-group-time-to-prepare-presenter}{%
\subsection{Rule 4: Give group time to prepare
(Presenter)}\label{rule-4-give-group-time-to-prepare-presenter}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Communicate with the group in advance so they can prepare

  \begin{enumerate}
  \def\labelenumii{\alph{enumii}.}
  \tightlist
  \item
    Provide code chunk and necessary data
  \item
    Provide instructions for appropriate dependencies
  \item
    Provide use case (where available/appropriate)
  \item
    Specify goals \& expectations
  \end{enumerate}
\end{enumerate}

\hypertarget{rule-5-prepare-in-advance-audience}{%
\subsection{Rule 5: Prepare in advance
(Audience)}\label{rule-5-prepare-in-advance-audience}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Prepare

  \begin{enumerate}
  \def\labelenumii{\alph{enumii}.}
  \tightlist
  \item
    Review the code in advance
  \item
    Download datasets, install packages
  \item
    Google what you don't understand
  \end{enumerate}
\end{enumerate}

Read chapter, paper, etc.

\hypertarget{rule-6-dont-give-busy-work-presenter}{%
\subsection{Rule 6: Don't give busy work
(Presenter)}\label{rule-6-dont-give-busy-work-presenter}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  For Help! format, select your code

  \begin{enumerate}
  \def\labelenumii{\alph{enumii}.}
  \tightlist
  \item
    That you are willing to receive and incorporate feedback on
  \item
    That is digestible in an hour (\textless100 lines)
  \item
    Use ``vanilla'' R to check for dependencies on a ``fresh''
    environment
  \item
    Prepare a small example dataset
  \end{enumerate}
\end{enumerate}

\hypertarget{rule-7-include-all-levels-of-learner-presenter}{%
\subsection{Rule 7: Include all levels of learner
(Presenter)}\label{rule-7-include-all-levels-of-learner-presenter}}

\begin{itemize}
\tightlist
\item
  Ask questions that are appropriate for different levels
\item
  Be purposeful about how you pair people off

  \begin{itemize}
  \tightlist
  \item
    Use think/pair/share
  \item
    Senior/Beginner
  \item
    Senior/senior and Beginner/beginner
  \item
    Self-selected
  \item
    Navigator/driver with a switch midway through
  \end{itemize}
\end{itemize}

\hypertarget{rule-8-engage-with-the-material-audience}{%
\subsection{Rule 8: Engage with the material
(Audience)}\label{rule-8-engage-with-the-material-audience}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Participate

  \begin{enumerate}
  \def\labelenumii{\alph{enumii}.}
  \tightlist
  \item
    Focus on the speaker's goals
  \item
    Avoid distractions and engage with the presenter
  \item
    Allow others to contribute
  \item
    Ask clarifying questions and note why you need the clarification
  \item
    Give feedback \& coding suggestions at a novice explanation level
  \item
    Give compliments
  \end{enumerate}
\end{enumerate}

\hypertarget{rule-9-let-the-material-change-you-audience}{%
\subsection{Rule 9: Let the material change you
(Audience)}\label{rule-9-let-the-material-change-you-audience}}

\begin{itemize}
\tightlist
\item
  Reflection

  \begin{itemize}
  \tightlist
  \item
    Use what you've learned to refactor your own code / improve new code
  \end{itemize}
\end{itemize}

\hypertarget{rule-10-speak-up-group}{%
\subsection{Rule 10: Speak up (Group)}\label{rule-10-speak-up-group}}

\begin{itemize}
\tightlist
\item
  Speak up!

  \begin{itemize}
  \tightlist
  \item
    Topics - check interest level of potential topics - check group is
    at sufficient skill level needed - need for specific skills/methods
    in group - May need to circle back to ``beginner'' stuff with
    natural cycle of research group development

    \begin{itemize}
    \tightlist
    \item
      What works and doesn't
    \end{itemize}
  \end{itemize}
\end{itemize}

\hypertarget{conclusion}{%
\subsection{Conclusion}\label{conclusion}}

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

\newpage

\hypertarget{references}{%
\subsection{References}\label{references}}

\hypertarget{refs}{}

\url{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004526}

\newpage

\textbf{Figure 1. XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX
XXXXX XXX.} XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX
XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX
XXXXX XXX XXXXX XXX XXXXX XXX.

\end{document}

### 2020-03-02 submission/manuscript.Rmd
Initial draft

@@
 ```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
-options(tidyverse.quiet = TRUE)
+library(knitr)
 
-library(tidyverse)
-library(knitr)
+options(tinytex.verbose = TRUE)
 
@@
 
-# Ten Simple Rules for running a Code Club to increasing computational skills among biologists
+# Ten Simple Rules for running a Code Club for increasing computational skills among biologists
 
 \vspace{35mm}
+
+*Author names here, order TBD*
 
@@
 
-* We are awash in data
-* Traditional undergraduate and graduate training is insufficient to develop data analysis skills needed to analyze one's own data
-* Common to find exceptional bench scientists who are inexperienced at analyzing large datasets
-* As new people join a research group there is a mix of coding experience in the group
+For most biologists, the ability to generate data has outpaced the ability to analyze those data.
+High throughput data comes to us from DNA and RNA sequencing, flow cytometry, metabolomics, molecular screens, and more.
+Although some accept the approach of compartmentalizing the ability to generate data and analyze data, we have found scientists feel empowered when they have the ability to both ask and answer their own biological questions.
+Yet, the standard undergraduate and graduate training in the biological sciences is insufficient to develop the data analysis skills needed to analyze one's data.
+In our experience performing microbiome research, it is more common to find exceptional bench scientists who are inexperienced at analyzing large datasets than to find the reverse.
+Of course this raises a challenge: How do we train bench scientists to analyze datasets to answer biological questions?
 
-### Our solution
-* This is not unlike the situation researchers find themselves in regards to an ever growing literature
-* Developed a Code Club to parallel the traditional Journal Club
-* Have experimented with different strategies
-* Overall goal is to enable intra-research group learning surrounding topics in data analysis
+The ever-growing ability to generate data and constant feeling of helplessness in analyzing it is analogous to the struggles we also face with engaging the voluminous scientific literature.
+Within our research group we have leveraged this similarity to address the challenges of training bench scientists to engage in reproducible data analysis practices.
+Our group and others hold a regular Journal Club to stay current on the literature and model to junior scientists how to engage that literature [@Lonsdale2016].
+Over the past 4 years we have experimented with using a Code Club to introduce new software packages, solve specific problems, and evaluate each other's code.
 
-### Initial format
-* Code review - project someone's script, go line by line through the code with each person in room stating the logic behind the code
-* Pros
-  - Emphasizes importance of readability
-  - People get to see real, smelly code
-* Cons
-  - Can be hard to integrate logic across lines of code
-  - Although beginners could contribute, experts dominated conversation
-  - Could become adversarial, especially if the presenter didn't articulate their goal for the review
-  - Not easy for presenter to take information back to modify their own code
+Initially, the Code Club was created as a way to review code that was being used in different projects.
+We would project someone's script onto a screen and go line by line through the code with each person in the room stating the logic behind the code.
+This approach emphasized the importance of readability and it gave beginners the opportunity to see the real, "smelly" code of more experienced people in the group.
+Unfortunately, the format only allowed us to review a small fraction of a project's code base and it became difficult to integrate the programmer's logic across their full project.
+We quickly realized that although beginners could contribute to the discussion, the more senior group members dominated conversation.
+This frequently created an adversarial environment, which led to the presenter not offering to present again.
+From this experience we learned that it is critical for the presenter to clearly articulate their goals for the Code Club and that the participants needed to engage with the code in that context.
 
-### Current format
-1. Help!
-   - Clearly state problem the presenter would like to solve
-   - Distribute code to pairs of lab members
-   - Pairs work on problem
-   - Diversity of solutions shared back to group and presenter
-   - Ex: make code DRY, convert to tidyverse framework, go through issue tracker fixing code
-2. Tutorial
-   - Introduce new package or way of doing things
-   - Start with brief presentation of package, workflow, book chapter
-   - Break into pairs to work through exercises
-   - Diversity of solutions shared back to group
-   - Ex: base vs ggplot, implement make/snakemake workflows
+Our Code Club sessions are each an hour long and are held roughly every other week as part of a weekly two hour lab meeting.
+Presenters voluntarily sign up to lead a session.
+The current format of our Code Club generally takes one of two approaches.
+The first is a more humane version of a group code review.
+The presenter will clearly state the problem they would like to solve, break the participants into smaller groups, and then ask each group to solve the problem.
+For example, someone may have an R script where a chunk of code is repeated throughout the script.
+The challenge for the session would be to convert the chunk into a function and call it throughout the script to make it "DRY" (i.e.
+Don't Repeat Yourself)[@Wilson2014].
+The presenter leaves the session with several partial or working solutions to their problem and everyone relearns the importance of writing DRY code.
+The second is a tutorial approach.
+The presenter introduces a new package or way of doing things.
+They then give the group an activity that forces them to replace their current approach with a new approach.
+For example, one of the more memorable activities was giving participants raw data and a finished plot and asking two pairs of participants to generate the plot using base R and asking the other two pairs to generate the plot with ggplot.
+The stipulation was that base R users had to build the plots in ggplot and ggplot users had to build the plot in base R.
+Each Code Club ends with an opportunity to report back to the presenter and larger group.
+Our Code Clubs typically have 7 to 10 participants.
+Because of the "think-pair-share" approach that is baked into each session it could likely be scaled to larger groups [@Kothiyal2014].
 
-## Rule 1: Reciprocate respect (Group)
+We continue to experiment with approaches to running Code Club.
+Although any individual Code Club session may be be more experimental than others, on the whole they have been a critical tool to providing the much needed training to use reproducible practices for analyzing data.
+Examples of successful code club topics are described in Table 1.
+These Ten Rules summarize what we have learned as presenters and participants.
 
-1.	Sign up
-*	Establish a code of conduct
-4.	Check your attitude
-	a.	Be humble
-	b.	Be willing to incorporate feedback
-	c.	Ask clarifying questions
-	b.	Remember that this isn’t a competition
+## Rule 1: Reciprocate respect
 
-7.	Reciprocate respect
-	a.	Facilitation
-		i.	Lead the discussion
-		ii.	 Have an action plan: Focus the conversation
-		iii.	Balance expertise in groups
-		iv.	Attempt to purposefully engage early learners
-		v.	Avoid letting (senior) individuals dominate
-	b.	Attitude:
-		i.	Both:
-			1.	 Be humble
-			2.	 Be willing to incorporate feedback
-			3.	 Ask clarifying questions and note why you need the clarification
-		ii.	Participants:
-			1.	Remember that this isn’t a competition
-			2.	Focus on the speaker’s goals
-			3.	Avoid distractions and engage with the presenter
-			4.	Allow others to contribute
-			5.	Give feedback & coding suggestions at a novice explanation level
-			6.	Give compliments
+It is critical that the presenter and participants respect each other and that there be an individual (e.g.
+the lab director) that enforces a code of conduct.
+Everyone in the session needs to have the humility to acknowledge that they still have more to learn about any topic.
+We have found that many problems are avoided when the presenter takes charge of the session by having a clear lesson plan, thoughtfully creates groups, and gives encouragement.
+Similarly, participants foster a positive environment by remembering that the task is not a competition, focusing on the presenter's goals, allowing their partner to contribute, asking clarifying questions when appropriate, and avoiding distractions (e.g.
+email, social media).
+Learning to program is challenging and too often it can lead to a toxic environment.
+All parties in a Code Club are responsible for preventing this toxicity.
 
-## Rule 2: Set specific goals (Presenter)
+## Rule 2: Set specific goals
 
-Preparation
-5.	Lead the discussion
-	a.	Have an action plan
-	c.	Focus the conversation
-1.	Set specific goals for each code review - Try to be as explicit and clear as possible for what you are asking the participants to do
-	a.	I have tried to be (what felt to me) painfully clear but I still was falling short making it clear to some people.
-	b.	People will always tend to notice or ask about things that are not the focus of the exercise, being specific about the goal may help.
-Also setting up a simplified scenario can be helpful (although requires much more prep time)
+In our early Code Clubs we noticed that if the presenter did not clearly state their goals for the session, it would often led to frustration for the presenter and participants.
+If the presenter shared their own code, did they want us to focus on their coding style or do they want help with incorporating a new package into their workflow?
+Participants will always notice or ask about things that are not the focus of the exercise.
+Be specific about the goal and bring them back to the task if they are distracted.
+Where possible, create a simplified scenario (i.e.
+minimal, reproducible example [@StackOverflow_MREE]) can be helpful that removes distractions.
+The presenter should make sure that the simplified example works and behaves the way they expect before coming to the Code Club.
+Beyond the content, participants should know what they can be expected to have completed by the end of the session.
 
-## Rule 3: Keep it simple (Presenter)
+## Rule 3: Keep it simple
 
-2.	Practice exercise
-	a.	When in doubt try it out! Run it by someone prior to lab meeting to make sure your request/goal is understandable
-	b.	Running on different systems will inevitably run into issues. Send out basic script ahead of time to try the dependencies/functions of focus
-3.	Create a *minimal working example* - Working directly with your own raw code can be challenging
-	a.	It is likely filled with code specific to you/your project
-	b.	It can take a lot of time to explain details and working of code
-	c.	Code can have multiple steps that are needed to run to get to script focus
-	d.	Using a whole script opens up the risk of getting off topic
+Our Code Club needs to fit within an hour slot.
+The presenter should anticipate enough time for an introduction and brief instruction, time for participants to engage the material, and time for everyone to report back.
+A typical schedule for Code Review is 10 minutes of introduction and instruction, 30 minutes of paired programming, 5 minutes to get groups to wrap up, and 10 minutes to report back to the group.
+Remember that a task that you can accomplish in 10 minutes may take a more junior member of your group 30 minutes.
+The goal of the Code Club should be to help that junior member to learn the topic and to reinforce the topic for the more senior person and help them see the concept in a new context.
+One possibility would be to create "stretch goals" for those that complete the task more quickly than everyone else.
+We once had a presenter try to teach the group some basic Julia syntax.
+By the time people installed the interpreter, the time was up.
+Alternatively, a session where participants were asked to make code DRY gave them multiple stopping points where they could feel successful (e.g.
+commenting code, creating function, implementing function, refactoring function).
 
-5.	Keep It Simple Silly (KISS)
-	a.	There is never enough time
-		i.	I typically have multiple activities planned, but rarely get through them all
-		ii.	Modularize the activity so that each can be its own and if accomplished you are able to move onto the next
-	b.	One activity per code review is typically sufficient
-		i.	This will take up most of the time
-			1.	Intro - 10 min
-			2.	Think/pair - 30 min
-			3.	Share in group - 10 min
-			4.	Debrief - 5 min
-		ii.	I have not been successful in trying to do more than that in different ways because it is going to be taking time away from one of the other steps
+## Rule 4: Give participants time to prepare
 
-## Rule 4: Give group time to prepare (Presenter)
+Similar to a Journal Club, the presenter should give participants a few days to prepare for the Code Club.
+Be sure to tell the participants what you will be covering.
+Provide the them with instructions on how to install dependencies, download data, and get the initial code.
+Make sure that the materials work together as intended.
+Considering the compressed schedule, asking participants to download materials beforehand is helpful.
+We have found that using separate GitHub repositories for each session can be a helpful way to get information to participants.
+Of course, using git and GitHub to engage in collaborative coding should probably be its own Code Club before using the tools more broadly with the group.
+An easier entry point is to have groups post their code to a Slack channel that we use to announce materials for our meetings.
 
-3.	Communicate with the group in advance so they can prepare
-	a.	Provide code chunk and necessary data
-	b.	Provide instructions for appropriate dependencies
-	c.	Provide use case (where available/appropriate)
-	d.	Specify goals & expectations
+## Rule 5: Prepare in advance to maximize your participation
 
-## Rule 5: Prepare in advance (Audience)
+It is the rare scientist that can fully participate in a Journal Club discussion about a paper that they have not read.
+Coming to Code Club without installing a dependency is like going to Journal Club and asking a simple question about the paper.
+It shows a lack of respect and preparation.
+Following Rule 4, review the code in advance, download the datasets, install the necessary packages, and perhaps read up on the topic.
+If your Code Club is reading a paper or chapters in a data science book be sure to read them before the session (e.g.
+[@Wickham_2016; @Schloss_2018; @Noble2009]).
+Consider how you could use the material for your own work.
 
-1.	Prepare
-	a.	Review the code in advance
-	b.	Download datasets, install packages
-	b.	Google what you don’t understand
+## Rule 6: Don't give your participants busy work
 
-Read chapter, paper, etc.
+Participants want to learn something because it will be useful to them or because they think it will help the presenter to improve their code.
+Presenters should do their best to satisfy those desires.
+It would not make much sense to hold a Code Club on downloading stock market data if the group was not interested in those data.
+At the same time, participants should not be asked to help the presenter improve their code if the presenter has no intention of incorporating the suggestions.
+It would be useful to keep a list of packages or tasks that members of the group want to learn about.
+That way if a presenter is having a hard time coming up with ideas, they can take something from the list and present it.
 
-## Rule 6: Don't give busy work (Presenter)
+## Rule 7: Include all levels of participants
 
-2.	For Help! format, select your code
-	a.	That you are willing to receive and incorporate feedback on
-	b.	That is digestible in an hour (<100 lines)
-	c.	Use “vanilla” R to check for dependencies on a “fresh” environment
-	d.	Prepare a small example dataset
+A significant challenge to presenting at Code Club is to pick topics and activities that appeal to a critical mass of the participants.
+For example, if there has recently been a large turn over in group membership, it can be useful to go back to the basics and cover commonly used tools.
+We have identified several strategies to overcome the challenges of having varying skill levels.
+As an alternative to letting participants form their own pairs, the presenter can be purposeful in how they form pairs.
+Depending on the goals of the session, forming pairs between participants that have similar or different skills can be effective.
+If pairs are formed where partners have similar skill levels, then the presenter needs to design activities that are appropriate to different levels of participants.
+We have found that a good activity for beginners is to have them comment code, which forces them to dissect a code chunk and understand what each line does.
+An advantage of forming pairs with disparate skill levels is that all of the pairs can be given the same task leading to a diversity of methods to achieve the same result.
+This approach to pairing also helps graft new members into the group that have emerging programming skills.
+Regardless of how partners are selected, asking the pairs to identify a navigator and driver is helpful [@Bockeler_2020].
+The driver types at the computer while the navigator tells them what to type.
+Midway through the activity the presenter can tell the partners to switch roles.
+Intentionally forming pairs can also be a mechanism to engineer interactions by breaking up cliques, avoiding potentially toxic combinations of members, or pairing reliable role models with new members of the group.
 
-## Rule 7: Include all levels of learner (Presenter)
+## Rule 8: Participate
 
-* Ask questions that are appropriate for different levels
-* Be purposeful about how you pair people off
-  - Use think/pair/share
-  - Senior/Beginner
-  - Senior/senior and Beginner/beginner
-  - Self-selected
-  - Navigator/driver with a switch midway through
+Having an open laptop on the table and permission to use it can feel like an invitation to get distracted by other work, emails, and browsing the internet.
+Fight that urge and focus on the presenter's goals.
+Allow your partner to contribute.
+Speak up for yourself and force your partner to let you contribute.
+Even if the material seems too advanced for you, programming languages like R and Python are generally expressive and should allow you to engage with the logic, even if the syntax is too hard.
+Frankly, understanding the logic of when to use one modeling approach over another is more important than how to use the modeling approach.
+If you understand the "why", the "how" will generally quickly follow.
+If you are a more experienced participant, give your feedback and coding suggestions at a level that seeks to inform everyone.
+Regardless of your skill level, your partner and the presenter are putting themselves at risk by revealing what they do or do not know.
+Encourage them and show your gratitude for helping you learn something new.
 
-## Rule 8: Engage with the material (Audience)
+## Rule 9: Let the material change you
 
-3.	Participate
-	a.	Focus on the speaker’s goals
-	b.	Avoid distractions and engage with the presenter
-	c.	Allow others to contribute
-	d.	Ask clarifying questions and note why you need the clarification
-	e.	Give feedback & coding suggestions at a novice explanation level
-	f.	Give compliments
+Part of the humility required to participate in a Code Club is both acknowledging your incomplete training and that learning something new is possible.
+Assume that the presenter has followed the other rules and that they think you should be using what they presented.
+After the Code Club, try to incorporate that material into your new code or by refactoring your old code.
+By practicing the material in a different context you will learn the material better.
+You may identify concepts that you are unsure of, which gives you another opportunity to ask the presenter for help.
 
-## Rule 9: Let the material change you (Audience)
+## Rule 10: Experiment!
 
-- Reflection
-  - Use what you've learned to refactor your own code / improve new code
-
-## Rule 10: Speak up (Group)
-
-* Speak up!
-  - Topics
-		- check interest level of potential topics
-		- check group is at sufficient skill level needed
-		- need for specific skills/methods in group
-		- May need to circle back to "beginner" stuff with natural cycle of research group development
-	- What works and doesn't
+The selection of content and structure for each Code Club is determined by the presenter making the process democratic and distributed.
+If someone thinks something is worth learning and wants to teach it, they have that power.
+If they want to experiment with a different format, they are free to try it out.
+Members of the Code Club need to feel like they have the power to shape the direction of the group.
+If members are following these Rules, they will naturally reflect on the skills and interests of the other members in the group.
+For example, there is always turn over in a research group and thus it is important to make sure that there are opportunities to go back to things people consider "simple" to teach the content to new people or provide a refresher to people that perhaps have been at the lab bench more than at the computer recently.
+The group should also feel free to experiment with the format.
+If we try a different format, then it is useful to give everyone an opportunity to debrief on what they did or did not like with the new approach.
 
@@
 
+The most important rules are the first and last.
+Members of the Code Club need to feel comfortable with other members of the group and sufficiently empowered to try something new or to ask for help.
+Aside from growing in our programming skills, we have noticed two other benefits that point to the efforts of creating a positive culture.
+First, we have intentionally interviewed postdoc candidates on the days we have held Code Club.
+Although it is important to make it clear that they are not being assessed on their coding skills, it is useful to see how the candidate interacts with other members of the research group in a relatively low stakes environment.
+At the same time, the candidate can learn about the culture of the research group.
+Second, members of other research groups where they are isolated in their desire to grow in these skills have integrated themselves into our Code Club.
+This speaks to the broader need for Code Club and the likelihood that it would work when expanded to a larger group of individuals that are not necessarily in the same research group.
+Ultimately, our Code Club has helped researchers feel more comfortable seeking out their colleagues outside of Code Club sessions for help in improving their code.
+
 ## Acknowledgements
+
+All co-authors have been participants in the Code Clubs run as part of PDS's lab meetings.
+They have had a critical role in shaping the evolution of the sessions.
+The order of co-authors was determined by...
 
@@
 
-https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004526
-
 \newpage
 
-**Figure 1.
-XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX.**  XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX.
+**Table 1. Examples of successful Code Club topics.**
+
+| **Title**             | **Description**                                    |
+| ----------------- | ---------------------------------------------- |
+| base vs. ggplot   | Given input data and a figure, recreate the figure using R's base graphics or ggplot syntax   |
+| Snakemake         | Given a bash script that contains an analysis pipeline, convert it to a series of recipes (can also be done with GNU Make) |
+| DRYing code       | Given script with repeated code, create functions to remove repetition   |
+| mothur and Vegan	| Given a pairwise community dissimilarity matrix, compare communities using the `adonis` function in the Vegan R package  |
+| tidy data         | Given a wide-formatted data table, convert it to a long, tidy-formatted data table using tools from R's tidyverse  |
+| GitFlow           | Participants file and claim an issue to add their name to a README file in a GitHub-hosted repository and file a pull request to complete the issue |
+| Integrating R and google docs   | Scrape a google docs workbook and clean the data to identify previous Code Club presenters |

### 2020-03-02 submission/manuscript.tex
Initial draft

@@
 \usepackage[margin=1.0in]{geometry}
+\usepackage{longtable,booktabs}
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx,grffile}
@@
 
-\hypertarget{ten-simple-rules-for-running-a-code-club-to-increasing-computational-skills-among-biologists}{%
-\section{Ten Simple Rules for running a Code Club to increasing
+\hypertarget{ten-simple-rules-for-running-a-code-club-for-increasing-computational-skills-among-biologists}{%
+\section{Ten Simple Rules for running a Code Club for increasing
 computational skills among
-biologists}\label{ten-simple-rules-for-running-a-code-club-to-increasing-computational-skills-among-biologists}}
+biologists}\label{ten-simple-rules-for-running-a-code-club-for-increasing-computational-skills-among-biologists}}
 
 \vspace{35mm}
+
+\emph{Author names here, order TBD}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  We are awash in data
-\item
-  Traditional undergraduate and graduate training is insufficient to
-  develop data analysis skills needed to analyze one's own data
-\item
-  Common to find exceptional bench scientists who are inexperienced at
-  analyzing large datasets
-\item
-  As new people join a research group there is a mix of coding
-  experience in the group
-\end{itemize}
+For most biologists, the ability to generate data has outpaced the
+ability to analyze those data. High throughput data comes to us from DNA
+and RNA sequencing, flow cytometry, metabolomics, molecular screens, and
+more. Although some accept the approach of compartmentalizing the
+ability to generate data and analyze data, we have found scientists feel
+empowered when they have the ability to both ask and answer their own
+biological questions. Yet, the standard undergraduate and graduate
+training in the biological sciences is insufficient to develop the data
+analysis skills needed to analyze one's data. In our experience
+performing microbiome research, it is more common to find exceptional
+bench scientists who are inexperienced at analyzing large datasets than
+to find the reverse. Of course this raises a challenge: How do we train
+bench scientists to analyze datasets to answer biological questions?
 
-\hypertarget{our-solution}{%
-\subsubsection{Our solution}\label{our-solution}}
+The ever-growing ability to generate data and constant feeling of
+helplessness in analyzing it is analogous to the struggles we also face
+with engaging the voluminous scientific literature. Within our research
+group we have leveraged this similarity to address the challenges of
+training bench scientists to engage in reproducible data analysis
+practices. Our group and others hold a regular Journal Club to stay
+current on the literature and model to junior scientists how to engage
+that literature {[}1{]}. Over the past 4 years we have experimented with
+using a Code Club to introduce new software packages, solve specific
+problems, and evaluate each other's code.
 
-\begin{itemize}
-\tightlist
-\item
-  This is not unlike the situation researchers find themselves in
-  regards to an ever growing literature
-\item
-  Developed a Code Club to parallel the traditional Journal Club
-\item
-  Have experimented with different strategies
-\item
-  Overall goal is to enable intra-research group learning surrounding
-  topics in data analysis
-\end{itemize}
+Initially, the Code Club was created as a way to review code that was
+being used in different projects. We would project someone's script onto
+a screen and go line by line through the code with each person in the
+room stating the logic behind the code. This approach emphasized the
+importance of readability and it gave beginners the opportunity to see
+the real, ``smelly'' code of more experienced people in the group.
+Unfortunately, the format only allowed us to review a small fraction of
+a project's code base and it became difficult to integrate the
+programmer's logic across their full project. We quickly realized that
+although beginners could contribute to the discussion, the more senior
+group members dominated conversation. This frequently created an
+adversarial environment, which led to the presenter not offering to
+present again. From this experience we learned that it is critical for
+the presenter to clearly articulate their goals for the Code Club and
+that the participants needed to engage with the code in that context.
 
-\hypertarget{initial-format}{%
-\subsubsection{Initial format}\label{initial-format}}
+Our Code Club sessions are each an hour long and are held roughly every
+other week as part of a weekly two hour lab meeting. Presenters
+voluntarily sign up to lead a session. The current format of our Code
+Club generally takes one of two approaches. The first is a more humane
+version of a group code review. The presenter will clearly state the
+problem they would like to solve, break the participants into smaller
+groups, and then ask each group to solve the problem. For example,
+someone may have an R script where a chunk of code is repeated
+throughout the script. The challenge for the session would be to convert
+the chunk into a function and call it throughout the script to make it
+``DRY'' (i.e.~Don't Repeat Yourself){[}2{]}. The presenter leaves the
+session with several partial or working solutions to their problem and
+everyone relearns the importance of writing DRY code. The second is a
+tutorial approach. The presenter introduces a new package or way of
+doing things. They then give the group an activity that forces them to
+replace their current approach with a new approach. For example, one of
+the more memorable activities was giving participants raw data and a
+finished plot and asking two pairs of participants to generate the plot
+using base R and asking the other two pairs to generate the plot with
+ggplot. The stipulation was that base R users had to build the plots in
+ggplot and ggplot users had to build the plot in base R. Each Code Club
+ends with an opportunity to report back to the presenter and larger
+group. Our Code Clubs typically have 7 to 10 participants. Because of
+the ``think-pair-share'' approach that is baked into each session it
+could likely be scaled to larger groups {[}3{]}.
 
-\begin{itemize}
-\tightlist
-\item
-  Code review - project someone's script, go line by line through the
-  code with each person in room stating the logic behind the code
-\item
-  Pros
+We continue to experiment with approaches to running Code Club. Although
+any individual Code Club session may be be more experimental than
+others, on the whole they have been a critical tool to providing the
+much needed training to use reproducible practices for analyzing data.
+Examples of successful code club topics are described in Table 1. These
+Ten Rules summarize what we have learned as presenters and participants.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Emphasizes importance of readability
-  \item
-    People get to see real, smelly code
-  \end{itemize}
-\item
-  Cons
+\hypertarget{rule-1-reciprocate-respect}{%
+\subsection{Rule 1: Reciprocate
+respect}\label{rule-1-reciprocate-respect}}
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Can be hard to integrate logic across lines of code
-  \item
-    Although beginners could contribute, experts dominated conversation
-  \item
-    Could become adversarial, especially if the presenter didn't
-    articulate their goal for the review
-  \item
-    Not easy for presenter to take information back to modify their own
-    code
-  \end{itemize}
-\end{itemize}
+It is critical that the presenter and participants respect each other
+and that there be an individual (e.g.~the lab director) that enforces a
+code of conduct. Everyone in the session needs to have the humility to
+acknowledge that they still have more to learn about any topic. We have
+found that many problems are avoided when the presenter takes charge of
+the session by having a clear lesson plan, thoughtfully creates groups,
+and gives encouragement. Similarly, participants foster a positive
+environment by remembering that the task is not a competition, focusing
+on the presenter's goals, allowing their partner to contribute, asking
+clarifying questions when appropriate, and avoiding distractions
+(e.g.~email, social media). Learning to program is challenging and too
+often it can lead to a toxic environment. All parties in a Code Club are
+responsible for preventing this toxicity.
 
-\hypertarget{current-format}{%
-\subsubsection{Current format}\label{current-format}}
+\hypertarget{rule-2-set-specific-goals}{%
+\subsection{Rule 2: Set specific
+goals}\label{rule-2-set-specific-goals}}
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Help!
+In our early Code Clubs we noticed that if the presenter did not clearly
+state their goals for the session, it would often led to frustration for
+the presenter and participants. If the presenter shared their own code,
+did they want us to focus on their coding style or do they want help
+with incorporating a new package into their workflow? Participants will
+always notice or ask about things that are not the focus of the
+exercise. Be specific about the goal and bring them back to the task if
+they are distracted. Where possible, create a simplified scenario
+(i.e.~minimal, reproducible example {[}4{]}) can be helpful that removes
+distractions. The presenter should make sure that the simplified example
+works and behaves the way they expect before coming to the Code Club.
+Beyond the content, participants should know what they can be expected
+to have completed by the end of the session.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Clearly state problem the presenter would like to solve
-  \item
-    Distribute code to pairs of lab members
-  \item
-    Pairs work on problem
-  \item
-    Diversity of solutions shared back to group and presenter
-  \item
-    Ex: make code DRY, convert to tidyverse framework, go through issue
-    tracker fixing code
-  \end{itemize}
-\item
-  Tutorial
+\hypertarget{rule-3-keep-it-simple}{%
+\subsection{Rule 3: Keep it simple}\label{rule-3-keep-it-simple}}
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Introduce new package or way of doing things
-  \item
-    Start with brief presentation of package, workflow, book chapter
-  \item
-    Break into pairs to work through exercises
-  \item
-    Diversity of solutions shared back to group
-  \item
-    Ex: base vs ggplot, implement make/snakemake workflows
-  \end{itemize}
-\end{enumerate}
+Our Code Club needs to fit within an hour slot. The presenter should
+anticipate enough time for an introduction and brief instruction, time
+for participants to engage the material, and time for everyone to report
+back. A typical schedule for Code Review is 10 minutes of introduction
+and instruction, 30 minutes of paired programming, 5 minutes to get
+groups to wrap up, and 10 minutes to report back to the group. Remember
+that a task that you can accomplish in 10 minutes may take a more junior
+member of your group 30 minutes. The goal of the Code Club should be to
+help that junior member to learn the topic and to reinforce the topic
+for the more senior person and help them see the concept in a new
+context. One possibility would be to create ``stretch goals'' for those
+that complete the task more quickly than everyone else. We once had a
+presenter try to teach the group some basic Julia syntax. By the time
+people installed the interpreter, the time was up. Alternatively, a
+session where participants were asked to make code DRY gave them
+multiple stopping points where they could feel successful
+(e.g.~commenting code, creating function, implementing function,
+refactoring function).
 
-\hypertarget{rule-1-reciprocate-respect-group}{%
-\subsection{Rule 1: Reciprocate respect
-(Group)}\label{rule-1-reciprocate-respect-group}}
+\hypertarget{rule-4-give-participants-time-to-prepare}{%
+\subsection{Rule 4: Give participants time to
+prepare}\label{rule-4-give-participants-time-to-prepare}}
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Sign up
-\end{enumerate}
+Similar to a Journal Club, the presenter should give participants a few
+days to prepare for the Code Club. Be sure to tell the participants what
+you will be covering. Provide the them with instructions on how to
+install dependencies, download data, and get the initial code. Make sure
+that the materials work together as intended. Considering the compressed
+schedule, asking participants to download materials beforehand is
+helpful. We have found that using separate GitHub repositories for each
+session can be a helpful way to get information to participants. Of
+course, using git and GitHub to engage in collaborative coding should
+probably be its own Code Club before using the tools more broadly with
+the group. An easier entry point is to have groups post their code to a
+Slack channel that we use to announce materials for our meetings.
 
-\begin{itemize}
-\tightlist
-\item
-  Establish a code of conduct
-\end{itemize}
+\hypertarget{rule-5-prepare-in-advance-to-maximize-your-participation}{%
+\subsection{Rule 5: Prepare in advance to maximize your
+participation}\label{rule-5-prepare-in-advance-to-maximize-your-participation}}
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{3}
-\tightlist
-\item
-  Check your attitude
+It is the rare scientist that can fully participate in a Journal Club
+discussion about a paper that they have not read. Coming to Code Club
+without installing a dependency is like going to Journal Club and asking
+a simple question about the paper. It shows a lack of respect and
+preparation. Following Rule 4, review the code in advance, download the
+datasets, install the necessary packages, and perhaps read up on the
+topic. If your Code Club is reading a paper or chapters in a data
+science book be sure to read them before the session (e.g.~{[}5--7{]}).
+Consider how you could use the material for your own work.
 
-  \begin{enumerate}
-  \def\labelenumii{\alph{enumii}.}
-  \tightlist
-  \item
-    Be humble
-  \item
-    Be willing to incorporate feedback
-  \item
-    Ask clarifying questions
-  \item
-    Remember that this isn't a competition
-  \end{enumerate}
-\item
-  Reciprocate respect
+\hypertarget{rule-6-dont-give-your-participants-busy-work}{%
+\subsection{Rule 6: Don't give your participants busy
+work}\label{rule-6-dont-give-your-participants-busy-work}}
 
-  \begin{enumerate}
-  \def\labelenumii{\alph{enumii}.}
-  \tightlist
-  \item
-    Facilitation
+Participants want to learn something because it will be useful to them
+or because they think it will help the presenter to improve their code.
+Presenters should do their best to satisfy those desires. It would not
+make much sense to hold a Code Club on downloading stock market data if
+the group was not interested in those data. At the same time,
+participants should not be asked to help the presenter improve their
+code if the presenter has no intention of incorporating the suggestions.
+It would be useful to keep a list of packages or tasks that members of
+the group want to learn about. That way if a presenter is having a hard
+time coming up with ideas, they can take something from the list and
+present it.
 
-    \begin{enumerate}
-    \def\labelenumiii{\roman{enumiii}.}
-    \tightlist
-    \item
-      Lead the discussion
-    \item
-      Have an action plan: Focus the conversation
-    \item
-      Balance expertise in groups
-    \item
-      Attempt to purposefully engage early learners
-    \item
-      Avoid letting (senior) individuals dominate
-    \end{enumerate}
-  \item
-    Attitude:
+\hypertarget{rule-7-include-all-levels-of-participants}{%
+\subsection{Rule 7: Include all levels of
+participants}\label{rule-7-include-all-levels-of-participants}}
 
-    \begin{enumerate}
-    \def\labelenumiii{\roman{enumiii}.}
-    \tightlist
-    \item
-      Both:
+A significant challenge to presenting at Code Club is to pick topics and
+activities that appeal to a critical mass of the participants. For
+example, if there has recently been a large turn over in group
+membership, it can be useful to go back to the basics and cover commonly
+used tools. We have identified several strategies to overcome the
+challenges of having varying skill levels. As an alternative to letting
+participants form their own pairs, the presenter can be purposeful in
+how they form pairs. Depending on the goals of the session, forming
+pairs between participants that have similar or different skills can be
+effective. If pairs are formed where partners have similar skill levels,
+then the presenter needs to design activities that are appropriate to
+different levels of participants. We have found that a good activity for
+beginners is to have them comment code, which forces them to dissect a
+code chunk and understand what each line does. An advantage of forming
+pairs with disparate skill levels is that all of the pairs can be given
+the same task leading to a diversity of methods to achieve the same
+result. This approach to pairing also helps graft new members into the
+group that have emerging programming skills. Regardless of how partners
+are selected, asking the pairs to identify a navigator and driver is
+helpful {[}8{]}. The driver types at the computer while the navigator
+tells them what to type. Midway through the activity the presenter can
+tell the partners to switch roles. Intentionally forming pairs can also
+be a mechanism to engineer interactions by breaking up cliques, avoiding
+potentially toxic combinations of members, or pairing reliable role
+models with new members of the group.
 
-      \begin{enumerate}
-      \def\labelenumiv{\arabic{enumiv}.}
-      \tightlist
-      \item
-        Be humble
-      \item
-        Be willing to incorporate feedback
-      \item
-        Ask clarifying questions and note why you need the clarification
-      \end{enumerate}
-    \item
-      Participants:
+\hypertarget{rule-8-participate}{%
+\subsection{Rule 8: Participate}\label{rule-8-participate}}
 
-      \begin{enumerate}
-      \def\labelenumiv{\arabic{enumiv}.}
-      \tightlist
-      \item
-        Remember that this isn't a competition
-      \item
-        Focus on the speaker's goals
-      \item
-        Avoid distractions and engage with the presenter
-      \item
-        Allow others to contribute
-      \item
-        Give feedback \& coding suggestions at a novice explanation
-        level
-      \item
-        Give compliments
-      \end{enumerate}
-    \end{enumerate}
-  \end{enumerate}
-\end{enumerate}
+Having an open laptop on the table and permission to use it can feel
+like an invitation to get distracted by other work, emails, and browsing
+the internet. Fight that urge and focus on the presenter's goals. Allow
+your partner to contribute. Speak up for yourself and force your partner
+to let you contribute. Even if the material seems too advanced for you,
+programming languages like R and Python are generally expressive and
+should allow you to engage with the logic, even if the syntax is too
+hard. Frankly, understanding the logic of when to use one modeling
+approach over another is more important than how to use the modeling
+approach. If you understand the ``why'', the ``how'' will generally
+quickly follow. If you are a more experienced participant, give your
+feedback and coding suggestions at a level that seeks to inform
+everyone. Regardless of your skill level, your partner and the presenter
+are putting themselves at risk by revealing what they do or do not know.
+Encourage them and show your gratitude for helping you learn something
+new.
 
-\hypertarget{rule-2-set-specific-goals-presenter}{%
-\subsection{Rule 2: Set specific goals
-(Presenter)}\label{rule-2-set-specific-goals-presenter}}
+\hypertarget{rule-9-let-the-material-change-you}{%
+\subsection{Rule 9: Let the material change
+you}\label{rule-9-let-the-material-change-you}}
 
-Preparation 5. Lead the discussion a. Have an action plan c.~Focus the
-conversation 1. Set specific goals for each code review - Try to be as
-explicit and clear as possible for what you are asking the participants
-to do a. I have tried to be (what felt to me) painfully clear but I
-still was falling short making it clear to some people. b. People will
-always tend to notice or ask about things that are not the focus of the
-exercise, being specific about the goal may help. Also setting up a
-simplified scenario can be helpful (although requires much more prep
-time)
+Part of the humility required to participate in a Code Club is both
+acknowledging your incomplete training and that learning something new
+is possible. Assume that the presenter has followed the other rules and
+that they think you should be using what they presented. After the Code
+Club, try to incorporate that material into your new code or by
+refactoring your old code. By practicing the material in a different
+context you will learn the material better. You may identify concepts
+that you are unsure of, which gives you another opportunity to ask the
+presenter for help.
 
-\hypertarget{rule-3-keep-it-simple-presenter}{%
-\subsection{Rule 3: Keep it simple
-(Presenter)}\label{rule-3-keep-it-simple-presenter}}
+\hypertarget{rule-10-experiment}{%
+\subsection{Rule 10: Experiment!}\label{rule-10-experiment}}
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{1}
-\tightlist
-\item
-  Practice exercise
-
-  \begin{enumerate}
-  \def\labelenumii{\alph{enumii}.}
-  \tightlist
-  \item
-    When in doubt try it out! Run it by someone prior to lab meeting to
-    make sure your request/goal is understandable
-  \item
-    Running on different systems will inevitably run into issues. Send
-    out basic script ahead of time to try the dependencies/functions of
-    focus
-  \end{enumerate}
-\item
-  Create a \emph{minimal working example} - Working directly with your
-  own raw code can be challenging
-
-  \begin{enumerate}
-  \def\labelenumii{\alph{enumii}.}
-  \tightlist
-  \item
-    It is likely filled with code specific to you/your project
-  \item
-    It can take a lot of time to explain details and working of code
-  \item
-    Code can have multiple steps that are needed to run to get to script
-    focus
-  \item
-    Using a whole script opens up the risk of getting off topic
-  \end{enumerate}
-\item
-  Keep It Simple Silly (KISS)
-
-  \begin{enumerate}
-  \def\labelenumii{\alph{enumii}.}
-  \tightlist
-  \item
-    There is never enough time
-
-    \begin{enumerate}
-    \def\labelenumiii{\roman{enumiii}.}
-    \tightlist
-    \item
-      I typically have multiple activities planned, but rarely get
-      through them all
-    \item
-      Modularize the activity so that each can be its own and if
-      accomplished you are able to move onto the next
-    \end{enumerate}
-  \item
-    One activity per code review is typically sufficient
-
-    \begin{enumerate}
-    \def\labelenumiii{\roman{enumiii}.}
-    \tightlist
-    \item
-      This will take up most of the time
-
-      \begin{enumerate}
-      \def\labelenumiv{\arabic{enumiv}.}
-      \tightlist
-      \item
-        Intro - 10 min
-      \item
-        Think/pair - 30 min
-      \item
-        Share in group - 10 min
-      \item
-        Debrief - 5 min
-      \end{enumerate}
-    \item
-      I have not been successful in trying to do more than that in
-      different ways because it is going to be taking time away from one
-      of the other steps
-    \end{enumerate}
-  \end{enumerate}
-\end{enumerate}
-
-\hypertarget{rule-4-give-group-time-to-prepare-presenter}{%
-\subsection{Rule 4: Give group time to prepare
-(Presenter)}\label{rule-4-give-group-time-to-prepare-presenter}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\tightlist
-\item
-  Communicate with the group in advance so they can prepare
-
-  \begin{enumerate}
-  \def\labelenumii{\alph{enumii}.}
-  \tightlist
-  \item
-    Provide code chunk and necessary data
-  \item
-    Provide instructions for appropriate dependencies
-  \item
-    Provide use case (where available/appropriate)
-  \item
-    Specify goals \& expectations
-  \end{enumerate}
-\end{enumerate}
-
-\hypertarget{rule-5-prepare-in-advance-audience}{%
-\subsection{Rule 5: Prepare in advance
-(Audience)}\label{rule-5-prepare-in-advance-audience}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Prepare
-
-  \begin{enumerate}
-  \def\labelenumii{\alph{enumii}.}
-  \tightlist
-  \item
-    Review the code in advance
-  \item
-    Download datasets, install packages
-  \item
-    Google what you don't understand
-  \end{enumerate}
-\end{enumerate}
-
-Read chapter, paper, etc.
-
-\hypertarget{rule-6-dont-give-busy-work-presenter}{%
-\subsection{Rule 6: Don't give busy work
-(Presenter)}\label{rule-6-dont-give-busy-work-presenter}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{1}
-\tightlist
-\item
-  For Help! format, select your code
-
-  \begin{enumerate}
-  \def\labelenumii{\alph{enumii}.}
-  \tightlist
-  \item
-    That you are willing to receive and incorporate feedback on
-  \item
-    That is digestible in an hour (\textless100 lines)
-  \item
-    Use ``vanilla'' R to check for dependencies on a ``fresh''
-    environment
-  \item
-    Prepare a small example dataset
-  \end{enumerate}
-\end{enumerate}
-
-\hypertarget{rule-7-include-all-levels-of-learner-presenter}{%
-\subsection{Rule 7: Include all levels of learner
-(Presenter)}\label{rule-7-include-all-levels-of-learner-presenter}}
-
-\begin{itemize}
-\tightlist
-\item
-  Ask questions that are appropriate for different levels
-\item
-  Be purposeful about how you pair people off
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Use think/pair/share
-  \item
-    Senior/Beginner
-  \item
-    Senior/senior and Beginner/beginner
-  \item
-    Self-selected
-  \item
-    Navigator/driver with a switch midway through
-  \end{itemize}
-\end{itemize}
-
-\hypertarget{rule-8-engage-with-the-material-audience}{%
-\subsection{Rule 8: Engage with the material
-(Audience)}\label{rule-8-engage-with-the-material-audience}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\tightlist
-\item
-  Participate
-
-  \begin{enumerate}
-  \def\labelenumii{\alph{enumii}.}
-  \tightlist
-  \item
-    Focus on the speaker's goals
-  \item
-    Avoid distractions and engage with the presenter
-  \item
-    Allow others to contribute
-  \item
-    Ask clarifying questions and note why you need the clarification
-  \item
-    Give feedback \& coding suggestions at a novice explanation level
-  \item
-    Give compliments
-  \end{enumerate}
-\end{enumerate}
-
-\hypertarget{rule-9-let-the-material-change-you-audience}{%
-\subsection{Rule 9: Let the material change you
-(Audience)}\label{rule-9-let-the-material-change-you-audience}}
-
-\begin{itemize}
-\tightlist
-\item
-  Reflection
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Use what you've learned to refactor your own code / improve new code
-  \end{itemize}
-\end{itemize}
-
-\hypertarget{rule-10-speak-up-group}{%
-\subsection{Rule 10: Speak up (Group)}\label{rule-10-speak-up-group}}
-
-\begin{itemize}
-\tightlist
-\item
-  Speak up!
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Topics - check interest level of potential topics - check group is
-    at sufficient skill level needed - need for specific skills/methods
-    in group - May need to circle back to ``beginner'' stuff with
-    natural cycle of research group development
-
-    \begin{itemize}
-    \tightlist
-    \item
-      What works and doesn't
-    \end{itemize}
-  \end{itemize}
-\end{itemize}
+The selection of content and structure for each Code Club is determined
+by the presenter making the process democratic and distributed. If
+someone thinks something is worth learning and wants to teach it, they
+have that power. If they want to experiment with a different format,
+they are free to try it out. Members of the Code Club need to feel like
+they have the power to shape the direction of the group. If members are
+following these Rules, they will naturally reflect on the skills and
+interests of the other members in the group. For example, there is
+always turn over in a research group and thus it is important to make
+sure that there are opportunities to go back to things people consider
+``simple'' to teach the content to new people or provide a refresher to
+people that perhaps have been at the lab bench more than at the computer
+recently. The group should also feel free to experiment with the format.
+If we try a different format, then it is useful to give everyone an
+opportunity to debrief on what they did or did not like with the new
+approach.
 
@@
 
+The most important rules are the first and last. Members of the Code
+Club need to feel comfortable with other members of the group and
+sufficiently empowered to try something new or to ask for help. Aside
+from growing in our programming skills, we have noticed two other
+benefits that point to the efforts of creating a positive culture.
+First, we have intentionally interviewed postdoc candidates on the days
+we have held Code Club. Although it is important to make it clear that
+they are not being assessed on their coding skills, it is useful to see
+how the candidate interacts with other members of the research group in
+a relatively low stakes environment. At the same time, the candidate can
+learn about the culture of the research group. Second, members of other
+research groups where they are isolated in their desire to grow in these
+skills have integrated themselves into our Code Club. This speaks to the
+broader need for Code Club and the likelihood that it would work when
+expanded to a larger group of individuals that are not necessarily in
+the same research group. Ultimately, our Code Club has helped
+researchers feel more comfortable seeking out their colleagues outside
+of Code Club sessions for help in improving their code.
+
 \hypertarget{acknowledgements}{%
 \subsection{Acknowledgements}\label{acknowledgements}}
+
+All co-authors have been participants in the Code Clubs run as part of
+PDS's lab meetings. They have had a critical role in shaping the
+evolution of the sessions. The order of co-authors was determined
+by\ldots{}
 
@@
 \hypertarget{refs}{}
+\leavevmode\hypertarget{ref-Lonsdale2016}{}%
+1. Lonsdale A, Penington JS, Rice T, Walker M, Dashnow H. Ten simple
+rules for a bioinformatics journal club. Lewitter F, editor. PLOS
+Computational Biology. 2016;12: e1004526.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1004526}{10.1371/journal.pcbi.1004526}
 
-\url{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004526}
+\leavevmode\hypertarget{ref-Wilson2014}{}%
+2. Wilson G, Aruliah DA, Brown CT, Hong NPC, Davis M, Guy RT, et al.
+Best practices for scientific computing. Eisen JA, editor. PLoS Biology.
+2014;12: e1001745.
+doi:\href{https://doi.org/10.1371/journal.pbio.1001745}{10.1371/journal.pbio.1001745}
+
+\leavevmode\hypertarget{ref-Kothiyal2014}{}%
+3. Kothiyal A, Murthy S, Iyer S. Think-pair-share in a large CS1 class.
+Proceedings of the 2014 conference on innovation \& technology in
+computer science education - ITiCSE 14. ACM Press; 2014.
+doi:\href{https://doi.org/10.1145/2591708.2591739}{10.1145/2591708.2591739}
+
+\leavevmode\hypertarget{ref-StackOverflow_MREE}{}%
+4. Stack Overflow. How to create a minimal, reproducible example. 2020.
+Available:
+\url{https://stackoverflow.com/help/minimal-reproducible-example}
+
+\leavevmode\hypertarget{ref-Wickham_2016}{}%
+5. Wickham H, Grolemund G. O'Reilly Media; 2016. Available:
+\url{https://r4ds.had.co.nz}
+
+\leavevmode\hypertarget{ref-Schloss_2018}{}%
+6. Schloss PD. The riffomonas reproducible research tutorial series.
+Journal of Open Source Education. 2018;1: 13.
+doi:\href{https://doi.org/10.21105/jose.00013}{10.21105/jose.00013}
+
+\leavevmode\hypertarget{ref-Noble2009}{}%
+7. Noble WS. A quick guide to organizing computational biology projects.
+Lewitter F, editor. PLoS Computational Biology. 2009;5: e1000424.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1000424}{10.1371/journal.pcbi.1000424}
+
+\leavevmode\hypertarget{ref-Bockeler_2020}{}%
+8. Böckeler B, Siessegger N. On pair programming. 2020. Available:
+\url{https://martinfowler.com/articles/on-pair-programming.html}
 
@@
 
-\textbf{Figure 1. XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX
-XXXXX XXX.} XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX
-XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX XXXXX XXX
-XXXXX XXX XXXXX XXX XXXXX XXX.
+\textbf{Table 1. Examples of successful Code Club topics.}
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule
+\begin{minipage}[b]{0.25\columnwidth}\raggedright
+\textbf{Title}\strut
+\end{minipage} & \begin{minipage}[b]{0.69\columnwidth}\raggedright
+\textbf{Description}\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
+base vs.~ggplot\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+Given input data and a figure, recreate the figure using R's base
+graphics or ggplot syntax\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
+Snakemake\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+Given a bash script that contains an analysis pipeline, convert it to a
+series of recipes (can also be done with GNU Make)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
+DRYing code\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+Given script with repeated code, create functions to remove
+repetition\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
+mothur and Vegan\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+Given a pairwise community dissimilarity matrix, compare communities
+using the \texttt{adonis} function in the Vegan R package\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
+tidy data\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+Given a wide-formatted data table, convert it to a long, tidy-formatted
+data table using tools from R's tidyverse\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
+GitFlow\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+Participants file and claim an issue to add their name to a README file
+in a GitHub-hosted repository and file a pull request to complete the
+issue\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
+Integrating R and google docs\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+Scrape a google docs workbook and clean the data to identify previous
+Code Club presenters\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
 

### 2020-03-03 submission/manuscript.Rmd
Emphasize 2 different abilities (generating, analzying)

@@
 High throughput data comes to us from DNA and RNA sequencing, flow cytometry, metabolomics, molecular screens, and more.
-Although some accept the approach of compartmentalizing the ability to generate data and analyze data, we have found scientists feel empowered when they have the ability to both ask and answer their own biological questions.
+Although some accept the approach of compartmentalizing the ability to generate data and the ability analyze data, we have found scientists feel empowered when they can both ask and answer their own biological questions.
 Yet, the standard undergraduate and graduate training in the biological sciences is insufficient to develop the data analysis skills needed to analyze one's data.

### 2020-03-03 submission/manuscript.tex
Emphasize 2 different abilities (generating, analzying)

@@
 \makesavenoteenv{longtable}
-\usepackage{graphicx,grffile}
+\usepackage{graphicx}
 \makeatletter
@@
 \usepackage{pdfpages}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
 
@@
 more. Although some accept the approach of compartmentalizing the
-ability to generate data and analyze data, we have found scientists feel
-empowered when they have the ability to both ask and answer their own
+ability to generate data and the ability analyze data, we have found
+scientists feel empowered when they can both ask and answer their own
 biological questions. Yet, the standard undergraduate and graduate
@@
 \hypertarget{refs}{}
+\begin{cslreferences}
 \leavevmode\hypertarget{ref-Lonsdale2016}{}%
@@
 \url{https://martinfowler.com/articles/on-pair-programming.html}
+\end{cslreferences}
 

### 2020-03-04 submission/manuscript.Rmd
add a line about ML pipeline exercise

--- 2020-03-02 submission/manuscript.Rmd
@@
 Our group and others hold a regular Journal Club to stay current on the literature and model to junior scientists how to engage that literature [@Lonsdale2016].
-Over the past 4 years we have experimented with using a Code Club to introduce new software packages, solve specific problems, and evaluate each other's code.
+Over the past 4 years we have experimented with using a Code Club to introduce new software packages, solve specific problems, and evaluate each other's code in terms of reproducibility.
 
@@
 We continue to experiment with approaches to running Code Club.
+For example, recently we evaluated a labmate's bioinformatics pipeline to make it more user-friendly so that everyone in the lab could use it for their own projects.
 Although any individual Code Club session may be be more experimental than others, on the whole they have been a critical tool to providing the much needed training to use reproducible practices for analyzing data.

### 2020-03-04 submission/manuscript.Rmd
Use the term 'Snakemake workflow' instead of 'recipe'

--- 2020-03-03 submission/manuscript.Rmd
@@
 | base vs. ggplot   | Given input data and a figure, recreate the figure using R's base graphics or ggplot syntax   |
-| Snakemake         | Given a bash script that contains an analysis pipeline, convert it to a series of recipes (can also be done with GNU Make) |
+| Snakemake         | Given a bash script that contains an analysis pipeline, convert it to a Snakemake workflow (can also be done with GNU Make) |
 | DRYing code       | Given script with repeated code, create functions to remove repetition   |

### 2020-03-04 submission/manuscript.tex
Use the term 'Snakemake workflow' instead of 'recipe'

@@
 Given a bash script that contains an analysis pipeline, convert it to a
-series of recipes (can also be done with GNU Make)\strut
+Snakemake workflow (can also be done with GNU Make)\strut
 \end{minipage}\tabularnewline

### 2020-03-04 submission/manuscript.Rmd
Grammar edits

@@
 It is critical that the presenter and participants respect each other and that there be an individual (e.g.
-the lab director) that enforces a code of conduct.
+the lab director) who enforces a code of conduct.
 Everyone in the session needs to have the humility to acknowledge that they still have more to learn about any topic.
-We have found that many problems are avoided when the presenter takes charge of the session by having a clear lesson plan, thoughtfully creates groups, and gives encouragement.
+We have found that many problems are avoided when the presenter takes charge of the session by having a clear lesson plan, thoughtfully creating groups, and giving encouragement.
 Similarly, participants foster a positive environment by remembering that the task is not a competition, focusing on the presenter's goals, allowing their partner to contribute, asking clarifying questions when appropriate, and avoiding distractions (e.g.
@@
 Be specific about the goal and bring them back to the task if they are distracted.
-Where possible, create a simplified scenario (i.e.
-minimal, reproducible example [@StackOverflow_MREE]) can be helpful that removes distractions.
+Where possible, creating a simplified scenario (i.e.
+minimal, reproducible example [@StackOverflow_MREE]) can be helpful to remove distractions.
 The presenter should make sure that the simplified example works and behaves the way they expect before coming to the Code Club.
@@
 
-Our Code Club needs to fit within an hour slot.
+Our Code Club needs to fit within an hour time slot.
 The presenter should anticipate enough time for an introduction and brief instruction, time for participants to engage the material, and time for everyone to report back.

### 2020-03-04 submission/manuscript.tex
Grammar edits

@@
 It is critical that the presenter and participants respect each other
-and that there be an individual (e.g.~the lab director) that enforces a
+and that there be an individual (e.g.~the lab director) who enforces a
 code of conduct. Everyone in the session needs to have the humility to
@@
 found that many problems are avoided when the presenter takes charge of
-the session by having a clear lesson plan, thoughtfully creates groups,
-and gives encouragement. Similarly, participants foster a positive
+the session by having a clear lesson plan, thoughtfully creating groups,
+and giving encouragement. Similarly, participants foster a positive
 environment by remembering that the task is not a competition, focusing
@@
 exercise. Be specific about the goal and bring them back to the task if
-they are distracted. Where possible, create a simplified scenario
-(i.e.~minimal, reproducible example {[}4{]}) can be helpful that removes
+they are distracted. Where possible, creating a simplified scenario
+(i.e.~minimal, reproducible example {[}4{]}) can be helpful to remove
 distractions. The presenter should make sure that the simplified example
@@
 
-Our Code Club needs to fit within an hour slot. The presenter should
-anticipate enough time for an introduction and brief instruction, time
-for participants to engage the material, and time for everyone to report
-back. A typical schedule for Code Review is 10 minutes of introduction
-and instruction, 30 minutes of paired programming, 5 minutes to get
-groups to wrap up, and 10 minutes to report back to the group. Remember
-that a task that you can accomplish in 10 minutes may take a more junior
-member of your group 30 minutes. The goal of the Code Club should be to
-help that junior member to learn the topic and to reinforce the topic
-for the more senior person and help them see the concept in a new
-context. One possibility would be to create ``stretch goals'' for those
-that complete the task more quickly than everyone else. We once had a
-presenter try to teach the group some basic Julia syntax. By the time
-people installed the interpreter, the time was up. Alternatively, a
-session where participants were asked to make code DRY gave them
-multiple stopping points where they could feel successful
+Our Code Club needs to fit within an hour time slot. The presenter
+should anticipate enough time for an introduction and brief instruction,
+time for participants to engage the material, and time for everyone to
+report back. A typical schedule for Code Review is 10 minutes of
+introduction and instruction, 30 minutes of paired programming, 5
+minutes to get groups to wrap up, and 10 minutes to report back to the
+group. Remember that a task that you can accomplish in 10 minutes may
+take a more junior member of your group 30 minutes. The goal of the Code
+Club should be to help that junior member to learn the topic and to
+reinforce the topic for the more senior person and help them see the
+concept in a new context. One possibility would be to create ``stretch
+goals'' for those that complete the task more quickly than everyone
+else. We once had a presenter try to teach the group some basic Julia
+syntax. By the time people installed the interpreter, the time was up.
+Alternatively, a session where participants were asked to make code DRY
+gave them multiple stopping points where they could feel successful
 (e.g.~commenting code, creating function, implementing function,

### 2020-03-06 submission/manuscript.Rmd
Minor edits

@@
 
-In our early Code Clubs we noticed that if the presenter did not clearly state their goals for the session, it would often led to frustration for the presenter and participants.
-If the presenter shared their own code, did they want us to focus on their coding style or do they want help with incorporating a new package into their workflow?
+In our early Code Clubs we noticed that if the presenter did not clearly state their goals for the session, it would often lead to frustration for the presenter and participants.
+If the presenter shared their own code, did they want us to focus on their coding style or did they want help with incorporating a new package into their workflow?
 Participants will always notice or ask about things that are not the focus of the exercise.
-Be specific about the goal and bring them back to the task if they are distracted.
+Be specific about the goals and bring them back to the task if they are distracted.
 Where possible, creating a simplified scenario (i.e.
@@
 Alternatively, a session where participants were asked to make code DRY gave them multiple stopping points where they could feel successful (e.g.
-commenting code, creating function, implementing function, refactoring function).
+documenting code, writing a function signature, implementing a function, refactoring code).
 
@@
 Be sure to tell the participants what you will be covering.
-Provide the them with instructions on how to install dependencies, download data, and get the initial code.
+Provide them with instructions on how to install dependencies, download data, and get the initial code.
 Make sure that the materials work together as intended.
@@
 Following Rule 4, review the code in advance, download the datasets, install the necessary packages, and perhaps read up on the topic.
-If your Code Club is reading a paper or chapters in a data science book be sure to read them before the session (e.g.
+If your Code Club is reading a paper or chapters in a data science book, be sure to read them before the session (e.g.
 [@Wickham_2016; @Schloss_2018; @Noble2009]).
@@
 An advantage of forming pairs with disparate skill levels is that all of the pairs can be given the same task leading to a diversity of methods to achieve the same result.
-This approach to pairing also helps graft new members into the group that have emerging programming skills.
+This approach to pairing also helps graft new members that have emerging programming skills into the group.
 Regardless of how partners are selected, asking the pairs to identify a navigator and driver is helpful [@Bockeler_2020].
 The driver types at the computer while the navigator tells them what to type.
-Midway through the activity the presenter can tell the partners to switch roles.
+Midway through the activity, the partners can switch roles.
 Intentionally forming pairs can also be a mechanism to engineer interactions by breaking up cliques, avoiding potentially toxic combinations of members, or pairing reliable role models with new members of the group.
@@
 Even if the material seems too advanced for you, programming languages like R and Python are generally expressive and should allow you to engage with the logic, even if the syntax is too hard.
-Frankly, understanding the logic of when to use one modeling approach over another is more important than how to use the modeling approach.
+Frankly, understanding the logic of _when_ to use one modeling approach over another is more important than knowing _how_ to use the modeling approach.
 If you understand the "why", the "how" will generally quickly follow.
-If you are a more experienced participant, give your feedback and coding suggestions at a level that seeks to inform everyone.
-Regardless of your skill level, your partner and the presenter are putting themselves at risk by revealing what they do or do not know.
+If you are a more experienced programmer, give your feedback and coding suggestions at a level that seeks to inform everyone.
+Regardless of your skill level, your partner and the presenter are putting themselves in a vulnerable position by revealing what they do or do not know.
 Encourage them and show your gratitude for helping you learn something new.
@@
 After the Code Club, try to incorporate that material into your new code or by refactoring your old code.
-By practicing the material in a different context you will learn the material better.
+By practicing the material in a different context, you will learn the material better.
 You may identify concepts that you are unsure of, which gives you another opportunity to ask the presenter for help.
@@
 
-The selection of content and structure for each Code Club is determined by the presenter making the process democratic and distributed.
+The selection of content and structure for each Code Club is determined by the presenter, making the process democratic and distributed.
 If someone thinks something is worth learning and wants to teach it, they have that power.

### 2020-03-06 submission/manuscript.tex
Minor edits

@@
 In our early Code Clubs we noticed that if the presenter did not clearly
-state their goals for the session, it would often led to frustration for
-the presenter and participants. If the presenter shared their own code,
-did they want us to focus on their coding style or do they want help
-with incorporating a new package into their workflow? Participants will
-always notice or ask about things that are not the focus of the
-exercise. Be specific about the goal and bring them back to the task if
+state their goals for the session, it would often lead to frustration
+for the presenter and participants. If the presenter shared their own
+code, did they want us to focus on their coding style or did they want
+help with incorporating a new package into their workflow? Participants
+will always notice or ask about things that are not the focus of the
+exercise. Be specific about the goals and bring them back to the task if
 they are distracted. Where possible, creating a simplified scenario
@@
 gave them multiple stopping points where they could feel successful
-(e.g.~commenting code, creating function, implementing function,
-refactoring function).
+(e.g.~documenting code, writing a function signature, implementing a
+function, refactoring code).
 
@@
 days to prepare for the Code Club. Be sure to tell the participants what
-you will be covering. Provide the them with instructions on how to
-install dependencies, download data, and get the initial code. Make sure
-that the materials work together as intended. Considering the compressed
+you will be covering. Provide them with instructions on how to install
+dependencies, download data, and get the initial code. Make sure that
+the materials work together as intended. Considering the compressed
 schedule, asking participants to download materials beforehand is
@@
 topic. If your Code Club is reading a paper or chapters in a data
-science book be sure to read them before the session (e.g.~{[}5--7{]}).
+science book, be sure to read them before the session (e.g.~{[}5--7{]}).
 Consider how you could use the material for your own work.
@@
 the same task leading to a diversity of methods to achieve the same
-result. This approach to pairing also helps graft new members into the
-group that have emerging programming skills. Regardless of how partners
+result. This approach to pairing also helps graft new members that have
+emerging programming skills into the group. Regardless of how partners
 are selected, asking the pairs to identify a navigator and driver is
 helpful {[}8{]}. The driver types at the computer while the navigator
-tells them what to type. Midway through the activity the presenter can
-tell the partners to switch roles. Intentionally forming pairs can also
-be a mechanism to engineer interactions by breaking up cliques, avoiding
-potentially toxic combinations of members, or pairing reliable role
-models with new members of the group.
+tells them what to type. Midway through the activity, the partners can
+switch roles. Intentionally forming pairs can also be a mechanism to
+engineer interactions by breaking up cliques, avoiding potentially toxic
+combinations of members, or pairing reliable role models with new
+members of the group.
 
@@
 should allow you to engage with the logic, even if the syntax is too
-hard. Frankly, understanding the logic of when to use one modeling
-approach over another is more important than how to use the modeling
-approach. If you understand the ``why'', the ``how'' will generally
-quickly follow. If you are a more experienced participant, give your
-feedback and coding suggestions at a level that seeks to inform
-everyone. Regardless of your skill level, your partner and the presenter
-are putting themselves at risk by revealing what they do or do not know.
-Encourage them and show your gratitude for helping you learn something
-new.
+hard. Frankly, understanding the logic of \emph{when} to use one
+modeling approach over another is more important than knowing \emph{how}
+to use the modeling approach. If you understand the ``why'', the ``how''
+will generally quickly follow. If you are a more experienced programmer,
+give your feedback and coding suggestions at a level that seeks to
+inform everyone. Regardless of your skill level, your partner and the
+presenter are putting themselves in a vulnerable position by revealing
+what they do or do not know. Encourage them and show your gratitude for
+helping you learn something new.
 
@@
 refactoring your old code. By practicing the material in a different
-context you will learn the material better. You may identify concepts
+context, you will learn the material better. You may identify concepts
 that you are unsure of, which gives you another opportunity to ask the
@@
 The selection of content and structure for each Code Club is determined
-by the presenter making the process democratic and distributed. If
+by the presenter, making the process democratic and distributed. If
 someone thinks something is worth learning and wants to teach it, they

### 2020-03-28 submission/manuscript.Rmd
revisions

--- 2020-03-04 submission/manuscript.Rmd
@@
 
-# Ten Simple Rules for running a Code Club for increasing computational skills among biologists
+# Ten Simple Rules to increase computational skills among biologists with Code Clubs
 
@@
 
-*Author names here, order TBD*
+Ada K. Hagan${^2}$, Nick Lesniak, *Author names here, order TBD*
 
@@
 $1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109
+
+$2$ Alliance SciComm & Consulting, Linden, MI 48418
 
@@
 High throughput data comes to us from DNA and RNA sequencing, flow cytometry, metabolomics, molecular screens, and more.
-Although some accept the approach of compartmentalizing the ability to generate data and the ability analyze data, we have found scientists feel empowered when they can both ask and answer their own biological questions.
-Yet, the standard undergraduate and graduate training in the biological sciences is insufficient to develop the data analysis skills needed to analyze one's data.
-In our experience performing microbiome research, it is more common to find exceptional bench scientists who are inexperienced at analyzing large datasets than to find the reverse.
-Of course this raises a challenge: How do we train bench scientists to analyze datasets to answer biological questions?
+Although some accept the approach of compartmentalizing data generation and data analysis, we have found scientists feel empowered when they can both ask and answer their own biological questions.
+Yet, the standard undergraduate and graduate training in the biological sciences is insufficient for developing these advanced data analysis skills.
+In our experience performing microbiome research, it is more common to find exceptional bench scientists who are inexperienced at analyzing large data sets than to find the reverse.
+Of course this raises a challenge: How do we train bench scientists that analyze data sets to answer biological questions?
 
 The ever-growing ability to generate data and constant feeling of helplessness in analyzing it is analogous to the struggles we also face with engaging the voluminous scientific literature.
-Within our research group we have leveraged this similarity to address the challenges of training bench scientists to engage in reproducible data analysis practices.
-Our group and others hold a regular Journal Club to stay current on the literature and model to junior scientists how to engage that literature [@Lonsdale2016].
-Over the past 4 years we have experimented with using a Code Club to introduce new software packages, solve specific problems, and evaluate each other's code.
+A common strategy to keeping up with the literature are journal clubs, which involve group discussion of a pre-selected paper.
+These range from informal discussions to PowerPoint presentations and course credit.
+In addition to staying current on the literature, journal clubs help strengthen skills in critical thinking, communication, and literature review [@Lonsdale2016].
+Our research group has leveraged the similarity between the overwhelming natures of scientific literature and data analysis to address the challenge of teaching reproducible data analysis practices to bench scientists.
+Over the past 4 years we have experimented with a Code Club model to improve data analysis skills in a community environment.
 
-Initially, the Code Club was created as a way to review code that was being used in different projects.
-We would project someone's script onto a screen and go line by line through the code with each person in the room stating the logic behind the code.
-This approach emphasized the importance of readability and it gave beginners the opportunity to see the real, "smelly" code of more experienced people in the group.
-Unfortunately, the format only allowed us to review a small fraction of a project's code base and it became difficult to integrate the programmer's logic across their full project.
-We quickly realized that although beginners could contribute to the discussion, the more senior group members dominated conversation.
-This frequently created an adversarial environment, which led to the presenter not offering to present again.
-From this experience we learned that it is critical for the presenter to clearly articulate their goals for the Code Club and that the participants needed to engage with the code in that context.
+Our Code Club sessions are an hour long and alternate with Journal Club as the second part of weekly two hour lab meetings.
+Presenters volunteer to lead a session with the expectation that they alternate leading Journal and Code Clubs.
+Initially, the Code Club was used to review code from trainee projects.
+Instead of a presentation, the presenter would project their code onto a screen and the participants would go line by line through the code, stating the logic behind each line.
+This approach emphasized the importance of code readability and gave beginners the opportunity to see the real-life, messy code of more experienced peers.
+Unfortunately, the format only allowed us to review a fraction of a project's code base, making it difficult to integrate the programmer's logic across their full project.
+A major issue with this model was that sometimes beginners couldn't contribute to improving the code, and even when they could, more senior group members would eventually dominate the discussion.
+This led to a lack of participation by beginners, who would mentally check out to focus on other activities, and frequently resulted in an adversarial environment between the presenter and senior group members.
+As a result, presenters rarely offered to present Code Club again.
 
-Our Code Club sessions are each an hour long and are held roughly every other week as part of a weekly two hour lab meeting.
-Presenters voluntarily sign up to lead a session.
-The current format of our Code Club generally takes one of two approaches.
-The first is a more humane version of a group code review.
-The presenter will clearly state the problem they would like to solve, break the participants into smaller groups, and then ask each group to solve the problem.
-For example, someone may have an R script where a chunk of code is repeated throughout the script.
-The challenge for the session would be to convert the chunk into a function and call it throughout the script to make it "DRY" (i.e.
+From these experiences, we began a collective conversation to improve our Code Club model and have evolved it into two approaches.
+The first is a more constructive version of a group code review.
+The presenter clearly states the problem they want to solve, breaks the participants into smaller groups, and then asks each group to solve the problem, or a portion of it.
+For example, someone may have an R script with a repeating chunk of code.
+The challenge for the session would be to convert the code chunk into a function to be called throughout the script to make it "DRY" (i.e.
 Don't Repeat Yourself)[@Wilson2014].
-The presenter leaves the session with several partial or working solutions to their problem and everyone relearns the importance of writing DRY code.
-The second is a tutorial approach.
-The presenter introduces a new package or way of doing things.
-They then give the group an activity that forces them to replace their current approach with a new approach.
-For example, one of the more memorable activities was giving participants raw data and a finished plot and asking two pairs of participants to generate the plot using base R and asking the other two pairs to generate the plot with ggplot.
-The stipulation was that base R users had to build the plots in ggplot and ggplot users had to build the plot in base R.
-Each Code Club ends with an opportunity to report back to the presenter and larger group.
-Our Code Clubs typically have 7 to 10 participants.
-Because of the "think-pair-share" approach that is baked into each session it could likely be scaled to larger groups [@Kothiyal2014].
+The presenter leaves the session with several partial or working solutions to their problem and the importance of writing DRY code is reinforced.
+The second approach is a tutorial.
+The presenter introduces a new package or technique and assigns an activity to practice the new approach.
+For example, a more memorable activity was giving participants raw data and a finished plot.
+Paired participants were tasked with generating the plot using either base R or ggplot, but base R users had to use ggplot and vice versa.
+In either approach, the Code Club ends with a report back to the presenter and larger group describing the approach each pair took.
+Our Code Clubs typically have 7 to 10 participants, but the inherent "think-pair-share" strategy should allow it to be scaled to larger groups [@Kothiyal2014].
 
-We continue to experiment with approaches to running Code Club.
-Although any individual Code Club session may be be more experimental than others, on the whole they have been a critical tool to providing the much needed training to use reproducible practices for analyzing data.
-Examples of successful code club topics are described in Table 1.
-These Ten Rules summarize what we have learned as presenters and participants.
+We continue to experiment with approaches to running Code Club, but have learned that it is critical for the presenter to clearly articulate their goals and facilitate participant engagement.
+Although any individual Code Club session may be more experimental than another, on the whole they are a critical tool to train bench scientists in reproducible data analysis practices.
+We have provided some examples of successful code club topics in Table 1 and summarized the results of our own experimentation as Code Club presenters and participants into Ten Rules.
+The first 3 rules apply to all participants at every Code Club, while Rules 4 through 8 are targeted to Code Club presenters and the last two for participants.
 
@@
 
-It is critical that the presenter and participants respect each other and that there be an individual (e.g.
-the lab director) who enforces a code of conduct.
-Everyone in the session needs to have the humility to acknowledge that they still have more to learn about any topic.
-We have found that many problems are avoided when the presenter takes charge of the session by having a clear lesson plan, thoughtfully creating groups, and giving encouragement.
+It is critical that the presenter and participants respect each other and that a designated individual (e.g.
+the lab director) enforces a code of conduct (see X for an example).
+Each member of the Code Club must have the humility to acknowledge that they have more to learn about any given topic.
+We have found that many problems are avoided when the presenter takes charge of the session with a clear lesson plan, thoughtful group creation, and constructive encouragement.
 Similarly, participants foster a positive environment by remembering that the task is not a competition, focusing on the presenter's goals, allowing their partner to contribute, asking clarifying questions when appropriate, and avoiding distractions (e.g.
 email, social media).
-Learning to program is challenging and too often it can lead to a toxic environment.
-All parties in a Code Club are responsible for preventing this toxicity.
+Learning to program is challenging and too often results in a toxic environment of nitpicking and nonconstructive criticism.
+All parties in a Code Club are responsible for preventing this toxicity by demonstrating respect for themselves and their colleagues.
 
-## Rule 2: Set specific goals
+## Rule 2: Let the material change you
 
-In our early Code Clubs we noticed that if the presenter did not clearly state their goals for the session, it would often led to frustration for the presenter and participants.
-If the presenter shared their own code, did they want us to focus on their coding style or do they want help with incorporating a new package into their workflow?
-Participants will always notice or ask about things that are not the focus of the exercise.
-Be specific about the goal and bring them back to the task if they are distracted.
-Where possible, creating a simplified scenario (i.e.
-minimal, reproducible example [@StackOverflow_MREE]) can be helpful to remove distractions.
-The presenter should make sure that the simplified example works and behaves the way they expect before coming to the Code Club.
-Beyond the content, participants should know what they can be expected to have completed by the end of the session.
+Part of the humility required to participate in a Code Club is acknowledging that your training is incomplete and that it is possible for you to learn something new.
+For participants, assume that the presenter has a plan and follow their presented approach.
+After the Code Club, try to incorporate that material into new code or refactor old code.
+By practicing the material in a different context you will learn the material better.
+For presenters, incorporate suggested changes into your code.
+Either party may identify concepts that they are unsure of, presenting opportunities for further conversation and learning.
 
-## Rule 3: Keep it simple
+## Rule 3: Experiment!
 
-Our Code Club needs to fit within an hour time slot.
-The presenter should anticipate enough time for an introduction and brief instruction, time for participants to engage the material, and time for everyone to report back.
-A typical schedule for Code Review is 10 minutes of introduction and instruction, 30 minutes of paired programming, 5 minutes to get groups to wrap up, and 10 minutes to report back to the group.
-Remember that a task that you can accomplish in 10 minutes may take a more junior member of your group 30 minutes.
-The goal of the Code Club should be to help that junior member to learn the topic and to reinforce the topic for the more senior person and help them see the concept in a new context.
-One possibility would be to create "stretch goals" for those that complete the task more quickly than everyone else.
-We once had a presenter try to teach the group some basic Julia syntax.
-By the time people installed the interpreter, the time was up.
-Alternatively, a session where participants were asked to make code DRY gave them multiple stopping points where they could feel successful (e.g.
-commenting code, creating function, implementing function, refactoring function).
-
-## Rule 4: Give participants time to prepare
-
-Similar to a Journal Club, the presenter should give participants a few days to prepare for the Code Club.
-Be sure to tell the participants what you will be covering.
-Provide the them with instructions on how to install dependencies, download data, and get the initial code.
-Make sure that the materials work together as intended.
-Considering the compressed schedule, asking participants to download materials beforehand is helpful.
-We have found that using separate GitHub repositories for each session can be a helpful way to get information to participants.
-Of course, using git and GitHub to engage in collaborative coding should probably be its own Code Club before using the tools more broadly with the group.
-An easier entry point is to have groups post their code to a Slack channel that we use to announce materials for our meetings.
-
-## Rule 5: Prepare in advance to maximize your participation
-
-It is the rare scientist that can fully participate in a Journal Club discussion about a paper that they have not read.
-Coming to Code Club without installing a dependency is like going to Journal Club and asking a simple question about the paper.
-It shows a lack of respect and preparation.
-Following Rule 4, review the code in advance, download the datasets, install the necessary packages, and perhaps read up on the topic.
-If your Code Club is reading a paper or chapters in a data science book be sure to read them before the session (e.g.
-[@Wickham_2016; @Schloss_2018; @Noble2009]).
-Consider how you could use the material for your own work.
-
-## Rule 6: Don't give your participants busy work
-
-Participants want to learn something because it will be useful to them or because they think it will help the presenter to improve their code.
-Presenters should do their best to satisfy those desires.
-It would not make much sense to hold a Code Club on downloading stock market data if the group was not interested in those data.
-At the same time, participants should not be asked to help the presenter improve their code if the presenter has no intention of incorporating the suggestions.
-It would be useful to keep a list of packages or tasks that members of the group want to learn about.
-That way if a presenter is having a hard time coming up with ideas, they can take something from the list and present it.
-
-## Rule 7: Include all levels of participants
-
-A significant challenge to presenting at Code Club is to pick topics and activities that appeal to a critical mass of the participants.
-For example, if there has recently been a large turn over in group membership, it can be useful to go back to the basics and cover commonly used tools.
-We have identified several strategies to overcome the challenges of having varying skill levels.
-As an alternative to letting participants form their own pairs, the presenter can be purposeful in how they form pairs.
-Depending on the goals of the session, forming pairs between participants that have similar or different skills can be effective.
-If pairs are formed where partners have similar skill levels, then the presenter needs to design activities that are appropriate to different levels of participants.
-We have found that a good activity for beginners is to have them comment code, which forces them to dissect a code chunk and understand what each line does.
-An advantage of forming pairs with disparate skill levels is that all of the pairs can be given the same task leading to a diversity of methods to achieve the same result.
-This approach to pairing also helps graft new members into the group that have emerging programming skills.
-Regardless of how partners are selected, asking the pairs to identify a navigator and driver is helpful [@Bockeler_2020].
-The driver types at the computer while the navigator tells them what to type.
-Midway through the activity the presenter can tell the partners to switch roles.
-Intentionally forming pairs can also be a mechanism to engineer interactions by breaking up cliques, avoiding potentially toxic combinations of members, or pairing reliable role models with new members of the group.
-
-## Rule 8: Participate
-
-Having an open laptop on the table and permission to use it can feel like an invitation to get distracted by other work, emails, and browsing the internet.
-Fight that urge and focus on the presenter's goals.
-Allow your partner to contribute.
-Speak up for yourself and force your partner to let you contribute.
-Even if the material seems too advanced for you, programming languages like R and Python are generally expressive and should allow you to engage with the logic, even if the syntax is too hard.
-Frankly, understanding the logic of when to use one modeling approach over another is more important than how to use the modeling approach.
-If you understand the "why", the "how" will generally quickly follow.
-If you are a more experienced participant, give your feedback and coding suggestions at a level that seeks to inform everyone.
-Regardless of your skill level, your partner and the presenter are putting themselves at risk by revealing what they do or do not know.
-Encourage them and show your gratitude for helping you learn something new.
-
-## Rule 9: Let the material change you
-
-Part of the humility required to participate in a Code Club is both acknowledging your incomplete training and that learning something new is possible.
-Assume that the presenter has followed the other rules and that they think you should be using what they presented.
-After the Code Club, try to incorporate that material into your new code or by refactoring your old code.
-By practicing the material in a different context you will learn the material better.
-You may identify concepts that you are unsure of, which gives you another opportunity to ask the presenter for help.
-
-## Rule 10: Experiment!
-
-The selection of content and structure for each Code Club is determined by the presenter making the process democratic and distributed.
-If someone thinks something is worth learning and wants to teach it, they have that power.
+The selection of content and structure for each Code Club is democratic and distributed.
+If someone thinks a technique is worth learning and wants to teach it, they have that power.
 If they want to experiment with a different format, they are free to try it out.
@@
 If members are following these Rules, they will naturally reflect on the skills and interests of the other members in the group.
-For example, there is always turn over in a research group and thus it is important to make sure that there are opportunities to go back to things people consider "simple" to teach the content to new people or provide a refresher to people that perhaps have been at the lab bench more than at the computer recently.
-The group should also feel free to experiment with the format.
-If we try a different format, then it is useful to give everyone an opportunity to debrief on what they did or did not like with the new approach.
+For example, there is always turnover in a research group, making it important to revisit basic concepts to teach it to new people and provide a refresher to others.
+The group should also feel free to experiment with the format and incorporate group feedback by ending with a debrief discussing the pros and cons of new formats.
+
+## Rule 4: Set specific goals
+
+In our early Code Clubs we noticed that if the presenter did not clearly state their goals for the session, it often led to frustration for both the presenter and participants.
+If the presenter shared their own code, did they want participants to focus on their coding style or did they want help incorporating a new package into their workflow?
+Participants will always notice or ask about code concepts that are not the focus of the exercise; a presenter with a specific goal can bring tangential conversations back to the planned task.
+Where possible, presenters should create a simplified scenario (i.e.
+minimal, reproducible example [@StackOverflow_MREE]), which can be helpful in focusing the participants.
+The presenter should verify ahead of time that the simplified example works and behaves the way they expect.
+Beyond the content, clear goals for participant activities will help both parties stay on task and avoid frustration.
+For more advanced learners, the presenter can create stretch goals or give an activity with multiple stopping points where participants would feel successful (e.g.
+commenting code, creating function, implementing function, refactoring function).
+Accomplishing specific goals is more likely to result in a positive outcome for presenters and participants.
+
+## Rule 5: Keep it simple
+
+Our Code Club needs to fit within an hour time slot.
+When considering their Code Club activity, the presenter should plan for an introduction and brief instruction, time for participants to engage the material, and time for everyone to report back within that hour.
+A typical schedule for Code Club is 10 minutes of introduction and instruction, 30 minutes of paired programming, 5 minutes to get groups to wrap up, and 10 minutes to report back to the group.
+Remember that learners may need up to three times as long to complete a task that is simple for the presenter, so Code Club is best kept simple.
+We once had a presenter try to teach basic Julia syntax, but time was up by the time participants installed the interpreter.
+Some tips to help keep it simple are to limit the presented code chunks to less than 50 or, conversely, consider the number of lines that might be required to accomplish a solution.
+
+## Rule 6: Give participants time to prepare
+
+Similar to a Journal Club, the presenter should give participants a few days (ideally a week) to prepare for the Code Club.
+Considering the compressed schedule described in Rule 5, asking participants to download materials beforehand is helpful to ensure a quick start.
+The presenter should provide the participants with instructions on how to install dependencies, download data, and get the initial code.
+This might also uncover weak points in the presenter's plan and enable them to ensure that the materials work as intended before the Code Club.
+We have found that using GitHub repositories for each Code Club can help make information, scripts, and data easily available to participants.
+Perhaps a first Code Club could be using git and GitHub to engage in collaborative coding.
+A lower barrier entry point is posting their code, data, and information in a lab meeting-dedicated Slack channel or sharing via email.
+Whatever method is used, the presenter should be sure to communicate the topic and necessary materials with the participants ahead of time.
+
+## Rule 7: Don't give participants busy work
+
+Participants want to learn topics that will either be useful to them or help their colleague (i.e., the presenter).
+Presenters should do their best to satisfy those motivations, whether it is through the relevancy of the concept or the data.
+It does not make sense to present a Code Club on downloading stock market data if it is not useful or interesting to the group.
+Similarly, participants should not be tasked with improving the presenter's code if the presenter has no intention of incorporating the suggestions.
+A list of packages or tasks that group members are interested in could help an uninspired presenter choose a topic that will make a rewarding Code Club.
+
+## Rule 8: Include all levels of participants
+
+As suggested by Rule 7, a significant challenge to presenting at Code Club is selecting topics and activities that appeal to a critical mass of the participants.
+This is particularly difficult if participants have a wide range of coding experience, such as follows a turnover in group membership.
+In these circumstances it can be useful to cover commonly used tools but that might result in disinterest of more experienced participants.
+We have identified several strategies to overcome the challenges presented by participants at varying skill levels.
+Instead of letting participants form their own pairs, the presenter can select pairs of participants with either similar or differing skill levels, depending on their goals.
+Partnerships between those with similar skill levels requires the presenter to design appropriate activites each skill level.
+We have found that commenting code is a good skill for beginners since it forces them to dissect and understand a code chunk line-by-line.
+An advantage of partnerships with disparate skill levels is assigning a single task, which provides the presenter with a diverse range of methods that achieve the same result.
+This approach to pairing also helps to graft new members into the group that have emerging programming skills.
+Regardless of how partners are selected, consider asking the pairs to identify a navigator and a driver [@Bockeler_2020].
+The driver types at the computer while the navigator tells them what to type, thus ensuring participation of both partners.
+Midway through the activity, the presenter can have the partners switch roles.
+Intentionally forming pairs can also engineer group interactions by breaking up cliques, avoiding potentially disruptive partnerships, or pairing reliable role models with new group members.
+
+## Rule 9: Prepare in advance to maximize participation
+
+It is not possible to fully participate in a Journal Club discussion about a paper that the participant has not read.
+In that context, coming to Code Club without having installed a dependency is similar to asking a simple question about the Journal Club paper.
+Both instances show a lack of respect and preparation.
+Just as a presenter must follow Rule 6 to provide materials ahead of time, participants must review the code in advance, download the data sets, install the necessary packages, and perhaps read up on the topic.
+If the Code Club is based on a paper or chapters in a data science book (e.g., [@Wickham_2016; @Schloss_2018; @Noble2009]) the participants should read them before the session and consider how they might incorporate the concepts into their own work.
+
+## Rule 10: Participate
+
+An essential ingredient of any Code Club is active participation from all parties.
+Having an open laptop on the table and permission to use it can feel like an invitation to get distracted by other work, emails, and browsing the internet; fight that urge and focus on the presenter's goals.
+Be respectful and allow your partner to contribute.
+Speak up for yourself and force your partner to let you contribute.
+If the material seems too advanced for you, it can be frustrating, and tempting to mentally check out.
+Fortuntately, programming languages like R and Python are generally expressive, which should allow you to engage with the logic, even if the syntax is too advanced.
+Frankly, understanding the logic of when to use one modeling approach over another is more important than how to use it.
+If you understand the "why", the "how" will quickly follow.
+More experienced participants should aim to communicate feedback and coding suggestions at a level that all participants can understand and engage in.
+Regardless of skill levels, your partner and the presenter put themselves at risk by revealing what they do or do not know.
+Encourage them, and show your gratitude for helping you learn something new, by fully participating in each Code Club.
 
@@
 The most important rules are the first and last.
-Members of the Code Club need to feel comfortable with other members of the group and sufficiently empowered to try something new or to ask for help.
-Aside from growing in our programming skills, we have noticed two other benefits that point to the efforts of creating a positive culture.
-First, we have intentionally interviewed postdoc candidates on the days we have held Code Club.
-Although it is important to make it clear that they are not being assessed on their coding skills, it is useful to see how the candidate interacts with other members of the research group in a relatively low stakes environment.
-At the same time, the candidate can learn about the culture of the research group.
-Second, members of other research groups where they are isolated in their desire to grow in these skills have integrated themselves into our Code Club.
-This speaks to the broader need for Code Club and the likelihood that it would work when expanded to a larger group of individuals that are not necessarily in the same research group.
-Ultimately, our Code Club has helped researchers feel more comfortable seeking out their colleagues outside of Code Club sessions for help in improving their code.
+Members of the Code Club need to feel comfortable with other group members and sufficiently empowered to try something new or ask for help.
+Aside from growing in our programming skills, we have noticed two other benefits that help create a positive work culture.
+First, we have intentionally interviewed postdoc candidates on the days we hold Code Club.
+We make it clear that they are not being assessed on their coding skills, but instead use it as an opportunity to see how the candidate interacts with other members of the research group in a relatively low stakes environment.
+At the same time, the candidate can learn about the culture of the research group through active participation.
+Second, members of other research groups that feel isolated in growing their skills have integrated themselves into our Code Club.
+This speaks to both the broader need for Code Club and the likelihood of success when expanded to include a larger group of individuals with broader research interests.
+Ultimately, Code Club has improved the overall data analysis skills, community, and research success of our lab by empowering researchers to seek help from their colleagues.
 
@@
 They have had a critical role in shaping the evolution of the sessions.
+AKH led a lab discussion of Code Clubs to identify the key concepts presented, which were supplemented by contributions from NL.
 The order of co-authors was determined by...
@@
 | base vs. ggplot   | Given input data and a figure, recreate the figure using R's base graphics or ggplot syntax   |
-| Snakemake         | Given a bash script that contains an analysis pipeline, convert it to a Snakemake workflow (can also be done with GNU Make) |
+| Snakemake         | Given a bash script that contains an analysis pipeline, convert it to a series of recipes (can also be done with GNU Make) |
 | DRYing code       | Given script with repeated code, create functions to remove repetition   |
@@
 | GitFlow           | Participants file and claim an issue to add their name to a README file in a GitHub-hosted repository and file a pull request to complete the issue |
-| Integrating R and google docs   | Scrape a google docs workbook and clean the data to identify previous Code Club presenters |
+| Integrating R and Google docs   | Scrape a Google docs workbook and clean the data to identify previous Code Club presenters |

### 2020-04-07 submission/manuscript.tex

--- 2020-03-04 submission/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  11pt,
-]{article}
+\documentclass[11pt,]{article}
 \usepackage{lmodern}
@@
 \usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+  \else
+    \usepackage{fontspec}
   \fi
-% Use upquote if available, for straight quotes in verbatim environments
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
+% use microtype if available
+\IfFileExists{microtype.sty}{%
 \usepackage[]{microtype}
@@
 }{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage[unicode=true]{hyperref}
 \hypersetup{
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
 \usepackage[margin=1.0in]{geometry}
 \usepackage{longtable,booktabs}
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
+% Fix footnotes in tables (requires footnote package)
+\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
+\usepackage{graphicx,grffile}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{0}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+% set default figure placement to htbp
 \makeatletter
@@
 \makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+
 \usepackage{helvet} % Helvetica font
@@
 \usepackage{pdfpages}
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
 
@@
 
-\hypertarget{ten-simple-rules-for-running-a-code-club-for-increasing-computational-skills-among-biologists}{%
-\section{Ten Simple Rules for running a Code Club for increasing
-computational skills among
-biologists}\label{ten-simple-rules-for-running-a-code-club-for-increasing-computational-skills-among-biologists}}
+\section{Ten Simple Rules to increase computational skills among
+biologists with Code
+Clubs}\label{ten-simple-rules-to-increase-computational-skills-among-biologists-with-code-clubs}
 
@@
 
-\emph{Author names here, order TBD}
+Ada K. Hagan\({^2}\), Nick Lesniak, \emph{Author names here, order TBD}
 
@@
 
+\(2\) Alliance SciComm \& Consulting, Linden, MI 48418
+
 \vspace{35mm}
 
-\hypertarget{ten-simple-rules}{%
-\subsubsection{Ten Simple Rules}\label{ten-simple-rules}}
+\subsubsection{Ten Simple Rules}\label{ten-simple-rules}
 
 \newpage
+
 \linenumbers
 
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
+\subsection{Introduction}\label{introduction}
 
@@
 and RNA sequencing, flow cytometry, metabolomics, molecular screens, and
-more. Although some accept the approach of compartmentalizing the
-ability to generate data and the ability analyze data, we have found
-scientists feel empowered when they can both ask and answer their own
-biological questions. Yet, the standard undergraduate and graduate
-training in the biological sciences is insufficient to develop the data
-analysis skills needed to analyze one's data. In our experience
-performing microbiome research, it is more common to find exceptional
-bench scientists who are inexperienced at analyzing large datasets than
-to find the reverse. Of course this raises a challenge: How do we train
-bench scientists to analyze datasets to answer biological questions?
+more. Although some accept the approach of compartmentalizing data
+generation and data analysis, we have found scientists feel empowered
+when they can both ask and answer their own biological questions. Yet,
+the standard undergraduate and graduate training in the biological
+sciences is insufficient for developing these advanced data analysis
+skills. In our experience performing microbiome research, it is more
+common to find exceptional bench scientists who are inexperienced at
+analyzing large data sets than to find the reverse. Of course this
+raises a challenge: How do we train bench scientists that analyze data
+sets to answer biological questions?
 
@@
 helplessness in analyzing it is analogous to the struggles we also face
-with engaging the voluminous scientific literature. Within our research
-group we have leveraged this similarity to address the challenges of
-training bench scientists to engage in reproducible data analysis
-practices. Our group and others hold a regular Journal Club to stay
-current on the literature and model to junior scientists how to engage
-that literature {[}1{]}. Over the past 4 years we have experimented with
-using a Code Club to introduce new software packages, solve specific
-problems, and evaluate each other's code.
+with engaging the voluminous scientific literature. A common strategy to
+keeping up with the literature are journal clubs, which involve group
+discussion of a pre-selected paper. These range from informal
+discussions to PowerPoint presentations and course credit. In addition
+to staying current on the literature, journal clubs help strengthen
+skills in critical thinking, communication, and literature review
+{[}1{]}. Our research group has leveraged the similarity between the
+overwhelming natures of scientific literature and data analysis to
+address the challenge of teaching reproducible data analysis practices
+to bench scientists. Over the past 4 years we have experimented with a
+Code Club model to improve data analysis skills in a community
+environment.
 
-Initially, the Code Club was created as a way to review code that was
-being used in different projects. We would project someone's script onto
-a screen and go line by line through the code with each person in the
-room stating the logic behind the code. This approach emphasized the
-importance of readability and it gave beginners the opportunity to see
-the real, ``smelly'' code of more experienced people in the group.
-Unfortunately, the format only allowed us to review a small fraction of
-a project's code base and it became difficult to integrate the
-programmer's logic across their full project. We quickly realized that
-although beginners could contribute to the discussion, the more senior
-group members dominated conversation. This frequently created an
-adversarial environment, which led to the presenter not offering to
-present again. From this experience we learned that it is critical for
-the presenter to clearly articulate their goals for the Code Club and
-that the participants needed to engage with the code in that context.
+Our Code Club sessions are an hour long and alternate with Journal Club
+as the second part of weekly two hour lab meetings. Presenters volunteer
+to lead a session with the expectation that they alternate leading
+Journal and Code Clubs. Initially, the Code Club was used to review code
+from trainee projects. Instead of a presentation, the presenter would
+project their code onto a screen and the participants would go line by
+line through the code, stating the logic behind each line. This approach
+emphasized the importance of code readability and gave beginners the
+opportunity to see the real-life, messy code of more experienced peers.
+Unfortunately, the format only allowed us to review a fraction of a
+project's code base, making it difficult to integrate the programmer's
+logic across their full project. A major issue with this model was that
+sometimes beginners couldn't contribute to improving the code, and even
+when they could, more senior group members would eventually dominate the
+discussion. This led to a lack of participation by beginners, who would
+mentally check out to focus on other activities, and frequently resulted
+in an adversarial environment between the presenter and senior group
+members. As a result, presenters rarely offered to present Code Club
+again.
 
-Our Code Club sessions are each an hour long and are held roughly every
-other week as part of a weekly two hour lab meeting. Presenters
-voluntarily sign up to lead a session. The current format of our Code
-Club generally takes one of two approaches. The first is a more humane
-version of a group code review. The presenter will clearly state the
-problem they would like to solve, break the participants into smaller
-groups, and then ask each group to solve the problem. For example,
-someone may have an R script where a chunk of code is repeated
-throughout the script. The challenge for the session would be to convert
-the chunk into a function and call it throughout the script to make it
-``DRY'' (i.e.~Don't Repeat Yourself){[}2{]}. The presenter leaves the
-session with several partial or working solutions to their problem and
-everyone relearns the importance of writing DRY code. The second is a
-tutorial approach. The presenter introduces a new package or way of
-doing things. They then give the group an activity that forces them to
-replace their current approach with a new approach. For example, one of
-the more memorable activities was giving participants raw data and a
-finished plot and asking two pairs of participants to generate the plot
-using base R and asking the other two pairs to generate the plot with
-ggplot. The stipulation was that base R users had to build the plots in
-ggplot and ggplot users had to build the plot in base R. Each Code Club
-ends with an opportunity to report back to the presenter and larger
-group. Our Code Clubs typically have 7 to 10 participants. Because of
-the ``think-pair-share'' approach that is baked into each session it
-could likely be scaled to larger groups {[}3{]}.
+From these experiences, we began a collective conversation to improve
+our Code Club model and have evolved it into two approaches. The first
+is a more constructive version of a group code review. The presenter
+clearly states the problem they want to solve, breaks the participants
+into smaller groups, and then asks each group to solve the problem, or a
+portion of it. For example, someone may have an R script with a
+repeating chunk of code. The challenge for the session would be to
+convert the code chunk into a function to be called throughout the
+script to make it ``DRY'' (i.e.~Don't Repeat Yourself){[}2{]}. The
+presenter leaves the session with several partial or working solutions
+to their problem and the importance of writing DRY code is reinforced.
+The second approach is a tutorial. The presenter introduces a new
+package or technique and assigns an activity to practice the new
+approach. For example, a more memorable activity was giving participants
+raw data and a finished plot. Paired participants were tasked with
+generating the plot using either base R or ggplot, but base R users had
+to use ggplot and vice versa. In either approach, the Code Club ends
+with a report back to the presenter and larger group describing the
+approach each pair took. Our Code Clubs typically have 7 to 10
+participants, but the inherent ``think-pair-share'' strategy should
+allow it to be scaled to larger groups {[}3{]}.
 
-We continue to experiment with approaches to running Code Club. Although
-any individual Code Club session may be be more experimental than
-others, on the whole they have been a critical tool to providing the
-much needed training to use reproducible practices for analyzing data.
-Examples of successful code club topics are described in Table 1. These
-Ten Rules summarize what we have learned as presenters and participants.
+We continue to experiment with approaches to running Code Club, but have
+learned that it is critical for the presenter to clearly articulate
+their goals and facilitate participant engagement. Although any
+individual Code Club session may be more experimental than another, on
+the whole they are a critical tool to train bench scientists in
+reproducible data analysis practices. We have provided some examples of
+successful code club topics in Table 1 and summarized the results of our
+own experimentation as Code Club presenters and participants into Ten
+Rules. The first 3 rules apply to all participants at every Code Club,
+while Rules 4 through 8 are targeted to Code Club presenters and the
+last two for participants.
 
-\hypertarget{rule-1-reciprocate-respect}{%
 \subsection{Rule 1: Reciprocate
-respect}\label{rule-1-reciprocate-respect}}
+respect}\label{rule-1-reciprocate-respect}
 
 It is critical that the presenter and participants respect each other
-and that there be an individual (e.g.~the lab director) who enforces a
-code of conduct. Everyone in the session needs to have the humility to
-acknowledge that they still have more to learn about any topic. We have
-found that many problems are avoided when the presenter takes charge of
-the session by having a clear lesson plan, thoughtfully creating groups,
-and giving encouragement. Similarly, participants foster a positive
-environment by remembering that the task is not a competition, focusing
-on the presenter's goals, allowing their partner to contribute, asking
-clarifying questions when appropriate, and avoiding distractions
-(e.g.~email, social media). Learning to program is challenging and too
-often it can lead to a toxic environment. All parties in a Code Club are
-responsible for preventing this toxicity.
+and that a designated individual (e.g.~the lab director) enforces a code
+of conduct (see X for an example). Each member of the Code Club must
+have the humility to acknowledge that they have more to learn about any
+given topic. We have found that many problems are avoided when the
+presenter takes charge of the session with a clear lesson plan,
+thoughtful group creation, and constructive encouragement. Similarly,
+participants foster a positive environment by remembering that the task
+is not a competition, focusing on the presenter's goals, allowing their
+partner to contribute, asking clarifying questions when appropriate, and
+avoiding distractions (e.g.~email, social media). Learning to program is
+challenging and too often results in a toxic environment of nitpicking
+and nonconstructive criticism. All parties in a Code Club are
+responsible for preventing this toxicity by demonstrating respect for
+themselves and their colleagues.
 
-\hypertarget{rule-2-set-specific-goals}{%
-\subsection{Rule 2: Set specific
-goals}\label{rule-2-set-specific-goals}}
+\subsection{Rule 2: Let the material change
+you}\label{rule-2-let-the-material-change-you}
+
+Part of the humility required to participate in a Code Club is
+acknowledging that your training is incomplete and that it is possible
+for you to learn something new. For participants, assume that the
+presenter has a plan and follow their presented approach. After the Code
+Club, try to incorporate that material into new code or refactor old
+code. By practicing the material in a different context you will learn
+the material better. For presenters, incorporate suggested changes into
+your code. Either party may identify concepts that they are unsure of,
+presenting opportunities for further conversation and learning.
+
+\subsection{Rule 3: Experiment!}\label{rule-3-experiment}
+
+The selection of content and structure for each Code Club is democratic
+and distributed. If someone thinks a technique is worth learning and
+wants to teach it, they have that power. If they want to experiment with
+a different format, they are free to try it out. Members of the Code
+Club need to feel like they have the power to shape the direction of the
+group. If members are following these Rules, they will naturally reflect
+on the skills and interests of the other members in the group. For
+example, there is always turnover in a research group, making it
+important to revisit basic concepts to teach it to new people and
+provide a refresher to others. The group should also feel free to
+experiment with the format and incorporate group feedback by ending with
+a debrief discussing the pros and cons of new formats.
+
+\subsection{Rule 4: Set specific goals}\label{rule-4-set-specific-goals}
 
 In our early Code Clubs we noticed that if the presenter did not clearly
-state their goals for the session, it would often led to frustration for
+state their goals for the session, it often led to frustration for both
 the presenter and participants. If the presenter shared their own code,
-did they want us to focus on their coding style or do they want help
-with incorporating a new package into their workflow? Participants will
-always notice or ask about things that are not the focus of the
-exercise. Be specific about the goal and bring them back to the task if
-they are distracted. Where possible, creating a simplified scenario
-(i.e.~minimal, reproducible example {[}4{]}) can be helpful to remove
-distractions. The presenter should make sure that the simplified example
-works and behaves the way they expect before coming to the Code Club.
-Beyond the content, participants should know what they can be expected
-to have completed by the end of the session.
+did they want participants to focus on their coding style or did they
+want help incorporating a new package into their workflow? Participants
+will always notice or ask about code concepts that are not the focus of
+the exercise; a presenter with a specific goal can bring tangential
+conversations back to the planned task. Where possible, presenters
+should create a simplified scenario (i.e.~minimal, reproducible example
+{[}4{]}), which can be helpful in focusing the participants. The
+presenter should verify ahead of time that the simplified example works
+and behaves the way they expect. Beyond the content, clear goals for
+participant activities will help both parties stay on task and avoid
+frustration. For more advanced learners, the presenter can create
+stretch goals or give an activity with multiple stopping points where
+participants would feel successful (e.g.~commenting code, creating
+function, implementing function, refactoring function). Accomplishing
+specific goals is more likely to result in a positive outcome for
+presenters and participants.
 
-\hypertarget{rule-3-keep-it-simple}{%
-\subsection{Rule 3: Keep it simple}\label{rule-3-keep-it-simple}}
+\subsection{Rule 5: Keep it simple}\label{rule-5-keep-it-simple}
 
-Our Code Club needs to fit within an hour time slot. The presenter
-should anticipate enough time for an introduction and brief instruction,
-time for participants to engage the material, and time for everyone to
-report back. A typical schedule for Code Review is 10 minutes of
-introduction and instruction, 30 minutes of paired programming, 5
-minutes to get groups to wrap up, and 10 minutes to report back to the
-group. Remember that a task that you can accomplish in 10 minutes may
-take a more junior member of your group 30 minutes. The goal of the Code
-Club should be to help that junior member to learn the topic and to
-reinforce the topic for the more senior person and help them see the
-concept in a new context. One possibility would be to create ``stretch
-goals'' for those that complete the task more quickly than everyone
-else. We once had a presenter try to teach the group some basic Julia
-syntax. By the time people installed the interpreter, the time was up.
-Alternatively, a session where participants were asked to make code DRY
-gave them multiple stopping points where they could feel successful
-(e.g.~commenting code, creating function, implementing function,
-refactoring function).
+Our Code Club needs to fit within an hour time slot. When considering
+their Code Club activity, the presenter should plan for an introduction
+and brief instruction, time for participants to engage the material, and
+time for everyone to report back within that hour. A typical schedule
+for Code Club is 10 minutes of introduction and instruction, 30 minutes
+of paired programming, 5 minutes to get groups to wrap up, and 10
+minutes to report back to the group. Remember that learners may need up
+to three times as long to complete a task that is simple for the
+presenter, so Code Club is best kept simple. We once had a presenter try
+to teach basic Julia syntax, but time was up by the time participants
+installed the interpreter. Some tips to help keep it simple are to limit
+the presented code chunks to less than 50 or, conversely, consider the
+number of lines that might be required to accomplish a solution.
 
-\hypertarget{rule-4-give-participants-time-to-prepare}{%
-\subsection{Rule 4: Give participants time to
-prepare}\label{rule-4-give-participants-time-to-prepare}}
+\subsection{Rule 6: Give participants time to
+prepare}\label{rule-6-give-participants-time-to-prepare}
 
 Similar to a Journal Club, the presenter should give participants a few
-days to prepare for the Code Club. Be sure to tell the participants what
-you will be covering. Provide the them with instructions on how to
-install dependencies, download data, and get the initial code. Make sure
-that the materials work together as intended. Considering the compressed
-schedule, asking participants to download materials beforehand is
-helpful. We have found that using separate GitHub repositories for each
-session can be a helpful way to get information to participants. Of
-course, using git and GitHub to engage in collaborative coding should
-probably be its own Code Club before using the tools more broadly with
-the group. An easier entry point is to have groups post their code to a
-Slack channel that we use to announce materials for our meetings.
+days (ideally a week) to prepare for the Code Club. Considering the
+compressed schedule described in Rule 5, asking participants to download
+materials beforehand is helpful to ensure a quick start. The presenter
+should provide the participants with instructions on how to install
+dependencies, download data, and get the initial code. This might also
+uncover weak points in the presenter's plan and enable them to ensure
+that the materials work as intended before the Code Club. We have found
+that using GitHub repositories for each Code Club can help make
+information, scripts, and data easily available to participants. Perhaps
+a first Code Club could be using git and GitHub to engage in
+collaborative coding. A lower barrier entry point is posting their code,
+data, and information in a lab meeting-dedicated Slack channel or
+sharing via email. Whatever method is used, the presenter should be sure
+to communicate the topic and necessary materials with the participants
+ahead of time.
 
-\hypertarget{rule-5-prepare-in-advance-to-maximize-your-participation}{%
-\subsection{Rule 5: Prepare in advance to maximize your
-participation}\label{rule-5-prepare-in-advance-to-maximize-your-participation}}
+\subsection{Rule 7: Don't give participants busy
+work}\label{rule-7-dont-give-participants-busy-work}
 
-It is the rare scientist that can fully participate in a Journal Club
-discussion about a paper that they have not read. Coming to Code Club
-without installing a dependency is like going to Journal Club and asking
-a simple question about the paper. It shows a lack of respect and
-preparation. Following Rule 4, review the code in advance, download the
-datasets, install the necessary packages, and perhaps read up on the
-topic. If your Code Club is reading a paper or chapters in a data
-science book be sure to read them before the session (e.g.~{[}5--7{]}).
-Consider how you could use the material for your own work.
+Participants want to learn topics that will either be useful to them or
+help their colleague (i.e., the presenter). Presenters should do their
+best to satisfy those motivations, whether it is through the relevancy
+of the concept or the data. It does not make sense to present a Code
+Club on downloading stock market data if it is not useful or interesting
+to the group. Similarly, participants should not be tasked with
+improving the presenter's code if the presenter has no intention of
+incorporating the suggestions. A list of packages or tasks that group
+members are interested in could help an uninspired presenter choose a
+topic that will make a rewarding Code Club.
 
-\hypertarget{rule-6-dont-give-your-participants-busy-work}{%
-\subsection{Rule 6: Don't give your participants busy
-work}\label{rule-6-dont-give-your-participants-busy-work}}
+\subsection{Rule 8: Include all levels of
+participants}\label{rule-8-include-all-levels-of-participants}
 
-Participants want to learn something because it will be useful to them
-or because they think it will help the presenter to improve their code.
-Presenters should do their best to satisfy those desires. It would not
-make much sense to hold a Code Club on downloading stock market data if
-the group was not interested in those data. At the same time,
-participants should not be asked to help the presenter improve their
-code if the presenter has no intention of incorporating the suggestions.
-It would be useful to keep a list of packages or tasks that members of
-the group want to learn about. That way if a presenter is having a hard
-time coming up with ideas, they can take something from the list and
-present it.
+As suggested by Rule 7, a significant challenge to presenting at Code
+Club is selecting topics and activities that appeal to a critical mass
+of the participants. This is particularly difficult if participants have
+a wide range of coding experience, such as follows a turnover in group
+membership. In these circumstances it can be useful to cover commonly
+used tools but that might result in disinterest of more experienced
+participants. We have identified several strategies to overcome the
+challenges presented by participants at varying skill levels. Instead of
+letting participants form their own pairs, the presenter can select
+pairs of participants with either similar or differing skill levels,
+depending on their goals. Partnerships between those with similar skill
+levels requires the presenter to design appropriate activites each skill
+level. We have found that commenting code is a good skill for beginners
+since it forces them to dissect and understand a code chunk
+line-by-line. An advantage of partnerships with disparate skill levels
+is assigning a single task, which provides the presenter with a diverse
+range of methods that achieve the same result. This approach to pairing
+also helps to graft new members into the group that have emerging
+programming skills. Regardless of how partners are selected, consider
+asking the pairs to identify a navigator and a driver {[}5{]}. The
+driver types at the computer while the navigator tells them what to
+type, thus ensuring participation of both partners. Midway through the
+activity, the presenter can have the partners switch roles.
+Intentionally forming pairs can also engineer group interactions by
+breaking up cliques, avoiding potentially disruptive partnerships, or
+pairing reliable role models with new group members.
 
-\hypertarget{rule-7-include-all-levels-of-participants}{%
-\subsection{Rule 7: Include all levels of
-participants}\label{rule-7-include-all-levels-of-participants}}
+\subsection{Rule 9: Prepare in advance to maximize
+participation}\label{rule-9-prepare-in-advance-to-maximize-participation}
 
-A significant challenge to presenting at Code Club is to pick topics and
-activities that appeal to a critical mass of the participants. For
-example, if there has recently been a large turn over in group
-membership, it can be useful to go back to the basics and cover commonly
-used tools. We have identified several strategies to overcome the
-challenges of having varying skill levels. As an alternative to letting
-participants form their own pairs, the presenter can be purposeful in
-how they form pairs. Depending on the goals of the session, forming
-pairs between participants that have similar or different skills can be
-effective. If pairs are formed where partners have similar skill levels,
-then the presenter needs to design activities that are appropriate to
-different levels of participants. We have found that a good activity for
-beginners is to have them comment code, which forces them to dissect a
-code chunk and understand what each line does. An advantage of forming
-pairs with disparate skill levels is that all of the pairs can be given
-the same task leading to a diversity of methods to achieve the same
-result. This approach to pairing also helps graft new members into the
-group that have emerging programming skills. Regardless of how partners
-are selected, asking the pairs to identify a navigator and driver is
-helpful {[}8{]}. The driver types at the computer while the navigator
-tells them what to type. Midway through the activity the presenter can
-tell the partners to switch roles. Intentionally forming pairs can also
-be a mechanism to engineer interactions by breaking up cliques, avoiding
-potentially toxic combinations of members, or pairing reliable role
-models with new members of the group.
+It is not possible to fully participate in a Journal Club discussion
+about a paper that the participant has not read. In that context, coming
+to Code Club without having installed a dependency is similar to asking
+a simple question about the Journal Club paper. Both instances show a
+lack of respect and preparation. Just as a presenter must follow Rule 6
+to provide materials ahead of time, participants must review the code in
+advance, download the data sets, install the necessary packages, and
+perhaps read up on the topic. If the Code Club is based on a paper or
+chapters in a data science book (e.g., {[}6--8{]}) the participants
+should read them before the session and consider how they might
+incorporate the concepts into their own work.
 
-\hypertarget{rule-8-participate}{%
-\subsection{Rule 8: Participate}\label{rule-8-participate}}
+\subsection{Rule 10: Participate}\label{rule-10-participate}
 
-Having an open laptop on the table and permission to use it can feel
-like an invitation to get distracted by other work, emails, and browsing
-the internet. Fight that urge and focus on the presenter's goals. Allow
-your partner to contribute. Speak up for yourself and force your partner
-to let you contribute. Even if the material seems too advanced for you,
-programming languages like R and Python are generally expressive and
-should allow you to engage with the logic, even if the syntax is too
-hard. Frankly, understanding the logic of when to use one modeling
-approach over another is more important than how to use the modeling
-approach. If you understand the ``why'', the ``how'' will generally
-quickly follow. If you are a more experienced participant, give your
-feedback and coding suggestions at a level that seeks to inform
-everyone. Regardless of your skill level, your partner and the presenter
-are putting themselves at risk by revealing what they do or do not know.
-Encourage them and show your gratitude for helping you learn something
-new.
+An essential ingredient of any Code Club is active participation from
+all parties. Having an open laptop on the table and permission to use it
+can feel like an invitation to get distracted by other work, emails, and
+browsing the internet; fight that urge and focus on the presenter's
+goals. Be respectful and allow your partner to contribute. Speak up for
+yourself and force your partner to let you contribute. If the material
+seems too advanced for you, it can be frustrating, and tempting to
+mentally check out. Fortuntately, programming languages like R and
+Python are generally expressive, which should allow you to engage with
+the logic, even if the syntax is too advanced. Frankly, understanding
+the logic of when to use one modeling approach over another is more
+important than how to use it. If you understand the ``why'', the ``how''
+will quickly follow. More experienced participants should aim to
+communicate feedback and coding suggestions at a level that all
+participants can understand and engage in. Regardless of skill levels,
+your partner and the presenter put themselves at risk by revealing what
+they do or do not know. Encourage them, and show your gratitude for
+helping you learn something new, by fully participating in each Code
+Club.
 
-\hypertarget{rule-9-let-the-material-change-you}{%
-\subsection{Rule 9: Let the material change
-you}\label{rule-9-let-the-material-change-you}}
-
-Part of the humility required to participate in a Code Club is both
-acknowledging your incomplete training and that learning something new
-is possible. Assume that the presenter has followed the other rules and
-that they think you should be using what they presented. After the Code
-Club, try to incorporate that material into your new code or by
-refactoring your old code. By practicing the material in a different
-context you will learn the material better. You may identify concepts
-that you are unsure of, which gives you another opportunity to ask the
-presenter for help.
-
-\hypertarget{rule-10-experiment}{%
-\subsection{Rule 10: Experiment!}\label{rule-10-experiment}}
-
-The selection of content and structure for each Code Club is determined
-by the presenter making the process democratic and distributed. If
-someone thinks something is worth learning and wants to teach it, they
-have that power. If they want to experiment with a different format,
-they are free to try it out. Members of the Code Club need to feel like
-they have the power to shape the direction of the group. If members are
-following these Rules, they will naturally reflect on the skills and
-interests of the other members in the group. For example, there is
-always turn over in a research group and thus it is important to make
-sure that there are opportunities to go back to things people consider
-``simple'' to teach the content to new people or provide a refresher to
-people that perhaps have been at the lab bench more than at the computer
-recently. The group should also feel free to experiment with the format.
-If we try a different format, then it is useful to give everyone an
-opportunity to debrief on what they did or did not like with the new
-approach.
-
-\hypertarget{conclusion}{%
-\subsection{Conclusion}\label{conclusion}}
+\subsection{Conclusion}\label{conclusion}
 
 The most important rules are the first and last. Members of the Code
-Club need to feel comfortable with other members of the group and
-sufficiently empowered to try something new or to ask for help. Aside
-from growing in our programming skills, we have noticed two other
-benefits that point to the efforts of creating a positive culture.
-First, we have intentionally interviewed postdoc candidates on the days
-we have held Code Club. Although it is important to make it clear that
-they are not being assessed on their coding skills, it is useful to see
-how the candidate interacts with other members of the research group in
-a relatively low stakes environment. At the same time, the candidate can
-learn about the culture of the research group. Second, members of other
-research groups where they are isolated in their desire to grow in these
-skills have integrated themselves into our Code Club. This speaks to the
-broader need for Code Club and the likelihood that it would work when
-expanded to a larger group of individuals that are not necessarily in
-the same research group. Ultimately, our Code Club has helped
-researchers feel more comfortable seeking out their colleagues outside
-of Code Club sessions for help in improving their code.
+Club need to feel comfortable with other group members and sufficiently
+empowered to try something new or ask for help. Aside from growing in
+our programming skills, we have noticed two other benefits that help
+create a positive work culture. First, we have intentionally interviewed
+postdoc candidates on the days we hold Code Club. We make it clear that
+they are not being assessed on their coding skills, but instead use it
+as an opportunity to see how the candidate interacts with other members
+of the research group in a relatively low stakes environment. At the
+same time, the candidate can learn about the culture of the research
+group through active participation. Second, members of other research
+groups that feel isolated in growing their skills have integrated
+themselves into our Code Club. This speaks to both the broader need for
+Code Club and the likelihood of success when expanded to include a
+larger group of individuals with broader research interests. Ultimately,
+Code Club has improved the overall data analysis skills, community, and
+research success of our lab by empowering researchers to seek help from
+their colleagues.
 
-\hypertarget{acknowledgements}{%
-\subsection{Acknowledgements}\label{acknowledgements}}
+\subsection{Acknowledgements}\label{acknowledgements}
 
@@
 PDS's lab meetings. They have had a critical role in shaping the
-evolution of the sessions. The order of co-authors was determined
-by\ldots{}
+evolution of the sessions. AKH led a lab discussion of Code Clubs to
+identify the key concepts presented, which were supplemented by
+contributions from NL. The order of co-authors was determined by\ldots{}
 
@@
 
-\hypertarget{references}{%
-\subsection{References}\label{references}}
+\subsection{References}\label{references}
 
 \hypertarget{refs}{}
-\begin{cslreferences}
-\leavevmode\hypertarget{ref-Lonsdale2016}{}%
+\hypertarget{ref-Lonsdale2016}{}
 1. Lonsdale A, Penington JS, Rice T, Walker M, Dashnow H. Ten simple
@@
 
-\leavevmode\hypertarget{ref-Wilson2014}{}%
+\hypertarget{ref-Wilson2014}{}
 2. Wilson G, Aruliah DA, Brown CT, Hong NPC, Davis M, Guy RT, et al.
@@
 
-\leavevmode\hypertarget{ref-Kothiyal2014}{}%
+\hypertarget{ref-Kothiyal2014}{}
 3. Kothiyal A, Murthy S, Iyer S. Think-pair-share in a large CS1 class.
@@
 
-\leavevmode\hypertarget{ref-StackOverflow_MREE}{}%
+\hypertarget{ref-StackOverflow_MREE}{}
 4. Stack Overflow. How to create a minimal, reproducible example. 2020.
@@
 
-\leavevmode\hypertarget{ref-Wickham_2016}{}%
-5. Wickham H, Grolemund G. O'Reilly Media; 2016. Available:
+\hypertarget{ref-Bockeler_2020}{}
+5. Böckeler B, Siessegger N. On pair programming. 2020. Available:
+\url{https://martinfowler.com/articles/on-pair-programming.html}
+
+\hypertarget{ref-Wickham_2016}{}
+6. Wickham H, Grolemund G. O'Reilly Media; 2016. Available:
 \url{https://r4ds.had.co.nz}
 
-\leavevmode\hypertarget{ref-Schloss_2018}{}%
-6. Schloss PD. The riffomonas reproducible research tutorial series.
+\hypertarget{ref-Schloss_2018}{}
+7. Schloss PD. The riffomonas reproducible research tutorial series.
 Journal of Open Source Education. 2018;1: 13.
@@
 
-\leavevmode\hypertarget{ref-Noble2009}{}%
-7. Noble WS. A quick guide to organizing computational biology projects.
+\hypertarget{ref-Noble2009}{}
+8. Noble WS. A quick guide to organizing computational biology projects.
 Lewitter F, editor. PLoS Computational Biology. 2009;5: e1000424.
 doi:\href{https://doi.org/10.1371/journal.pcbi.1000424}{10.1371/journal.pcbi.1000424}
-
-\leavevmode\hypertarget{ref-Bockeler_2020}{}%
-8. Böckeler B, Siessegger N. On pair programming. 2020. Available:
-\url{https://martinfowler.com/articles/on-pair-programming.html}
-\end{cslreferences}
 
@@
 \toprule
-\begin{minipage}[b]{0.25\columnwidth}\raggedright
+\begin{minipage}[b]{0.24\columnwidth}\raggedright\strut
 \textbf{Title}\strut
-\end{minipage} & \begin{minipage}[b]{0.69\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[b]{0.61\columnwidth}\raggedright\strut
 \textbf{Description}\strut
@@
 \endhead
-\begin{minipage}[t]{0.25\columnwidth}\raggedright
+\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
 base vs.~ggplot\strut
-\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
 Given input data and a figure, recreate the figure using R's base
@@
 \end{minipage}\tabularnewline
-\begin{minipage}[t]{0.25\columnwidth}\raggedright
+\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
 Snakemake\strut
-\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
 Given a bash script that contains an analysis pipeline, convert it to a
-Snakemake workflow (can also be done with GNU Make)\strut
+series of recipes (can also be done with GNU Make)\strut
 \end{minipage}\tabularnewline
-\begin{minipage}[t]{0.25\columnwidth}\raggedright
+\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
 DRYing code\strut
-\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
 Given script with repeated code, create functions to remove
@@
 \end{minipage}\tabularnewline
-\begin{minipage}[t]{0.25\columnwidth}\raggedright
+\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
 mothur and Vegan\strut
-\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
 Given a pairwise community dissimilarity matrix, compare communities
@@
 \end{minipage}\tabularnewline
-\begin{minipage}[t]{0.25\columnwidth}\raggedright
+\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
 tidy data\strut
-\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
 Given a wide-formatted data table, convert it to a long, tidy-formatted
@@
 \end{minipage}\tabularnewline
-\begin{minipage}[t]{0.25\columnwidth}\raggedright
+\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
 GitFlow\strut
-\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
 Participants file and claim an issue to add their name to a README file
@@
 \end{minipage}\tabularnewline
-\begin{minipage}[t]{0.25\columnwidth}\raggedright
-Integrating R and google docs\strut
-\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
-Scrape a google docs workbook and clean the data to identify previous
+\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
+Integrating R and Google docs\strut
+\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
+Scrape a Google docs workbook and clean the data to identify previous
 Code Club presenters\strut

### 2020-04-07 submission/manuscript.Rmd
Incorporate edits from AKH, KLS, BT, and PDS

@@
 
-\vspace{10mm}
+\vspace*{5mm}
 
@@
 
-\vspace{35mm}
+\vspace{25mm}
 
-Ada K. Hagan${^2}$, Nick Lesniak, *Author names here, order TBD*
-
-Patrick D. Schloss${^1}$${^\dagger}$
+Ada K. Hagan${^2}$, Nick Lesniak, *Author names here, order alphabetical*, Patrick D. Schloss${^1}$${^\dagger}$
 
@@
 
-We continue to experiment with approaches to running Code Club, but have learned that it is critical for the presenter to clearly articulate their goals and facilitate participant engagement.
+We continue to experiment with approaches to running Code Club.
+During the period of the Covid-19 pandemic in 2020 when many research labs were shuttered, PDS started hosting weekly Code Club virtually.
+Participants could engage the material posted to https://www.riffomonas.org/code_club/ via a Zoom meeting or asynchronously using the recorded sessions available on YouTube.
+Regardless of our approach to leading a Code Club, we have learned that it is critical for the presenter to clearly articulate their goals and facilitate participant engagement.
 Although any individual Code Club session may be more experimental than another, on the whole they are a critical tool to train bench scientists in reproducible data analysis practices.
-We have provided some examples of successful code club topics in Table 1 and summarized the results of our own experimentation as Code Club presenters and participants into Ten Rules.
+We have provided some examples of successful code club topics in Table 1.
+We have summarized the results of our own experimentation as Code Club presenters and participants into Ten Rules.
 The first 3 rules apply to all participants at every Code Club, while Rules 4 through 8 are targeted to Code Club presenters and the last two for participants.
@@
 Each member of the Code Club must have the humility to acknowledge that they have more to learn about any given topic.
-We have found that many problems are avoided when the presenter takes charge of the session with a clear lesson plan, thoughtful group creation, and constructive encouragement.
+We have found that many problems are avoided when the presenter takes charge of the session with a clear lesson plan, thoughtfully creating groups, and giving encouragement.
 Similarly, participants foster a positive environment by remembering that the task is not a competition, focusing on the presenter's goals, allowing their partner to contribute, asking clarifying questions when appropriate, and avoiding distractions (e.g.
@@
 After the Code Club, try to incorporate that material into new code or refactor old code.
-By practicing the material in a different context you will learn the material better.
+By practicing the material in a different context, you will learn the material better.
 For presenters, incorporate suggested changes into your code.
@@
 Instead of letting participants form their own pairs, the presenter can select pairs of participants with either similar or differing skill levels, depending on their goals.
-Partnerships between those with similar skill levels requires the presenter to design appropriate activites each skill level.
+Partnerships between those with similar skill levels requires the presenter to design appropriate activities each skill level.
 We have found that commenting code is a good skill for beginners since it forces them to dissect and understand a code chunk line-by-line.
@@
 Just as a presenter must follow Rule 6 to provide materials ahead of time, participants must review the code in advance, download the data sets, install the necessary packages, and perhaps read up on the topic.
-If the Code Club is based on a paper or chapters in a data science book (e.g., [@Wickham_2016; @Schloss_2018; @Noble2009]) the participants should read them before the session and consider how they might incorporate the concepts into their own work.
+If the Code Club is based on a paper or chapters in a data science book (e.g., [@Wickham_2016; @Schloss_2018; @Noble2009]), the participants should read them before the session and consider how they might incorporate the concepts into their own work.
 
@@
 If the material seems too advanced for you, it can be frustrating, and tempting to mentally check out.
-Fortuntately, programming languages like R and Python are generally expressive, which should allow you to engage with the logic, even if the syntax is too advanced.
-Frankly, understanding the logic of when to use one modeling approach over another is more important than how to use it.
+Fortunately, programming languages like R and Python are generally expressive, which should allow you to engage with the logic, even if the syntax is too advanced.
+Frankly, understanding the logic of *when* to use one modeling approach over another is more important than knowing *how* to use it.
 If you understand the "why", the "how" will quickly follow.
 More experienced participants should aim to communicate feedback and coding suggestions at a level that all participants can understand and engage in.
-Regardless of skill levels, your partner and the presenter put themselves at risk by revealing what they do or do not know.
+Regardless of skill levels, your partner and the presenter put themselves in a vulnerable position by revealing what they do or do not know.
 Encourage them, and show your gratitude for helping you learn something new, by fully participating in each Code Club.
@@
 AKH led a lab discussion of Code Clubs to identify the key concepts presented, which were supplemented by contributions from NL.
-The order of co-authors was determined by...
+The order of other co-authors is listed alphabetically.
 
@@
 | GitFlow           | Participants file and claim an issue to add their name to a README file in a GitHub-hosted repository and file a pull request to complete the issue |
-| Integrating R and Google docs   | Scrape a Google docs workbook and clean the data to identify previous Code Club presenters |
+| R with Google docs   | Scrape a Google docs workbook and clean the data to identify previous Code Club presenters |
+| Develop an R Package  | Convert a lab member's collection of scripts into package over a series of sessions  |

### 2020-04-07 submission/manuscript.tex
Incorporate edits from AKH, KLS, BT, and PDS

@@
-\documentclass[11pt,]{article}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+  11pt,
+]{article}
 \usepackage{lmodern}
@@
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
 \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
+\IfFileExists{microtype.sty}{% use microtype if available
   \usepackage[]{microtype}
@@
 }{}
-\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
-\usepackage[unicode=true]{hyperref}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \hypersetup{
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1.0in]{geometry}
 \usepackage{longtable,booktabs}
-% Fix footnotes in tables (requires footnote package)
-\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx,grffile}
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{helvet} % Helvetica font
@@
 
-\vspace{10mm}
+\vspace*{5mm}
 
+\hypertarget{ten-simple-rules-to-increase-computational-skills-among-biologists-with-code-clubs}{%
 \section{Ten Simple Rules to increase computational skills among
 biologists with Code
-Clubs}\label{ten-simple-rules-to-increase-computational-skills-among-biologists-with-code-clubs}
+Clubs}\label{ten-simple-rules-to-increase-computational-skills-among-biologists-with-code-clubs}}
 
-\vspace{35mm}
+\vspace{25mm}
 
-Ada K. Hagan\({^2}\), Nick Lesniak, \emph{Author names here, order TBD}
-
-Patrick D. Schloss\({^1}\)\({^\dagger}\)
+Ada K. Hagan\({^2}\), Nick Lesniak, \emph{Author names here, order
+alphabetical}, Patrick D. Schloss\({^1}\)\({^\dagger}\)
 
@@
 
-\subsubsection{Ten Simple Rules}\label{ten-simple-rules}
+\hypertarget{ten-simple-rules}{%
+\subsubsection{Ten Simple Rules}\label{ten-simple-rules}}
 
 \newpage
-
 \linenumbers
 
-\subsection{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\subsection{Introduction}\label{introduction}}
 
@@
 
-We continue to experiment with approaches to running Code Club, but have
-learned that it is critical for the presenter to clearly articulate
-their goals and facilitate participant engagement. Although any
-individual Code Club session may be more experimental than another, on
-the whole they are a critical tool to train bench scientists in
-reproducible data analysis practices. We have provided some examples of
-successful code club topics in Table 1 and summarized the results of our
-own experimentation as Code Club presenters and participants into Ten
-Rules. The first 3 rules apply to all participants at every Code Club,
-while Rules 4 through 8 are targeted to Code Club presenters and the
-last two for participants.
+We continue to experiment with approaches to running Code Club. During
+the period of the Covid-19 pandemic in 2020 when many research labs were
+shuttered, PDS started hosting weekly Code Club virtually. Participants
+could engage the material posted to
+\url{https://www.riffomonas.org/code_club/} via a Zoom meeting or
+asynchronously using the recorded sessions available on YouTube.
+Regardless of our approach to leading a Code Club, we have learned that
+it is critical for the presenter to clearly articulate their goals and
+facilitate participant engagement. Although any individual Code Club
+session may be more experimental than another, on the whole they are a
+critical tool to train bench scientists in reproducible data analysis
+practices. We have provided some examples of successful code club topics
+in Table 1. We have summarized the results of our own experimentation as
+Code Club presenters and participants into Ten Rules. The first 3 rules
+apply to all participants at every Code Club, while Rules 4 through 8
+are targeted to Code Club presenters and the last two for participants.
 
+\hypertarget{rule-1-reciprocate-respect}{%
 \subsection{Rule 1: Reciprocate
-respect}\label{rule-1-reciprocate-respect}
+respect}\label{rule-1-reciprocate-respect}}
 
@@
 presenter takes charge of the session with a clear lesson plan,
-thoughtful group creation, and constructive encouragement. Similarly,
+thoughtfully creating groups, and giving encouragement. Similarly,
 participants foster a positive environment by remembering that the task
@@
 
+\hypertarget{rule-2-let-the-material-change-you}{%
 \subsection{Rule 2: Let the material change
-you}\label{rule-2-let-the-material-change-you}
+you}\label{rule-2-let-the-material-change-you}}
 
@@
 Club, try to incorporate that material into new code or refactor old
-code. By practicing the material in a different context you will learn
+code. By practicing the material in a different context, you will learn
 the material better. For presenters, incorporate suggested changes into
@@
 
-\subsection{Rule 3: Experiment!}\label{rule-3-experiment}
+\hypertarget{rule-3-experiment}{%
+\subsection{Rule 3: Experiment!}\label{rule-3-experiment}}
 
@@
 
-\subsection{Rule 4: Set specific goals}\label{rule-4-set-specific-goals}
+\hypertarget{rule-4-set-specific-goals}{%
+\subsection{Rule 4: Set specific
+goals}\label{rule-4-set-specific-goals}}
 
@@
 
-\subsection{Rule 5: Keep it simple}\label{rule-5-keep-it-simple}
+\hypertarget{rule-5-keep-it-simple}{%
+\subsection{Rule 5: Keep it simple}\label{rule-5-keep-it-simple}}
 
@@
 
+\hypertarget{rule-6-give-participants-time-to-prepare}{%
 \subsection{Rule 6: Give participants time to
-prepare}\label{rule-6-give-participants-time-to-prepare}
+prepare}\label{rule-6-give-participants-time-to-prepare}}
 
@@
 
+\hypertarget{rule-7-dont-give-participants-busy-work}{%
 \subsection{Rule 7: Don't give participants busy
-work}\label{rule-7-dont-give-participants-busy-work}
+work}\label{rule-7-dont-give-participants-busy-work}}
 
@@
 
+\hypertarget{rule-8-include-all-levels-of-participants}{%
 \subsection{Rule 8: Include all levels of
-participants}\label{rule-8-include-all-levels-of-participants}
+participants}\label{rule-8-include-all-levels-of-participants}}
 
@@
 depending on their goals. Partnerships between those with similar skill
-levels requires the presenter to design appropriate activites each skill
-level. We have found that commenting code is a good skill for beginners
-since it forces them to dissect and understand a code chunk
+levels requires the presenter to design appropriate activities each
+skill level. We have found that commenting code is a good skill for
+beginners since it forces them to dissect and understand a code chunk
 line-by-line. An advantage of partnerships with disparate skill levels
@@
 
+\hypertarget{rule-9-prepare-in-advance-to-maximize-participation}{%
 \subsection{Rule 9: Prepare in advance to maximize
-participation}\label{rule-9-prepare-in-advance-to-maximize-participation}
+participation}\label{rule-9-prepare-in-advance-to-maximize-participation}}
 
@@
 perhaps read up on the topic. If the Code Club is based on a paper or
-chapters in a data science book (e.g., {[}6--8{]}) the participants
+chapters in a data science book (e.g., {[}6--8{]}), the participants
 should read them before the session and consider how they might
@@
 
-\subsection{Rule 10: Participate}\label{rule-10-participate}
+\hypertarget{rule-10-participate}{%
+\subsection{Rule 10: Participate}\label{rule-10-participate}}
 
@@
 seems too advanced for you, it can be frustrating, and tempting to
-mentally check out. Fortuntately, programming languages like R and
-Python are generally expressive, which should allow you to engage with
-the logic, even if the syntax is too advanced. Frankly, understanding
-the logic of when to use one modeling approach over another is more
-important than how to use it. If you understand the ``why'', the ``how''
-will quickly follow. More experienced participants should aim to
-communicate feedback and coding suggestions at a level that all
-participants can understand and engage in. Regardless of skill levels,
-your partner and the presenter put themselves at risk by revealing what
-they do or do not know. Encourage them, and show your gratitude for
-helping you learn something new, by fully participating in each Code
-Club.
+mentally check out. Fortunately, programming languages like R and Python
+are generally expressive, which should allow you to engage with the
+logic, even if the syntax is too advanced. Frankly, understanding the
+logic of \emph{when} to use one modeling approach over another is more
+important than knowing \emph{how} to use it. If you understand the
+``why'', the ``how'' will quickly follow. More experienced participants
+should aim to communicate feedback and coding suggestions at a level
+that all participants can understand and engage in. Regardless of skill
+levels, your partner and the presenter put themselves in a vulnerable
+position by revealing what they do or do not know. Encourage them, and
+show your gratitude for helping you learn something new, by fully
+participating in each Code Club.
 
-\subsection{Conclusion}\label{conclusion}
+\hypertarget{conclusion}{%
+\subsection{Conclusion}\label{conclusion}}
 
@@
 
-\subsection{Acknowledgements}\label{acknowledgements}
+\hypertarget{acknowledgements}{%
+\subsection{Acknowledgements}\label{acknowledgements}}
 
@@
 identify the key concepts presented, which were supplemented by
-contributions from NL. The order of co-authors was determined by\ldots{}
+contributions from NL. The order of other co-authors is listed
+alphabetically.
 
@@
 
-\subsection{References}\label{references}
+\hypertarget{references}{%
+\subsection{References}\label{references}}
 
 \hypertarget{refs}{}
-\hypertarget{ref-Lonsdale2016}{}
+\leavevmode\hypertarget{ref-Lonsdale2016}{}%
 1. Lonsdale A, Penington JS, Rice T, Walker M, Dashnow H. Ten simple
@@
 
-\hypertarget{ref-Wilson2014}{}
+\leavevmode\hypertarget{ref-Wilson2014}{}%
 2. Wilson G, Aruliah DA, Brown CT, Hong NPC, Davis M, Guy RT, et al.
@@
 
-\hypertarget{ref-Kothiyal2014}{}
+\leavevmode\hypertarget{ref-Kothiyal2014}{}%
 3. Kothiyal A, Murthy S, Iyer S. Think-pair-share in a large CS1 class.
@@
 
-\hypertarget{ref-StackOverflow_MREE}{}
+\leavevmode\hypertarget{ref-StackOverflow_MREE}{}%
 4. Stack Overflow. How to create a minimal, reproducible example. 2020.
@@
 
-\hypertarget{ref-Bockeler_2020}{}
+\leavevmode\hypertarget{ref-Bockeler_2020}{}%
 5. Böckeler B, Siessegger N. On pair programming. 2020. Available:
@@
 
-\hypertarget{ref-Wickham_2016}{}
+\leavevmode\hypertarget{ref-Wickham_2016}{}%
 6. Wickham H, Grolemund G. O'Reilly Media; 2016. Available:
@@
 
-\hypertarget{ref-Schloss_2018}{}
+\leavevmode\hypertarget{ref-Schloss_2018}{}%
 7. Schloss PD. The riffomonas reproducible research tutorial series.
@@
 
-\hypertarget{ref-Noble2009}{}
+\leavevmode\hypertarget{ref-Noble2009}{}%
 8. Noble WS. A quick guide to organizing computational biology projects.
@@
 \toprule
-\begin{minipage}[b]{0.24\columnwidth}\raggedright\strut
+\begin{minipage}[b]{0.25\columnwidth}\raggedright
 \textbf{Title}\strut
-\end{minipage} & \begin{minipage}[b]{0.61\columnwidth}\raggedright\strut
+\end{minipage} & \begin{minipage}[b]{0.69\columnwidth}\raggedright
 \textbf{Description}\strut
@@
 \endhead
-\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
 base vs.~ggplot\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
 Given input data and a figure, recreate the figure using R's base
@@
 \end{minipage}\tabularnewline
-\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
 Snakemake\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
 Given a bash script that contains an analysis pipeline, convert it to a
-series of recipes (can also be done with GNU Make)\strut
+Snakemake workflow (can also be done with GNU Make)\strut
 \end{minipage}\tabularnewline
-\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
 DRYing code\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
 Given script with repeated code, create functions to remove
@@
 \end{minipage}\tabularnewline
-\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
 mothur and Vegan\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
 Given a pairwise community dissimilarity matrix, compare communities
@@
 \end{minipage}\tabularnewline
-\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
 tidy data\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
 Given a wide-formatted data table, convert it to a long, tidy-formatted
@@
 \end{minipage}\tabularnewline
-\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
 GitFlow\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
 Participants file and claim an issue to add their name to a README file
@@
 \end{minipage}\tabularnewline
-\begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
-Integrating R and Google docs\strut
-\end{minipage} & \begin{minipage}[t]{0.61\columnwidth}\raggedright\strut
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
+R with Google docs\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
 Scrape a Google docs workbook and clean the data to identify previous
 Code Club presenters\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
+Develop an R Package\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+Convert a lab member's collection of scripts into package over a series
+of sessions\strut
 \end{minipage}\tabularnewline

### 2020-04-28 submission/manuscript.Rmd
First submission

@@
 
-\vspace*{5mm}
-
 # Ten Simple Rules to increase computational skills among biologists with Code Clubs
 
-\vspace{25mm}
+\vspace{5mm}
 
-Ada K. Hagan${^2}$, Nick Lesniak, *Author names here, order alphabetical*, Patrick D. Schloss${^1}$${^\dagger}$
+Ada K.
+Hagan${^{1,2}}$, Nicholas A.
+Lesniak${^1}$, Marcy J.
+Balunas${^3}$, Lucas Bishop${^1}$, William L.
+Close${^1}$, Matthew D.
+Doherty${^1}$, Amanda G.
+Elmore${^1}$, Kaitlin J.
+Flynn${^{1,4}}$, Geoffrey D.
+Hannigan${^{1,5}}$, Charlie C.
+Koumpouras${^{1,6}}$, Matthew L.
+Jenior${^{1,7}}$, Ariangela J.
+Kozik${^8}$, Kathryn McBride${^1}$, Samara B.
+Rifkin${^8}$, Joshua M.A.
+Stough${^1}$, Kelly L.
+Sovacool${^9}$, Marc A.
+Sze${^{1,10}}$, Sarah Tomkovich${^1}$, Begum D.
+Topcuoglu${^{1,5}}$, and Patrick D.
+Schloss${^1}$${^\dagger}$
 
-\vspace{40mm}
+\vspace{5mm}
 
@@
 
-$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109
+1. Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
+2. Current affiliation: Alliance SciComm & Consulting, Linden, MI
+3. Division of Medicinal Chemistry, Department of Pharmaceutical Sciences, University of Connecticut, Storrs, CT
+4. Current affiliation: Benaroya Research Institute, Seattle, WA
+5. Current affiliation: Exploratory Science Center, Merck & Co., Inc., Cambridge, Massachusetts, USA
+6. Current affiliation: Roche Diagnostics, Clinical Operations Services & eSystems, Indianapolis, IN
+7. Current affiliation: Department of Biomedical Engineering, University of Virginia, Charlottesville, VA
+8. Department of Internal Medicine, University of Michigan
+9. Department of Computational Medicine & Bioinformatics, University of Michigan
+10. Current affiliation: Genomics and Pharmacogenomics, Merck & Co., Inc., Cambridge, MA
 
-$2$ Alliance SciComm & Consulting, Linden, MI 48418
-
-\vspace{35mm}
+\vspace{5mm}
 
@@
 High throughput data comes to us from DNA and RNA sequencing, flow cytometry, metabolomics, molecular screens, and more.
-Although some accept the approach of compartmentalizing data generation and data analysis, we have found scientists feel empowered when they can both ask and answer their own biological questions.
-Yet, the standard undergraduate and graduate training in the biological sciences is insufficient for developing these advanced data analysis skills.
+Although some accept the approach of compartmentalizing data generation and data analysis, we have found scientists feel empowered when they can both ask and answer their own biological questions [@Carey2018].
+Yet, the standard undergraduate and graduate training in the biological sciences is typically insufficient for developing these advanced data analysis skills.
 In our experience performing microbiome research, it is more common to find exceptional bench scientists who are inexperienced at analyzing large data sets than to find the reverse.
-Of course this raises a challenge: How do we train bench scientists that analyze data sets to answer biological questions?
+Of course this raises a challenge: How do we train bench scientists to effectively answer biological questions with these larger datasets?
 
-The ever-growing ability to generate data and constant feeling of helplessness in analyzing it is analogous to the struggles we also face with engaging the voluminous scientific literature.
-A common strategy to keeping up with the literature are journal clubs, which involve group discussion of a pre-selected paper.
+The ever-growing ability to generate data and feeling of unease in analyzing the data are analogous to the struggles scientists also face with engaging the voluminous scientific literature.
+A common strategy for keeping up with the literature is participating in journal clubs, which involve group discussion of a pre-selected paper.
 These range from informal discussions to PowerPoint presentations and course credit.
-In addition to staying current on the literature, journal clubs help strengthen skills in critical thinking, communication, and literature review [@Lonsdale2016].
-Our research group has leveraged the similarity between the overwhelming natures of scientific literature and data analysis to address the challenge of teaching reproducible data analysis practices to bench scientists.
-Over the past 4 years we have experimented with a Code Club model to improve data analysis skills in a community environment.
+In addition to staying current on the literature, journal clubs help strengthen skills in critical thinking, communication, and integrating the literature [@Lonsdale2016].
+Our research group has leveraged the similarity between the overwhelming nature of both scientific literature and data analysis to address the challenge of teaching reproducible data analysis practices to bench scientists.
+Over the past four years we have experimented with a Code Club model to improve data analysis skills in a community environment.
 
 Our Code Club sessions are an hour long and alternate with Journal Club as the second part of weekly two hour lab meetings.
-Presenters volunteer to lead a session with the expectation that they alternate leading Journal and Code Clubs.
 Initially, the Code Club was used to review code from trainee projects.
-Instead of a presentation, the presenter would project their code onto a screen and the participants would go line by line through the code, stating the logic behind each line.
+Instead of a presentation, the presenter would project their code onto a screen and the participants would go through the code, stating the logic behind each line.
 This approach emphasized the importance of code readability and gave beginners the opportunity to see the real-life, messy code of more experienced peers.
-Unfortunately, the format only allowed us to review a fraction of a project's code base, making it difficult to integrate the programmer's logic across their full project.
-A major issue with this model was that sometimes beginners couldn't contribute to improving the code, and even when they could, more senior group members would eventually dominate the discussion.
-This led to a lack of participation by beginners, who would mentally check out to focus on other activities, and frequently resulted in an adversarial environment between the presenter and senior group members.
-As a result, presenters rarely offered to present Code Club again.
+Unfortunately, the format only allowed us to review a fraction of a project's code, making it difficult to integrate the programmer's logic across their full project.
+A major issue with this model was that sometimes beginners could not contribute to improving the code, and even when they could, more experienced group members would eventually dominate the discussion.
+This led to a lack of participation by beginners, who would sometimes mentally check out and resulted in an adversarial environment between the presenter and more experienced members.
+As a result, presenters were reluctant to offer to present again.
 
-From these experiences, we began a collective conversation to improve our Code Club model and have evolved it into two approaches.
+From these experiences, we began a collective conversation to improve our Code Club model.
+We have identified two successful approaches.
 The first is a more constructive version of a group code review.
@@
 The challenge for the session would be to convert the code chunk into a function to be called throughout the script to make it "DRY" (i.e.
-Don't Repeat Yourself)[@Wilson2014].
+Don't Repeat Yourself [@Wilson2014]).
 The presenter leaves the session with several partial or working solutions to their problem and the importance of writing DRY code is reinforced.
@@
 The presenter introduces a new package or technique and assigns an activity to practice the new approach.
-For example, a more memorable activity was giving participants raw data and a finished plot.
-Paired participants were tasked with generating the plot using either base R or ggplot, but base R users had to use ggplot and vice versa.
-In either approach, the Code Club ends with a report back to the presenter and larger group describing the approach each pair took.
+For example, at one Code Club participants were given raw data and a finished plot.
+Paired participants were tasked with generating the plot from the data using either R syntax from the base language or the ggplot2 package.
+For this exercise, base R users had to use ggplot2 and vice versa.
+In either approach, the Code Club ends with a report back to the larger group describing the approach each pair took.
 Our Code Clubs typically have 7 to 10 participants, but the inherent "think-pair-share" strategy should allow it to be scaled to larger groups [@Kothiyal2014].
 
-We continue to experiment with approaches to running Code Club.
-During the period of the Covid-19 pandemic in 2020 when many research labs were shuttered, PDS started hosting weekly Code Club virtually.
+We continue to experiment with approaches for running Code Club.
+During the period of the Covid-19 pandemic in 2020 when many research labs were shuttered, PDS started hosting an open virtual weekly Code Club.
 Participants could engage the material posted to https://www.riffomonas.org/code_club/ via a Zoom meeting or asynchronously using the recorded sessions available on YouTube.
 Regardless of our approach to leading a Code Club, we have learned that it is critical for the presenter to clearly articulate their goals and facilitate participant engagement.
-Although any individual Code Club session may be more experimental than another, on the whole they are a critical tool to train bench scientists in reproducible data analysis practices.
-We have provided some examples of successful code club topics in Table 1.
+Although some Code Club sessions may be more experimental than others, on the whole they are a critical tool to train bench scientists in reproducible data analysis practices.
+We have provided some examples of successful topics in Table 1.
 We have summarized the results of our own experimentation as Code Club presenters and participants into Ten Rules.
-The first 3 rules apply to all participants at every Code Club, while Rules 4 through 8 are targeted to Code Club presenters and the last two for participants.
+The first 3 rules apply to all participants, Rules 4 through 8 are targeted to presenters, and the last two for participants.
 
@@
 It is critical that the presenter and participants respect each other and that a designated individual (e.g.
-the lab director) enforces a code of conduct (see X for an example).
+the lab director) enforces a code of conduct.
 Each member of the Code Club must have the humility to acknowledge that they have more to learn about any given topic.
-We have found that many problems are avoided when the presenter takes charge of the session with a clear lesson plan, thoughtfully creating groups, and giving encouragement.
-Similarly, participants foster a positive environment by remembering that the task is not a competition, focusing on the presenter's goals, allowing their partner to contribute, asking clarifying questions when appropriate, and avoiding distractions (e.g.
+We have found that many problems are avoided when the presenter takes charge of the session with a clear lesson plan, thoughtfully creates groups, and gives encouragement.
+Similarly, participants foster a positive environment by remembering that the task is not a competition, instead focusing on the presenter's goals, allowing their partner to contribute, asking clarifying questions when appropriate, and avoiding distractions (e.g.
 email, social media).
-Learning to program is challenging and too often results in a toxic environment of nitpicking and nonconstructive criticism.
-All parties in a Code Club are responsible for preventing this toxicity by demonstrating respect for themselves and their colleagues.
+Learning to program is challenging.
+Too often it is attempted in a environment with nonconstructive criticism.
+All parties in a Code Club are responsible for preventing this by demonstrating respect for themselves and their colleagues.
 
@@
 
-Part of the humility required to participate in a Code Club is acknowledging that your training is incomplete and that it is possible for you to learn something new.
+Part of the humility required to participate in a Code Club is acknowledging that your training is incomplete and that it is possible for everyone to learn something new.
 For participants, assume that the presenter has a plan and follow their presented approach.
@@
 
-The selection of content and structure for each Code Club is democratic and distributed.
+The selection of content and structure for each Code Club works best when it is democratic and distributed.
 If someone thinks a technique is worth learning and wants to teach it, they have that power.
@@
 If members are following these Rules, they will naturally reflect on the skills and interests of the other members in the group.
-For example, there is always turnover in a research group, making it important to revisit basic concepts to teach it to new people and provide a refresher to others.
+For example, there is always turnover in a research group, making it important to revisit basic concepts to teach to new people and provide a refresher to others.
+Keeping an online log of previous topics or topics of interest for future sessions is a useful organizational tool.
 The group should also feel free to experiment with the format and incorporate group feedback by ending with a debrief discussing the pros and cons of new formats.
@@
 A typical schedule for Code Club is 10 minutes of introduction and instruction, 30 minutes of paired programming, 5 minutes to get groups to wrap up, and 10 minutes to report back to the group.
+We once had a presenter try to teach basic, but unfamiliar, Julia syntax.
+Unfortunately, the time was up before the participants had installed the interpreter.
+Some tips to help keep it simple are to limit the presented code chunks to less than 50 lines or, conversely, consider the number of lines that might be required to accomplish a solution.
 Remember that learners may need up to three times as long to complete a task that is simple for the presenter, so Code Club is best kept simple.
-We once had a presenter try to teach basic Julia syntax, but time was up by the time participants installed the interpreter.
-Some tips to help keep it simple are to limit the presented code chunks to less than 50 or, conversely, consider the number of lines that might be required to accomplish a solution.
 
@@
 We have found that using GitHub repositories for each Code Club can help make information, scripts, and data easily available to participants.
-Perhaps a first Code Club could be using git and GitHub to engage in collaborative coding.
-A lower barrier entry point is posting their code, data, and information in a lab meeting-dedicated Slack channel or sharing via email.
+Perhaps a first Code Club could introduce using git and GitHub to engage in collaborative coding.
+A lower barrier entry point for the presenter is posting their code, data, and information in a lab meeting-dedicated Slack channel or via email.
 Whatever method is used, the presenter should be sure to communicate the topic and necessary materials with the participants ahead of time.
@@
 
-Participants want to learn topics that will either be useful to them or help their colleague (i.e., the presenter).
+Participants want to learn topics that will either be useful to them or help their colleague (i.e.
+the presenter).
 Presenters should do their best to satisfy those motivations, whether it is through the relevancy of the concept or the data.
@@
 Similarly, participants should not be tasked with improving the presenter's code if the presenter has no intention of incorporating the suggestions.
-A list of packages or tasks that group members are interested in could help an uninspired presenter choose a topic that will make a rewarding Code Club.
+A list of packages or tasks that group members are interested in could help a presenter struggling to find a topic choose one that will make a rewarding Code Club.
 
@@
 As suggested by Rule 7, a significant challenge to presenting at Code Club is selecting topics and activities that appeal to a critical mass of the participants.
-This is particularly difficult if participants have a wide range of coding experience, such as follows a turnover in group membership.
+This is particularly difficult if participants have a wide range of coding experience, which can follow a turnover in group membership.
 In these circumstances it can be useful to cover commonly used tools but that might result in disinterest of more experienced participants.
@@
 Instead of letting participants form their own pairs, the presenter can select pairs of participants with either similar or differing skill levels, depending on their goals.
-Partnerships between those with similar skill levels requires the presenter to design appropriate activities each skill level.
+Partnerships between those with similar skill levels requires the presenter to design appropriate activities for each skill level.
 We have found that commenting code is a good skill for beginners since it forces them to dissect and understand a code chunk line-by-line.
-An advantage of partnerships with disparate skill levels is assigning a single task, which provides the presenter with a diverse range of methods that achieve the same result.
-This approach to pairing also helps to graft new members into the group that have emerging programming skills.
+It also reinforces the value of commenting as they develop their skills and independence.
+An advantage of forming partnership between people with disparate skill levels is that it is more likely for groups to provide the presenter with a diverse range of methods that achieve the same result.
+This approach to pairing also helps to graft new members that have emerging programming skills.
 Regardless of how partners are selected, consider asking the pairs to identify a navigator and a driver [@Bockeler_2020].
@@
 Midway through the activity, the presenter can have the partners switch roles.
-Intentionally forming pairs can also engineer group interactions by breaking up cliques, avoiding potentially disruptive partnerships, or pairing reliable role models with new group members.
+Intentionally forming pairs can also engineer group interactions by avoiding potentially disruptive partnerships or pairing reliable role models with new group members.
 
@@
 In that context, coming to Code Club without having installed a dependency is similar to asking a simple question about the Journal Club paper.
-Both instances show a lack of respect and preparation.
+Both instances show a lack of preparation.
 Just as a presenter must follow Rule 6 to provide materials ahead of time, participants must review the code in advance, download the data sets, install the necessary packages, and perhaps read up on the topic.
@@
 Fortunately, programming languages like R and Python are generally expressive, which should allow you to engage with the logic, even if the syntax is too advanced.
-Frankly, understanding the logic of *when* to use one modeling approach over another is more important than knowing *how* to use it.
+Oftentimes, understanding the logic of *when* to use one modeling approach over another is more important than knowing *how* to use it.
 If you understand the "why", the "how" will quickly follow.
@@
 Members of the Code Club need to feel comfortable with other group members and sufficiently empowered to try something new or ask for help.
-Aside from growing in our programming skills, we have noticed two other benefits that help create a positive work culture.
+Aside from expanding our programming skills, we have noticed two other benefits that help create a positive work culture.
 First, we have intentionally interviewed postdoc candidates on the days we hold Code Club.
-We make it clear that they are not being assessed on their coding skills, but instead use it as an opportunity to see how the candidate interacts with other members of the research group in a relatively low stakes environment.
+We make it clear that they are not being assessed on their coding skills, but instead use it as an opportunity to see how the candidate interacts with other members of the research group.
 At the same time, the candidate can learn about the culture of the research group through active participation.
-Second, members of other research groups that feel isolated in growing their skills have integrated themselves into our Code Club.
+Second, members of other research groups have integrated themselves into our Code Club to minimize the isolation they feel in growing their skills within smaller research groups.
 This speaks to both the broader need for Code Club and the likelihood of success when expanded to include a larger group of individuals with broader research interests.
@@
 They have had a critical role in shaping the evolution of the sessions.
-AKH led a lab discussion of Code Clubs to identify the key concepts presented, which were supplemented by contributions from NL.
+AKH led a lab discussion of Code Clubs to identify the key concepts presented, which were supplemented by contributions from NAL.
 The order of other co-authors is listed alphabetically.
@@
 | ----------------- | ---------------------------------------------- |
-| base vs. ggplot   | Given input data and a figure, recreate the figure using R's base graphics or ggplot syntax   |
+| base vs. ggplot2   | Given input data and a figure, recreate the figure using R's base graphics or ggplot2 syntax   |
 | Snakemake         | Given a bash script that contains an analysis pipeline, convert it to a Snakemake workflow (can also be done with GNU Make) |
@@
 | R with Google docs   | Scrape a Google docs workbook and clean the data to identify previous Code Club presenters |
-| Develop an R Package  | Convert a lab member's collection of scripts into package over a series of sessions  |
+| Develop an R Package  | Convert a lab member's collection of scripts into an R package over a series of sessions  |

### 2020-04-28 submission/manuscript.tex
First submission

@@
 
-\vspace*{5mm}
-
 \hypertarget{ten-simple-rules-to-increase-computational-skills-among-biologists-with-code-clubs}{%
@@
 
-\vspace{25mm}
+\vspace{5mm}
 
-Ada K. Hagan\({^2}\), Nick Lesniak, \emph{Author names here, order
-alphabetical}, Patrick D. Schloss\({^1}\)\({^\dagger}\)
+Ada K. Hagan\({^{1,2}}\), Nicholas A. Lesniak\({^1}\), Marcy J.
+Balunas\({^3}\), Lucas Bishop\({^1}\), William L. Close\({^1}\), Matthew
+D. Doherty\({^1}\), Amanda G. Elmore\({^1}\), Kaitlin J.
+Flynn\({^{1,4}}\), Geoffrey D. Hannigan\({^{1,5}}\), Charlie C.
+Koumpouras\({^{1,6}}\), Matthew L. Jenior\({^{1,7}}\), Ariangela J.
+Kozik\({^8}\), Kathryn McBride\({^1}\), Samara B. Rifkin\({^8}\), Joshua
+M.A.~Stough\({^1}\), Kelly L. Sovacool\({^9}\), Marc A.
+Sze\({^{1,10}}\), Sarah Tomkovich\({^1}\), Begum D.
+Topcuoglu\({^{1,5}}\), and Patrick D. Schloss\({^1}\)\({^\dagger}\)
 
-\vspace{40mm}
+\vspace{5mm}
 
@@
 
-\(1\) Department of Microbiology and Immunology, University of Michigan,
-Ann Arbor, MI 48109
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Department of Microbiology and Immunology, University of Michigan, Ann
+  Arbor, MI
+\item
+  Current affiliation: Alliance SciComm \& Consulting, Linden, MI
+\item
+  Division of Medicinal Chemistry, Department of Pharmaceutical
+  Sciences, University of Connecticut, Storrs, CT
+\item
+  Current affiliation: Benaroya Research Institute, Seattle, WA
+\item
+  Current affiliation: Exploratory Science Center, Merck \& Co., Inc.,
+  Cambridge, Massachusetts, USA
+\item
+  Current affiliation: Roche Diagnostics, Clinical Operations Services
+  \& eSystems, Indianapolis, IN
+\item
+  Current affiliation: Department of Biomedical Engineering, University
+  of Virginia, Charlottesville, VA
+\item
+  Department of Internal Medicine, University of Michigan
+\item
+  Department of Computational Medicine \& Bioinformatics, University of
+  Michigan
+\item
+  Current affiliation: Genomics and Pharmacogenomics, Merck \& Co.,
+  Inc., Cambridge, MA
+\end{enumerate}
 
-\(2\) Alliance SciComm \& Consulting, Linden, MI 48418
-
-\vspace{35mm}
+\vspace{5mm}
 
@@
 generation and data analysis, we have found scientists feel empowered
-when they can both ask and answer their own biological questions. Yet,
-the standard undergraduate and graduate training in the biological
-sciences is insufficient for developing these advanced data analysis
-skills. In our experience performing microbiome research, it is more
-common to find exceptional bench scientists who are inexperienced at
-analyzing large data sets than to find the reverse. Of course this
-raises a challenge: How do we train bench scientists that analyze data
-sets to answer biological questions?
+when they can both ask and answer their own biological questions
+{[}1{]}. Yet, the standard undergraduate and graduate training in the
+biological sciences is typically insufficient for developing these
+advanced data analysis skills. In our experience performing microbiome
+research, it is more common to find exceptional bench scientists who are
+inexperienced at analyzing large data sets than to find the reverse. Of
+course this raises a challenge: How do we train bench scientists to
+effectively answer biological questions with these larger datasets?
 
-The ever-growing ability to generate data and constant feeling of
-helplessness in analyzing it is analogous to the struggles we also face
-with engaging the voluminous scientific literature. A common strategy to
-keeping up with the literature are journal clubs, which involve group
-discussion of a pre-selected paper. These range from informal
-discussions to PowerPoint presentations and course credit. In addition
-to staying current on the literature, journal clubs help strengthen
-skills in critical thinking, communication, and literature review
-{[}1{]}. Our research group has leveraged the similarity between the
-overwhelming natures of scientific literature and data analysis to
-address the challenge of teaching reproducible data analysis practices
-to bench scientists. Over the past 4 years we have experimented with a
-Code Club model to improve data analysis skills in a community
-environment.
+The ever-growing ability to generate data and feeling of unease in
+analyzing the data are analogous to the struggles scientists also face
+with engaging the voluminous scientific literature. A common strategy
+for keeping up with the literature is participating in journal clubs,
+which involve group discussion of a pre-selected paper. These range from
+informal discussions to PowerPoint presentations and course credit. In
+addition to staying current on the literature, journal clubs help
+strengthen skills in critical thinking, communication, and integrating
+the literature {[}2{]}. Our research group has leveraged the similarity
+between the overwhelming nature of both scientific literature and data
+analysis to address the challenge of teaching reproducible data analysis
+practices to bench scientists. Over the past four years we have
+experimented with a Code Club model to improve data analysis skills in a
+community environment.
 
 Our Code Club sessions are an hour long and alternate with Journal Club
-as the second part of weekly two hour lab meetings. Presenters volunteer
-to lead a session with the expectation that they alternate leading
-Journal and Code Clubs. Initially, the Code Club was used to review code
-from trainee projects. Instead of a presentation, the presenter would
-project their code onto a screen and the participants would go line by
-line through the code, stating the logic behind each line. This approach
-emphasized the importance of code readability and gave beginners the
-opportunity to see the real-life, messy code of more experienced peers.
-Unfortunately, the format only allowed us to review a fraction of a
-project's code base, making it difficult to integrate the programmer's
-logic across their full project. A major issue with this model was that
-sometimes beginners couldn't contribute to improving the code, and even
-when they could, more senior group members would eventually dominate the
-discussion. This led to a lack of participation by beginners, who would
-mentally check out to focus on other activities, and frequently resulted
-in an adversarial environment between the presenter and senior group
-members. As a result, presenters rarely offered to present Code Club
-again.
+as the second part of weekly two hour lab meetings. Initially, the Code
+Club was used to review code from trainee projects. Instead of a
+presentation, the presenter would project their code onto a screen and
+the participants would go through the code, stating the logic behind
+each line. This approach emphasized the importance of code readability
+and gave beginners the opportunity to see the real-life, messy code of
+more experienced peers. Unfortunately, the format only allowed us to
+review a fraction of a project's code, making it difficult to integrate
+the programmer's logic across their full project. A major issue with
+this model was that sometimes beginners could not contribute to
+improving the code, and even when they could, more experienced group
+members would eventually dominate the discussion. This led to a lack of
+participation by beginners, who would sometimes mentally check out and
+resulted in an adversarial environment between the presenter and more
+experienced members. As a result, presenters were reluctant to offer to
+present again.
 
 From these experiences, we began a collective conversation to improve
-our Code Club model and have evolved it into two approaches. The first
-is a more constructive version of a group code review. The presenter
-clearly states the problem they want to solve, breaks the participants
-into smaller groups, and then asks each group to solve the problem, or a
-portion of it. For example, someone may have an R script with a
-repeating chunk of code. The challenge for the session would be to
-convert the code chunk into a function to be called throughout the
-script to make it ``DRY'' (i.e.~Don't Repeat Yourself){[}2{]}. The
+our Code Club model. We have identified two successful approaches. The
+first is a more constructive version of a group code review. The
+presenter clearly states the problem they want to solve, breaks the
+participants into smaller groups, and then asks each group to solve the
+problem, or a portion of it. For example, someone may have an R script
+with a repeating chunk of code. The challenge for the session would be
+to convert the code chunk into a function to be called throughout the
+script to make it ``DRY'' (i.e.~Don't Repeat Yourself {[}3{]}). The
 presenter leaves the session with several partial or working solutions
@@
 package or technique and assigns an activity to practice the new
-approach. For example, a more memorable activity was giving participants
-raw data and a finished plot. Paired participants were tasked with
-generating the plot using either base R or ggplot, but base R users had
-to use ggplot and vice versa. In either approach, the Code Club ends
-with a report back to the presenter and larger group describing the
-approach each pair took. Our Code Clubs typically have 7 to 10
-participants, but the inherent ``think-pair-share'' strategy should
-allow it to be scaled to larger groups {[}3{]}.
+approach. For example, at one Code Club participants were given raw data
+and a finished plot. Paired participants were tasked with generating the
+plot from the data using either R syntax from the base language or the
+ggplot2 package. For this exercise, base R users had to use ggplot2 and
+vice versa. In either approach, the Code Club ends with a report back to
+the larger group describing the approach each pair took. Our Code Clubs
+typically have 7 to 10 participants, but the inherent
+``think-pair-share'' strategy should allow it to be scaled to larger
+groups {[}4{]}.
 
-We continue to experiment with approaches to running Code Club. During
+We continue to experiment with approaches for running Code Club. During
 the period of the Covid-19 pandemic in 2020 when many research labs were
-shuttered, PDS started hosting weekly Code Club virtually. Participants
-could engage the material posted to
+shuttered, PDS started hosting an open virtual weekly Code Club.
+Participants could engage the material posted to
 \url{https://www.riffomonas.org/code_club/} via a Zoom meeting or
@@
 it is critical for the presenter to clearly articulate their goals and
-facilitate participant engagement. Although any individual Code Club
-session may be more experimental than another, on the whole they are a
-critical tool to train bench scientists in reproducible data analysis
-practices. We have provided some examples of successful code club topics
-in Table 1. We have summarized the results of our own experimentation as
-Code Club presenters and participants into Ten Rules. The first 3 rules
-apply to all participants at every Code Club, while Rules 4 through 8
-are targeted to Code Club presenters and the last two for participants.
+facilitate participant engagement. Although some Code Club sessions may
+be more experimental than others, on the whole they are a critical tool
+to train bench scientists in reproducible data analysis practices. We
+have provided some examples of successful topics in Table 1. We have
+summarized the results of our own experimentation as Code Club
+presenters and participants into Ten Rules. The first 3 rules apply to
+all participants, Rules 4 through 8 are targeted to presenters, and the
+last two for participants.
 
@@
 and that a designated individual (e.g.~the lab director) enforces a code
-of conduct (see X for an example). Each member of the Code Club must
-have the humility to acknowledge that they have more to learn about any
-given topic. We have found that many problems are avoided when the
-presenter takes charge of the session with a clear lesson plan,
-thoughtfully creating groups, and giving encouragement. Similarly,
-participants foster a positive environment by remembering that the task
-is not a competition, focusing on the presenter's goals, allowing their
-partner to contribute, asking clarifying questions when appropriate, and
-avoiding distractions (e.g.~email, social media). Learning to program is
-challenging and too often results in a toxic environment of nitpicking
-and nonconstructive criticism. All parties in a Code Club are
-responsible for preventing this toxicity by demonstrating respect for
-themselves and their colleagues.
+of conduct. Each member of the Code Club must have the humility to
+acknowledge that they have more to learn about any given topic. We have
+found that many problems are avoided when the presenter takes charge of
+the session with a clear lesson plan, thoughtfully creates groups, and
+gives encouragement. Similarly, participants foster a positive
+environment by remembering that the task is not a competition, instead
+focusing on the presenter's goals, allowing their partner to contribute,
+asking clarifying questions when appropriate, and avoiding distractions
+(e.g.~email, social media). Learning to program is challenging. Too
+often it is attempted in a environment with nonconstructive criticism.
+All parties in a Code Club are responsible for preventing this by
+demonstrating respect for themselves and their colleagues.
 
@@
 acknowledging that your training is incomplete and that it is possible
-for you to learn something new. For participants, assume that the
+for everyone to learn something new. For participants, assume that the
 presenter has a plan and follow their presented approach. After the Code
@@
 
-The selection of content and structure for each Code Club is democratic
-and distributed. If someone thinks a technique is worth learning and
-wants to teach it, they have that power. If they want to experiment with
-a different format, they are free to try it out. Members of the Code
-Club need to feel like they have the power to shape the direction of the
-group. If members are following these Rules, they will naturally reflect
-on the skills and interests of the other members in the group. For
-example, there is always turnover in a research group, making it
-important to revisit basic concepts to teach it to new people and
-provide a refresher to others. The group should also feel free to
-experiment with the format and incorporate group feedback by ending with
-a debrief discussing the pros and cons of new formats.
+The selection of content and structure for each Code Club works best
+when it is democratic and distributed. If someone thinks a technique is
+worth learning and wants to teach it, they have that power. If they want
+to experiment with a different format, they are free to try it out.
+Members of the Code Club need to feel like they have the power to shape
+the direction of the group. If members are following these Rules, they
+will naturally reflect on the skills and interests of the other members
+in the group. For example, there is always turnover in a research group,
+making it important to revisit basic concepts to teach to new people and
+provide a refresher to others. Keeping an online log of previous topics
+or topics of interest for future sessions is a useful organizational
+tool. The group should also feel free to experiment with the format and
+incorporate group feedback by ending with a debrief discussing the pros
+and cons of new formats.
 
@@
 should create a simplified scenario (i.e.~minimal, reproducible example
-{[}4{]}), which can be helpful in focusing the participants. The
+{[}5{]}), which can be helpful in focusing the participants. The
 presenter should verify ahead of time that the simplified example works
@@
 of paired programming, 5 minutes to get groups to wrap up, and 10
-minutes to report back to the group. Remember that learners may need up
-to three times as long to complete a task that is simple for the
-presenter, so Code Club is best kept simple. We once had a presenter try
-to teach basic Julia syntax, but time was up by the time participants
-installed the interpreter. Some tips to help keep it simple are to limit
-the presented code chunks to less than 50 or, conversely, consider the
-number of lines that might be required to accomplish a solution.
+minutes to report back to the group. We once had a presenter try to
+teach basic, but unfamiliar, Julia syntax. Unfortunately, the time was
+up before the participants had installed the interpreter. Some tips to
+help keep it simple are to limit the presented code chunks to less than
+50 lines or, conversely, consider the number of lines that might be
+required to accomplish a solution. Remember that learners may need up to
+three times as long to complete a task that is simple for the presenter,
+so Code Club is best kept simple.
 
@@
 information, scripts, and data easily available to participants. Perhaps
-a first Code Club could be using git and GitHub to engage in
-collaborative coding. A lower barrier entry point is posting their code,
-data, and information in a lab meeting-dedicated Slack channel or
-sharing via email. Whatever method is used, the presenter should be sure
-to communicate the topic and necessary materials with the participants
-ahead of time.
+a first Code Club could introduce using git and GitHub to engage in
+collaborative coding. A lower barrier entry point for the presenter is
+posting their code, data, and information in a lab meeting-dedicated
+Slack channel or via email. Whatever method is used, the presenter
+should be sure to communicate the topic and necessary materials with the
+participants ahead of time.
 
@@
 Participants want to learn topics that will either be useful to them or
-help their colleague (i.e., the presenter). Presenters should do their
+help their colleague (i.e.~the presenter). Presenters should do their
 best to satisfy those motivations, whether it is through the relevancy
@@
 incorporating the suggestions. A list of packages or tasks that group
-members are interested in could help an uninspired presenter choose a
-topic that will make a rewarding Code Club.
+members are interested in could help a presenter struggling to find a
+topic choose one that will make a rewarding Code Club.
 
@@
 of the participants. This is particularly difficult if participants have
-a wide range of coding experience, such as follows a turnover in group
+a wide range of coding experience, which can follow a turnover in group
 membership. In these circumstances it can be useful to cover commonly
@@
 depending on their goals. Partnerships between those with similar skill
-levels requires the presenter to design appropriate activities each
+levels requires the presenter to design appropriate activities for each
 skill level. We have found that commenting code is a good skill for
 beginners since it forces them to dissect and understand a code chunk
-line-by-line. An advantage of partnerships with disparate skill levels
-is assigning a single task, which provides the presenter with a diverse
-range of methods that achieve the same result. This approach to pairing
-also helps to graft new members into the group that have emerging
-programming skills. Regardless of how partners are selected, consider
-asking the pairs to identify a navigator and a driver {[}5{]}. The
-driver types at the computer while the navigator tells them what to
-type, thus ensuring participation of both partners. Midway through the
-activity, the presenter can have the partners switch roles.
-Intentionally forming pairs can also engineer group interactions by
-breaking up cliques, avoiding potentially disruptive partnerships, or
+line-by-line. It also reinforces the value of commenting as they develop
+their skills and independence. An advantage of forming partnership
+between people with disparate skill levels is that it is more likely for
+groups to provide the presenter with a diverse range of methods that
+achieve the same result. This approach to pairing also helps to graft
+new members that have emerging programming skills. Regardless of how
+partners are selected, consider asking the pairs to identify a navigator
+and a driver {[}6{]}. The driver types at the computer while the
+navigator tells them what to type, thus ensuring participation of both
+partners. Midway through the activity, the presenter can have the
+partners switch roles. Intentionally forming pairs can also engineer
+group interactions by avoiding potentially disruptive partnerships or
 pairing reliable role models with new group members.
@@
 a simple question about the Journal Club paper. Both instances show a
-lack of respect and preparation. Just as a presenter must follow Rule 6
-to provide materials ahead of time, participants must review the code in
-advance, download the data sets, install the necessary packages, and
-perhaps read up on the topic. If the Code Club is based on a paper or
-chapters in a data science book (e.g., {[}6--8{]}), the participants
-should read them before the session and consider how they might
-incorporate the concepts into their own work.
+lack of preparation. Just as a presenter must follow Rule 6 to provide
+materials ahead of time, participants must review the code in advance,
+download the data sets, install the necessary packages, and perhaps read
+up on the topic. If the Code Club is based on a paper or chapters in a
+data science book (e.g., {[}7--9{]}), the participants should read them
+before the session and consider how they might incorporate the concepts
+into their own work.
 
@@
 are generally expressive, which should allow you to engage with the
-logic, even if the syntax is too advanced. Frankly, understanding the
+logic, even if the syntax is too advanced. Oftentimes, understanding the
 logic of \emph{when} to use one modeling approach over another is more
@@
 Club need to feel comfortable with other group members and sufficiently
-empowered to try something new or ask for help. Aside from growing in
-our programming skills, we have noticed two other benefits that help
-create a positive work culture. First, we have intentionally interviewed
+empowered to try something new or ask for help. Aside from expanding our
+programming skills, we have noticed two other benefits that help create
+a positive work culture. First, we have intentionally interviewed
 postdoc candidates on the days we hold Code Club. We make it clear that
@@
 as an opportunity to see how the candidate interacts with other members
-of the research group in a relatively low stakes environment. At the
-same time, the candidate can learn about the culture of the research
-group through active participation. Second, members of other research
-groups that feel isolated in growing their skills have integrated
-themselves into our Code Club. This speaks to both the broader need for
+of the research group. At the same time, the candidate can learn about
+the culture of the research group through active participation. Second,
+members of other research groups have integrated themselves into our
+Code Club to minimize the isolation they feel in growing their skills
+within smaller research groups. This speaks to both the broader need for
 Code Club and the likelihood of success when expanded to include a
@@
 identify the key concepts presented, which were supplemented by
-contributions from NL. The order of other co-authors is listed
+contributions from NAL. The order of other co-authors is listed
 alphabetically.
@@
 \hypertarget{refs}{}
+\leavevmode\hypertarget{ref-Carey2018}{}%
+1. Carey MA, Papin JA. Ten simple rules for biologists learning to
+program. Markel S, editor. PLOS Computational Biology. 2018;14:
+e1005871.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1005871}{10.1371/journal.pcbi.1005871}
+
 \leavevmode\hypertarget{ref-Lonsdale2016}{}%
-1. Lonsdale A, Penington JS, Rice T, Walker M, Dashnow H. Ten simple
+2. Lonsdale A, Penington JS, Rice T, Walker M, Dashnow H. Ten simple
 rules for a bioinformatics journal club. Lewitter F, editor. PLOS
@@
 \leavevmode\hypertarget{ref-Wilson2014}{}%
-2. Wilson G, Aruliah DA, Brown CT, Hong NPC, Davis M, Guy RT, et al.
+3. Wilson G, Aruliah DA, Brown CT, Hong NPC, Davis M, Guy RT, et al.
 Best practices for scientific computing. Eisen JA, editor. PLoS Biology.
@@
 \leavevmode\hypertarget{ref-Kothiyal2014}{}%
-3. Kothiyal A, Murthy S, Iyer S. Think-pair-share in a large CS1 class.
+4. Kothiyal A, Murthy S, Iyer S. Think-pair-share in a large CS1 class.
 Proceedings of the 2014 conference on innovation \& technology in
@@
 \leavevmode\hypertarget{ref-StackOverflow_MREE}{}%
-4. Stack Overflow. How to create a minimal, reproducible example. 2020.
+5. Stack Overflow. How to create a minimal, reproducible example. 2020.
 Available:
@@
 \leavevmode\hypertarget{ref-Bockeler_2020}{}%
-5. Böckeler B, Siessegger N. On pair programming. 2020. Available:
+6. Böckeler B, Siessegger N. On pair programming. 2020. Available:
 \url{https://martinfowler.com/articles/on-pair-programming.html}
@@
 \leavevmode\hypertarget{ref-Wickham_2016}{}%
-6. Wickham H, Grolemund G. O'Reilly Media; 2016. Available:
+7. Wickham H, Grolemund G. O'Reilly Media; 2016. Available:
 \url{https://r4ds.had.co.nz}
@@
 \leavevmode\hypertarget{ref-Schloss_2018}{}%
-7. Schloss PD. The riffomonas reproducible research tutorial series.
+8. Schloss PD. The riffomonas reproducible research tutorial series.
 Journal of Open Source Education. 2018;1: 13.
@@
 \leavevmode\hypertarget{ref-Noble2009}{}%
-8. Noble WS. A quick guide to organizing computational biology projects.
+9. Noble WS. A quick guide to organizing computational biology projects.
 Lewitter F, editor. PLoS Computational Biology. 2009;5: e1000424.
@@
 \begin{minipage}[t]{0.25\columnwidth}\raggedright
-base vs.~ggplot\strut
+base vs.~ggplot2\strut
 \end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
 Given input data and a figure, recreate the figure using R's base
-graphics or ggplot syntax\strut
+graphics or ggplot2 syntax\strut
 \end{minipage}\tabularnewline
@@
 \end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
-Convert a lab member's collection of scripts into package over a series
-of sessions\strut
+Convert a lab member's collection of scripts into an R package over a
+series of sessions\strut
 \end{minipage}\tabularnewline

### 2020-06-18 submission/response_to_reviewers.md
Pat's first take at revision

**Reviewer #1**

**The authors put together an interesting 10 simple rules article sharing their experiences with code clubs.**

**I find the proposed code clubs an interesting idea and would be interested in trying this out in the lab. However, quite a bit more practical advice from this manuscript would be helpful to get up and running. Practical advice can be in the form of things that worked, but also explicitly mentioning what didn't work so that others can learn from the authors experience.**

> Thank you for the feedback on where the manuscript could provide more context. Throughout their review, the reviewer encourages us to make more comments about groups where there are more people with CS backgrounds. Admittedly, our lab is made of researchers with non-CS backgrounds and so our implementation of Code Clubs is with that audience in mind. We have added text as appropriate to respond to the following suggestions.

**Some suggestions:**
**Put this into context of the literature: is there any formal description of code clubs? What other learning approaches are commonly used to introduce and improve coding skills in people with no formal programming training?**

**How does this connect to other activities such as hackathons, pair-programming (buddy programming)?**

> We have provided a broader context in the second paragraph of the Introduction and with an expanded discussion of pair-programming in Rule 8.

**Highlight more the benefits of both junior and senior coders, for now the benefits appears to be solely for the junior coders. What motivates the senior folks to join? Or alternatively, in the context of a more diverse lab, how would this work with people with CS and non-CS backgrounds?**

> We have added text to Rule 3 describing the development of more challenging problems for more experienced group members. There is also now a longer section explicitly stating the benefits to more experienced group members and ways to keep their attention by developing sessions around packages and concepts that are new and accessible to everyone.

**Some suggestions on how to implement this would be nice, i.e. how to get started, how to get people motivated to join, etc.**

> It is challenging to comment on this point since we have only done this a part of our normal lab meetings where lab members are compelled to participate. We have added some text to the Conclusion paragraph describing why this model should be successful for more independent groups.

**Are these code clubs unique to biologists? How would this set up work in a hybrid cs/bio environment or perhaps even in a non-bio environment?**

> There's no reason to think that Code Clubs would be unique to biologists. We have added text to the end of the Introduction indicating that as long as the Rules are followed it should be generalizable to groups that are more proficient at programming. Just as Journal Clubs are valuable to all fields of science and levels of experience, we think Code Clubs should also be valuable to the same diversity of participants.

**I would be interested to read more about how much effort it takes in reality from both participants and presenters to prepare? How many hours each week need to be dedicated?**

> In the second to last paragraph of the Introduction we added a sentence indicating that we have found that it takes about as much effort to put together material for a Code Club as it does for a Journal Club session.

**It would be interesting to see a much more extensive proposed program. The examples in Table 1 are illustrative, but far from complete. Can the list be expanded with repeatable activities (git was mentioned in the text) that could constitute a rotating 1 year program and a list of example side activities that are more one-off.**

> Our goal in this manuscript is not to be proscriptive about what the session should contain. We are presenting a framework that diverse groups can use to achieve their own goals with their interests and level of background. We suspect one could certainly create a "schedule" of topics, but the challenge would be sustaining interest across years with the same cohort of participants.

**It would be nice if the above suggestions can be embedded within the rules. To create the space, I think rule 7 and 8 can be merged. Rule 7 seems rather obvious. Rule 6 and 9 are connected and would make sense to have them subsequently, or alternatively as a single rule with two sides of the coin.**

> In our experience, Rule 7 was not obvious and led to the deeper discussion that caused us to re-form our Code Club format. Rather than moving Rules around, we've incorporated the reviewer's comments throughout the manuscript.

**Reviewer #2**
**The manuscript from Hagan, Schloss et al. presents a guide for developing computational skills in labs/groups/buildings/departments, and especially in those environments that include wet lab scientists. This is a timely piece given both the increased importance of computational analyses over the past 1-2 decades, as well as the more recent shuttering of many wet labs due to COVID-19. The piece describes steps to take to develop 1-hour "Code Club" sessions. At these sessions, presenters pose a problem and pair students to solve the problem. After the pair programming, there is a opportunity to reconvene and present solutions.**

**Two general scenarios are explored -- one in which the presenters are helping to solve a problem for the presenter, and another in which the problem is didactic and the key beneficiaries are the other participants. Specific examples for sessions are provided, and additional guidance is offered throughout the rules. Overall, the rules are well organized and the direction is clear.**

**It would appear that the authors violated the first rule of "Code Club" in writing about it, and that is to the benefit of all of us. I have only minor comments:**

> We thank the reviewer for their comments on the manuscript and suggestions for where to add more detail and where we can clarify the text.

**L81-91 (Rule 3): Can you provide more detail in this section? What are examples here of how to "experiment", or what should be avoided? Additionally, returning to basics for new participants seems to conflict with the message to try something new.**

> We have added examples of how we've been experimenting with format during the COVID-19 pandemic to the Introduction as well as variation in difficulty of problems that are assigned in Rule 3.

**L125 (Rule 6): Are the authors recommending a separate GitHub repository for each club meeting, or one general one for the club? Are any of these materials public that they could be shared here as an example? I reviewed the materials at "https://www.riffomonas.org/code_club" but I assume these are more elaborate than what is used in-house.**

> We have added text to Rule 6 detailing what we have found to be pros and cons of hosting material on GitHub. Although interacting with git and GitHub can seem trivial to more experienced people, it can be intimidating to beginners. In many cases, using a dedicated Slack channel or email has lower barriers and allows us to focus on the topic for the session rather than worrying about git configurations or GitHub permissions. Although not all of the examples in Table 1 have repositories on our group's GitHub account, we have included links to those that are available. Looking at these, one will see the variation in "polish" across the sessions, but will also get a sense of the overall approach we have taken to Code Club sessions.

**L32: Suggest rewording "and resulted" to ", resulting".**

> We have edited the text as suggested.

**L60: Suggest rewording "last two for participants." to "last two for non-presenters."**

> We have edited the text as suggested.

### 2020-06-18 submission/manuscript.Rmd
Pat's first take at revision

@@
 Although some accept the approach of compartmentalizing data generation and data analysis, we have found scientists feel empowered when they can both ask and answer their own biological questions [@Carey2018].
-Yet, the standard undergraduate and graduate training in the biological sciences is typically insufficient for developing these advanced data analysis skills.
 In our experience performing microbiome research, it is more common to find exceptional bench scientists who are inexperienced at analyzing large data sets than to find the reverse.
@@
 
-The ever-growing ability to generate data and feeling of unease in analyzing the data are analogous to the struggles scientists also face with engaging the voluminous scientific literature.
-A common strategy for keeping up with the literature is participating in journal clubs, which involve group discussion of a pre-selected paper.
+The standard undergraduate and graduate training in the biological sciences is typically insufficient for developing these skills [@Barone2017].
+To fill this void, there has been growth in online data science training resources through companies including Data Camp (https://www.datacamp.com) and Codecademy (https://www.codecademy.com), massive open online courses (MOOCs), and in person workshops hosted by organizations including The Carpentries (https://carpentries.org).
+Each of these formats are highly popular among their audience (e.g.
+[@Wilson2016]); however, recent analyses of workshops suggests that they have little impact on the long term retention of the material that is taught [@Feldon2017].
+This result should not be surprising.
+These learning formats ask participants to engage in massed practice; an approach that is known to be ineffective [@Carpenter2012; @Rohrer2015b; @Bud2010; @Karpicke2011].
+In contrast, participants need regular opportunities to engage in repeated deliberate practice that encourage them to practice their retrieval and application skills of the material [@Rohrer2015a; @Karpicke2008].
+There is clearly a need to develop a framework that can build upon material introduced in coursework and workshops and that can help researchers to stay up to date on the latest best practices, algorithms, and tools.
+
+To address this need, we noted the similarity between the feeling of unease in analyzing the data and the struggles scientists face with engaging the voluminous scientific literature.
+A common strategy for keeping up with the literature is by participating in a journal club.
+Regardless of the discipline, a Journal Club involves group discussion of a pre-selected paper.
 These range from informal discussions to PowerPoint presentations and course credit.
-In addition to staying current on the literature, journal clubs help strengthen skills in critical thinking, communication, and integrating the literature [@Lonsdale2016].
-Our research group has leveraged the similarity between the overwhelming nature of both scientific literature and data analysis to address the challenge of teaching reproducible data analysis practices to bench scientists.
-Over the past four years we have experimented with a Code Club model to improve data analysis skills in a community environment.
+In addition to building upon material from traditional coursework and staying current on the literature, journal clubs help strengthen skills in critical thinking, communication, and integrating the literature [@Lonsdale2016].
+Since most journal clubs occur on a regular schedule, they are effective because they incorporate repeated deliberate practices.
+With this model in mind, over the past four years we have experimented with creating a Code Club model with the goal of improving reproducible data analysis skills in a community environment.
 
-Our Code Club sessions are an hour long and alternate with Journal Club as the second part of weekly two hour lab meetings.
+Our Code Club sessions are an hour long and alternate with our lab's Journal Club as the second part of weekly two hour lab meetings.
 Initially, the Code Club was used to review code from trainee projects.
@@
 A major issue with this model was that sometimes beginners could not contribute to improving the code, and even when they could, more experienced group members would eventually dominate the discussion.
-This led to a lack of participation by beginners, who would sometimes mentally check out and resulted in an adversarial environment between the presenter and more experienced members.
+This led to a lack of participation by beginners, who would sometimes mentally withdraw, resulting in an adversarial environment between the presenter and more experienced members.
 As a result, presenters were reluctant to offer to present again.
@@
 In either approach, the Code Club ends with a report back to the larger group describing the approach each pair took.
-Our Code Clubs typically have 7 to 10 participants, but the inherent "think-pair-share" strategy should allow it to be scaled to larger groups [@Kothiyal2014].
+We generally find that preparing a Code Club session takes as much effort as preparing to present at a journal club.
+Our Code Clubs typically have 7 to 10 participants, but the inherent "think-pair-share" approach should allow it to be scaled to larger groups [@Kothiyal2014].
 
 We continue to experiment with approaches for running Code Club.
-During the period of the Covid-19 pandemic in 2020 when many research labs were shuttered, PDS started hosting an open virtual weekly Code Club.
-Participants could engage the material posted to https://www.riffomonas.org/code_club/ via a Zoom meeting or asynchronously using the recorded sessions available on YouTube.
+For example, during the period of the Covid-19 pandemic in 2020 when many research labs were shuttered, PDS started posting Code Club materials (https://www.riffomonas.org/code_club/) that individuals or research groups can use to practice their coding skills.
+During this time, we continued our group's Code Club sessions by making use of the breakout room and screen sharing features in Zoom.
 Regardless of our approach to leading a Code Club, we have learned that it is critical for the presenter to clearly articulate their goals and facilitate participant engagement.
@@
 We have summarized the results of our own experimentation as Code Club presenters and participants into Ten Rules.
-The first 3 rules apply to all participants, Rules 4 through 8 are targeted to presenters, and the last two for participants.
+The first 3 rules apply to all participants, Rules 4 through 8 are targeted to presenters, and the last two for non-presenters.
+As the rules describe, this model can easily be adapted to groups of non-biologists and to groups with stronger background in computer science.
 
@@
 For example, there is always turnover in a research group, making it important to revisit basic concepts to teach to new people and provide a refresher to others.
-Keeping an online log of previous topics or topics of interest for future sessions is a useful organizational tool.
+One successful experiment found that we could keep basic content interesting to more experienced group members by creating related problems that varied in their difficulty.
 The group should also feel free to experiment with the format and incorporate group feedback by ending with a debrief discussing the pros and cons of new formats.
@@
 We have found that using GitHub repositories for each Code Club can help make information, scripts, and data easily available to participants.
+Although this can be convenient, introducing GitHub on top of the session's activities can impose a significant cognitive load and frustration to new group members as well as to the presenter if they are not comfortable with GitHub.
 Perhaps a first Code Club could introduce using git and GitHub to engage in collaborative coding.
@@
 Similarly, participants should not be tasked with improving the presenter's code if the presenter has no intention of incorporating the suggestions.
-A list of packages or tasks that group members are interested in could help a presenter struggling to find a topic choose one that will make a rewarding Code Club.
+A list of packages or tasks that group members are interested in along with a log of previous topics could help a presenter struggling to find a topic choose one that will make a rewarding Code Club.
 
@@
 This is particularly difficult if participants have a wide range of coding experience, which can follow a turnover in group membership.
-In these circumstances it can be useful to cover commonly used tools but that might result in disinterest of more experienced participants.
-We have identified several strategies to overcome the challenges presented by participants at varying skill levels.
-Instead of letting participants form their own pairs, the presenter can select pairs of participants with either similar or differing skill levels, depending on their goals.
+Beginners will benefit from sessions that cover fundamental concepts and functions.
+The benefits experienced by more experienced participants include the improved understanding of concepts that comes through teaching the concepts and breaking down problems into simpler elements.
+Furthermore, instead of focusing on core functions, the group could balance Code Club sessions that cover basics with those that cover methods and packages that are new to everyone in the group.
+We have also identified several strategies to overcome the challenges presented by participants at varying skill levels.
+Central to the Code Club format is the use of paired-programming [@Hanks2011].
+Instead of letting participants form their own pairs, the presenter can select pairs of participants with either similar or differing skill levels, depending on the presenter's goals.
 Partnerships between those with similar skill levels requires the presenter to design appropriate activities for each skill level.
@@
 It is not possible to fully participate in a Journal Club discussion about a paper that the participant has not read.
-In that context, coming to Code Club without having installed a dependency is similar to asking a simple question about the Journal Club paper.
+In that context, coming to Code Club without having installed a dependency is similar to asking a simple question about the Journal Club paper without first reading it.
 Both instances show a lack of preparation.
@@
 Members of the Code Club need to feel comfortable with other group members and sufficiently empowered to try something new or ask for help.
-Aside from expanding our programming skills, we have noticed two other benefits that help create a positive work culture.
+Aside from expanding our programming skills, we have noticed two other benefits that help create a positive culture.
 First, we have intentionally interviewed postdoc candidates on the days we hold Code Club.
@@
 This speaks to both the broader need for Code Club and the likelihood of success when expanded to include a larger group of individuals with broader research interests.
+There is no reason that the Code Club format would not work with a group of people independent of a research group as long as the follow the Rules.
+Finding datasets and applications that are interesting to a critical mass of people is likely critical to starting and sustaining such a group.
 Ultimately, Code Club has improved the overall data analysis skills, community, and research success of our lab by empowering researchers to seek help from their colleagues.
@@
 | base vs. ggplot2   | Given input data and a figure, recreate the figure using R's base graphics or ggplot2 syntax   |
-| Snakemake         | Given a bash script that contains an analysis pipeline, convert it to a Snakemake workflow (can also be done with GNU Make) |
+| Snakemake         | Given a bash script that contains an analysis pipeline, convert it to a Snakemake workflow (https://github.com/SchlossLab/snakemake_riffomonas_tutorial) |
 | DRYing code       | Given script with repeated code, create functions to remove repetition   |
-| mothur and Vegan	| Given a pairwise community dissimilarity matrix, compare communities using the `adonis` function in the Vegan R package  |
+| Vegan R package	| Compare microbial communities using the `adonis` function in the Vegan R package (https://github.com/SchlossLab/Code_Review_42717)  |
 | tidy data         | Given a wide-formatted data table, convert it to a long, tidy-formatted data table using tools from R's tidyverse  |
 | GitFlow           | Participants file and claim an issue to add their name to a README file in a GitHub-hosted repository and file a pull request to complete the issue |
-| R with Google docs   | Scrape a Google docs workbook and clean the data to identify previous Code Club presenters |
+| googledocs4 R package | Scrape a Google docs workbook and clean the data to identify previous Code Club presenters |
 | Develop an R Package  | Convert a lab member's collection of scripts into an R package over a series of sessions  |
+| Documenting R code  | Use roxygen2 to supplement comments in R code to improve documentation (https://github.com/SchlossLab/documenting-R) |
+| gganimate R package  |  Convert static plots generated with ggplot2 package into GIFs (https://github.com/SchlossLab/2020-04-12-CodeClub_PlotAnimation/) |

### 2020-06-18 submission/manuscript.tex
Pat's first take at revision

@@
 when they can both ask and answer their own biological questions
-{[}1{]}. Yet, the standard undergraduate and graduate training in the
-biological sciences is typically insufficient for developing these
-advanced data analysis skills. In our experience performing microbiome
-research, it is more common to find exceptional bench scientists who are
-inexperienced at analyzing large data sets than to find the reverse. Of
-course this raises a challenge: How do we train bench scientists to
-effectively answer biological questions with these larger datasets?
+{[}1{]}. In our experience performing microbiome research, it is more
+common to find exceptional bench scientists who are inexperienced at
+analyzing large data sets than to find the reverse. Of course this
+raises a challenge: How do we train bench scientists to effectively
+answer biological questions with these larger datasets?
 
-The ever-growing ability to generate data and feeling of unease in
-analyzing the data are analogous to the struggles scientists also face
-with engaging the voluminous scientific literature. A common strategy
-for keeping up with the literature is participating in journal clubs,
-which involve group discussion of a pre-selected paper. These range from
-informal discussions to PowerPoint presentations and course credit. In
-addition to staying current on the literature, journal clubs help
-strengthen skills in critical thinking, communication, and integrating
-the literature {[}2{]}. Our research group has leveraged the similarity
-between the overwhelming nature of both scientific literature and data
-analysis to address the challenge of teaching reproducible data analysis
-practices to bench scientists. Over the past four years we have
-experimented with a Code Club model to improve data analysis skills in a
+The standard undergraduate and graduate training in the biological
+sciences is typically insufficient for developing these skills {[}2{]}.
+To fill this void, there has been growth in online data science training
+resources through companies including Data Camp
+(\url{https://www.datacamp.com}) and Codecademy
+(\url{https://www.codecademy.com}), massive open online courses (MOOCs),
+and in person workshops hosted by organizations including The
+Carpentries (\url{https://carpentries.org}). Each of these formats are
+highly popular among their audience (e.g.~{[}3{]}); however, recent
+analyses of workshops suggests that they have little impact on the long
+term retention of the material that is taught {[}4{]}. This result
+should not be surprising. These learning formats ask participants to
+engage in massed practice; an approach that is known to be ineffective
+{[}5--8{]}. In contrast, participants need regular opportunities to
+engage in repeated deliberate practice that encourage them to practice
+their retrieval and application skills of the material {[}9,10{]}. There
+is clearly a need to develop a framework that can build upon material
+introduced in coursework and workshops and that can help researchers to
+stay up to date on the latest best practices, algorithms, and tools.
+
+To address this need, we noted the similarity between the feeling of
+unease in analyzing the data and the struggles scientists face with
+engaging the voluminous scientific literature. A common strategy for
+keeping up with the literature is by participating in a journal club.
+Regardless of the discipline, a Journal Club involves group discussion
+of a pre-selected paper. These range from informal discussions to
+PowerPoint presentations and course credit. In addition to building upon
+material from traditional coursework and staying current on the
+literature, journal clubs help strengthen skills in critical thinking,
+communication, and integrating the literature {[}11{]}. Since most
+journal clubs occur on a regular schedule, they are effective because
+they incorporate repeated deliberate practices. With this model in mind,
+over the past four years we have experimented with creating a Code Club
+model with the goal of improving reproducible data analysis skills in a
 community environment.
 
-Our Code Club sessions are an hour long and alternate with Journal Club
-as the second part of weekly two hour lab meetings. Initially, the Code
-Club was used to review code from trainee projects. Instead of a
-presentation, the presenter would project their code onto a screen and
-the participants would go through the code, stating the logic behind
-each line. This approach emphasized the importance of code readability
-and gave beginners the opportunity to see the real-life, messy code of
-more experienced peers. Unfortunately, the format only allowed us to
-review a fraction of a project's code, making it difficult to integrate
-the programmer's logic across their full project. A major issue with
-this model was that sometimes beginners could not contribute to
-improving the code, and even when they could, more experienced group
+Our Code Club sessions are an hour long and alternate with our lab's
+Journal Club as the second part of weekly two hour lab meetings.
+Initially, the Code Club was used to review code from trainee projects.
+Instead of a presentation, the presenter would project their code onto a
+screen and the participants would go through the code, stating the logic
+behind each line. This approach emphasized the importance of code
+readability and gave beginners the opportunity to see the real-life,
+messy code of more experienced peers. Unfortunately, the format only
+allowed us to review a fraction of a project's code, making it difficult
+to integrate the programmer's logic across their full project. A major
+issue with this model was that sometimes beginners could not contribute
+to improving the code, and even when they could, more experienced group
 members would eventually dominate the discussion. This led to a lack of
-participation by beginners, who would sometimes mentally check out and
-resulted in an adversarial environment between the presenter and more
+participation by beginners, who would sometimes mentally withdraw,
+resulting in an adversarial environment between the presenter and more
 experienced members. As a result, presenters were reluctant to offer to
@@
 to convert the code chunk into a function to be called throughout the
-script to make it ``DRY'' (i.e.~Don't Repeat Yourself {[}3{]}). The
+script to make it ``DRY'' (i.e.~Don't Repeat Yourself {[}12{]}). The
 presenter leaves the session with several partial or working solutions
@@
 vice versa. In either approach, the Code Club ends with a report back to
-the larger group describing the approach each pair took. Our Code Clubs
-typically have 7 to 10 participants, but the inherent
-``think-pair-share'' strategy should allow it to be scaled to larger
-groups {[}4{]}.
+the larger group describing the approach each pair took. We generally
+find that preparing a Code Club session takes as much effort as
+preparing to present at a journal club. Our Code Clubs typically have 7
+to 10 participants, but the inherent ``think-pair-share'' approach
+should allow it to be scaled to larger groups {[}13{]}.
 
-We continue to experiment with approaches for running Code Club. During
-the period of the Covid-19 pandemic in 2020 when many research labs were
-shuttered, PDS started hosting an open virtual weekly Code Club.
-Participants could engage the material posted to
-\url{https://www.riffomonas.org/code_club/} via a Zoom meeting or
-asynchronously using the recorded sessions available on YouTube.
-Regardless of our approach to leading a Code Club, we have learned that
-it is critical for the presenter to clearly articulate their goals and
-facilitate participant engagement. Although some Code Club sessions may
-be more experimental than others, on the whole they are a critical tool
-to train bench scientists in reproducible data analysis practices. We
-have provided some examples of successful topics in Table 1. We have
+We continue to experiment with approaches for running Code Club. For
+example, during the period of the Covid-19 pandemic in 2020 when many
+research labs were shuttered, PDS started posting Code Club materials
+(\url{https://www.riffomonas.org/code_club/}) that individuals or
+research groups can use to practice their coding skills. During this
+time, we continued our group's Code Club sessions by making use of the
+breakout room and screen sharing features in Zoom. Regardless of our
+approach to leading a Code Club, we have learned that it is critical for
+the presenter to clearly articulate their goals and facilitate
+participant engagement. Although some Code Club sessions may be more
+experimental than others, on the whole they are a critical tool to train
+bench scientists in reproducible data analysis practices. We have
+provided some examples of successful topics in Table 1. We have
 summarized the results of our own experimentation as Code Club
@@
 all participants, Rules 4 through 8 are targeted to presenters, and the
-last two for participants.
+last two for non-presenters. As the rules describe, this model can
+easily be adapted to groups of non-biologists and to groups with
+stronger background in computer science.
 
@@
 making it important to revisit basic concepts to teach to new people and
-provide a refresher to others. Keeping an online log of previous topics
-or topics of interest for future sessions is a useful organizational
-tool. The group should also feel free to experiment with the format and
-incorporate group feedback by ending with a debrief discussing the pros
-and cons of new formats.
+provide a refresher to others. One successful experiment found that we
+could keep basic content interesting to more experienced group members
+by creating related problems that varied in their difficulty. The group
+should also feel free to experiment with the format and incorporate
+group feedback by ending with a debrief discussing the pros and cons of
+new formats.
 
@@
 should create a simplified scenario (i.e.~minimal, reproducible example
-{[}5{]}), which can be helpful in focusing the participants. The
+{[}14{]}), which can be helpful in focusing the participants. The
 presenter should verify ahead of time that the simplified example works
@@
 that using GitHub repositories for each Code Club can help make
-information, scripts, and data easily available to participants. Perhaps
-a first Code Club could introduce using git and GitHub to engage in
-collaborative coding. A lower barrier entry point for the presenter is
-posting their code, data, and information in a lab meeting-dedicated
-Slack channel or via email. Whatever method is used, the presenter
-should be sure to communicate the topic and necessary materials with the
-participants ahead of time.
+information, scripts, and data easily available to participants.
+Although this can be convenient, introducing GitHub on top of the
+session's activities can impose a significant cognitive load and
+frustration to new group members as well as to the presenter if they are
+not comfortable with GitHub. Perhaps a first Code Club could introduce
+using git and GitHub to engage in collaborative coding. A lower barrier
+entry point for the presenter is posting their code, data, and
+information in a lab meeting-dedicated Slack channel or via email.
+Whatever method is used, the presenter should be sure to communicate the
+topic and necessary materials with the participants ahead of time.
 
@@
 incorporating the suggestions. A list of packages or tasks that group
-members are interested in could help a presenter struggling to find a
-topic choose one that will make a rewarding Code Club.
+members are interested in along with a log of previous topics could help
+a presenter struggling to find a topic choose one that will make a
+rewarding Code Club.
 
@@
 a wide range of coding experience, which can follow a turnover in group
-membership. In these circumstances it can be useful to cover commonly
-used tools but that might result in disinterest of more experienced
-participants. We have identified several strategies to overcome the
-challenges presented by participants at varying skill levels. Instead of
-letting participants form their own pairs, the presenter can select
-pairs of participants with either similar or differing skill levels,
-depending on their goals. Partnerships between those with similar skill
-levels requires the presenter to design appropriate activities for each
-skill level. We have found that commenting code is a good skill for
-beginners since it forces them to dissect and understand a code chunk
+membership. Beginners will benefit from sessions that cover fundamental
+concepts and functions. The benefits experienced by more experienced
+participants include the improved understanding of concepts that comes
+through teaching the concepts and breaking down problems into simpler
+elements. Furthermore, instead of focusing on core functions, the group
+could balance Code Club sessions that cover basics with those that cover
+methods and packages that are new to everyone in the group. We have also
+identified several strategies to overcome the challenges presented by
+participants at varying skill levels. Central to the Code Club format is
+the use of paired-programming {[}15{]}. Instead of letting participants
+form their own pairs, the presenter can select pairs of participants
+with either similar or differing skill levels, depending on the
+presenter's goals. Partnerships between those with similar skill levels
+requires the presenter to design appropriate activities for each skill
+level. We have found that commenting code is a good skill for beginners
+since it forces them to dissect and understand a code chunk
 line-by-line. It also reinforces the value of commenting as they develop
@@
 partners are selected, consider asking the pairs to identify a navigator
-and a driver {[}6{]}. The driver types at the computer while the
+and a driver {[}16{]}. The driver types at the computer while the
 navigator tells them what to type, thus ensuring participation of both
@@
 to Code Club without having installed a dependency is similar to asking
-a simple question about the Journal Club paper. Both instances show a
-lack of preparation. Just as a presenter must follow Rule 6 to provide
-materials ahead of time, participants must review the code in advance,
-download the data sets, install the necessary packages, and perhaps read
-up on the topic. If the Code Club is based on a paper or chapters in a
-data science book (e.g., {[}7--9{]}), the participants should read them
-before the session and consider how they might incorporate the concepts
-into their own work.
+a simple question about the Journal Club paper without first reading it.
+Both instances show a lack of preparation. Just as a presenter must
+follow Rule 6 to provide materials ahead of time, participants must
+review the code in advance, download the data sets, install the
+necessary packages, and perhaps read up on the topic. If the Code Club
+is based on a paper or chapters in a data science book (e.g.,
+{[}17--19{]}), the participants should read them before the session and
+consider how they might incorporate the concepts into their own work.
 
@@
 programming skills, we have noticed two other benefits that help create
-a positive work culture. First, we have intentionally interviewed
-postdoc candidates on the days we hold Code Club. We make it clear that
-they are not being assessed on their coding skills, but instead use it
-as an opportunity to see how the candidate interacts with other members
-of the research group. At the same time, the candidate can learn about
-the culture of the research group through active participation. Second,
+a positive culture. First, we have intentionally interviewed postdoc
+candidates on the days we hold Code Club. We make it clear that they are
+not being assessed on their coding skills, but instead use it as an
+opportunity to see how the candidate interacts with other members of the
+research group. At the same time, the candidate can learn about the
+culture of the research group through active participation. Second,
 members of other research groups have integrated themselves into our
@@
 Code Club and the likelihood of success when expanded to include a
-larger group of individuals with broader research interests. Ultimately,
-Code Club has improved the overall data analysis skills, community, and
-research success of our lab by empowering researchers to seek help from
-their colleagues.
+larger group of individuals with broader research interests. There is no
+reason that the Code Club format would not work with a group of people
+independent of a research group as long as the follow the Rules. Finding
+datasets and applications that are interesting to a critical mass of
+people is likely critical to starting and sustaining such a group.
+Ultimately, Code Club has improved the overall data analysis skills,
+community, and research success of our lab by empowering researchers to
+seek help from their colleagues.
 
@@
 
+\leavevmode\hypertarget{ref-Barone2017}{}%
+2. Barone L, Williams J, Micklos D. Unmet needs for analyzing biological
+big data: A survey of 704 NSF principal investigators. Ouellette F,
+editor. PLOS Computational Biology. 2017;13: e1005755.
+doi:\href{https://doi.org/10.1371/journal.pcbi.1005755}{10.1371/journal.pcbi.1005755}
+
+\leavevmode\hypertarget{ref-Wilson2016}{}%
+3. Wilson G. Software carpentry: Lessons learned. F1000Research. 2016.
+doi:\href{https://doi.org/10.12688/f1000research.3-62.v2}{10.12688/f1000research.3-62.v2}
+
+\leavevmode\hypertarget{ref-Feldon2017}{}%
+4. Feldon DF, Jeong S, Peugh J, Roksa J, Maahs-Fladung C, Shenoy A, et
+al. Null effects of boot camps and short-format training for PhD
+students in life sciences. Proceedings of the National Academy of
+Sciences. 2017;114: 9854--9858.
+doi:\href{https://doi.org/10.1073/pnas.1705783114}{10.1073/pnas.1705783114}
+
+\leavevmode\hypertarget{ref-Carpenter2012}{}%
+5. Carpenter SK, Cepeda NJ, Rohrer D, Kang SHK, Pashler H. Using spacing
+to enhance diverse forms of learning: Review of recent research and
+implications for instruction. Educational Psychology Review. 2012;24:
+369--378.
+doi:\href{https://doi.org/10.1007/s10648-012-9205-z}{10.1007/s10648-012-9205-z}
+
+\leavevmode\hypertarget{ref-Rohrer2015b}{}%
+6. Rohrer D. Student instruction should be distributed over long time
+periods. Educational Psychology Review. 2015;27: 635--643.
+doi:\href{https://doi.org/10.1007/s10648-015-9332-4}{10.1007/s10648-015-9332-4}
+
+\leavevmode\hypertarget{ref-Bud2010}{}%
+7. Budé L, Imbos T, Wiel MW van de, Berger MP. The effect of distributed
+practice on students' conceptual understanding of statistics. Higher
+Education. 2010;62: 69--79.
+doi:\href{https://doi.org/10.1007/s10734-010-9366-y}{10.1007/s10734-010-9366-y}
+
+\leavevmode\hypertarget{ref-Karpicke2011}{}%
+8. Karpicke JD, Blunt JR. Retrieval practice produces more learning than
+elaborative studying with concept mapping. Science. 2011;331: 772--775.
+doi:\href{https://doi.org/10.1126/science.1199327}{10.1126/science.1199327}
+
+\leavevmode\hypertarget{ref-Rohrer2015a}{}%
+9. Rohrer D, Dedrick RF, Stershic S. Interleaved practice improves
+mathematics learning. Journal of Educational Psychology. 2015;107:
+900--908.
+doi:\href{https://doi.org/10.1037/edu0000001}{10.1037/edu0000001}
+
+\leavevmode\hypertarget{ref-Karpicke2008}{}%
+10. Karpicke JD, Roediger HL. The critical importance of retrieval for
+learning. Science. 2008;319: 966--968.
+doi:\href{https://doi.org/10.1126/science.1152408}{10.1126/science.1152408}
+
 \leavevmode\hypertarget{ref-Lonsdale2016}{}%
-2. Lonsdale A, Penington JS, Rice T, Walker M, Dashnow H. Ten simple
+11. Lonsdale A, Penington JS, Rice T, Walker M, Dashnow H. Ten simple
 rules for a bioinformatics journal club. Lewitter F, editor. PLOS
@@
 \leavevmode\hypertarget{ref-Wilson2014}{}%
-3. Wilson G, Aruliah DA, Brown CT, Hong NPC, Davis M, Guy RT, et al.
+12. Wilson G, Aruliah DA, Brown CT, Hong NPC, Davis M, Guy RT, et al.
 Best practices for scientific computing. Eisen JA, editor. PLoS Biology.
@@
 \leavevmode\hypertarget{ref-Kothiyal2014}{}%
-4. Kothiyal A, Murthy S, Iyer S. Think-pair-share in a large CS1 class.
+13. Kothiyal A, Murthy S, Iyer S. Think-pair-share in a large CS1 class.
 Proceedings of the 2014 conference on innovation \& technology in
@@
 \leavevmode\hypertarget{ref-StackOverflow_MREE}{}%
-5. Stack Overflow. How to create a minimal, reproducible example. 2020.
+14. Stack Overflow. How to create a minimal, reproducible example. 2020.
 Available:
@@
 
+\leavevmode\hypertarget{ref-Hanks2011}{}%
+15. Hanks B, Fitzgerald S, McCauley R, Murphy L, Zander C. Pair
+programming in education: A literature review. Computer Science
+Education. 2011;21: 135--173.
+doi:\href{https://doi.org/10.1080/08993408.2011.579808}{10.1080/08993408.2011.579808}
+
 \leavevmode\hypertarget{ref-Bockeler_2020}{}%
-6. Böckeler B, Siessegger N. On pair programming. 2020. Available:
+16. Böckeler B, Siessegger N. On pair programming. 2020. Available:
 \url{https://martinfowler.com/articles/on-pair-programming.html}
@@
 \leavevmode\hypertarget{ref-Wickham_2016}{}%
-7. Wickham H, Grolemund G. O'Reilly Media; 2016. Available:
+17. Wickham H, Grolemund G. O'Reilly Media; 2016. Available:
 \url{https://r4ds.had.co.nz}
@@
 \leavevmode\hypertarget{ref-Schloss_2018}{}%
-8. Schloss PD. The riffomonas reproducible research tutorial series.
+18. Schloss PD. The Riffomonas reproducible research tutorial series.
 Journal of Open Source Education. 2018;1: 13.
@@
 \leavevmode\hypertarget{ref-Noble2009}{}%
-9. Noble WS. A quick guide to organizing computational biology projects.
-Lewitter F, editor. PLoS Computational Biology. 2009;5: e1000424.
+19. Noble WS. A quick guide to organizing computational biology
+projects. Lewitter F, editor. PLoS Computational Biology. 2009;5:
+e1000424.
 doi:\href{https://doi.org/10.1371/journal.pcbi.1000424}{10.1371/journal.pcbi.1000424}
@@
 Given a bash script that contains an analysis pipeline, convert it to a
-Snakemake workflow (can also be done with GNU Make)\strut
+Snakemake workflow
+(\url{https://github.com/SchlossLab/snakemake_riffomonas_tutorial})\strut
 \end{minipage}\tabularnewline
@@
 \begin{minipage}[t]{0.25\columnwidth}\raggedright
-mothur and Vegan\strut
+Vegan R package\strut
 \end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
-Given a pairwise community dissimilarity matrix, compare communities
-using the \texttt{adonis} function in the Vegan R package\strut
+Compare microbial communities using the \texttt{adonis} function in the
+Vegan R package
+(\url{https://github.com/SchlossLab/Code_Review_42717})\strut
 \end{minipage}\tabularnewline
@@
 \begin{minipage}[t]{0.25\columnwidth}\raggedright
-R with Google docs\strut
+googledocs4 R package\strut
 \end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
@@
 \end{minipage}\tabularnewline
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
+Documenting R code\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+Use roxygen2 to supplement comments in R code to improve documentation
+(\url{https://github.com/SchlossLab/documenting-R})\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.25\columnwidth}\raggedright
+gganimate R package\strut
+\end{minipage} & \begin{minipage}[t]{0.69\columnwidth}\raggedright
+Convert static plots generated with ggplot2 package into GIFs
+(\url{https://github.com/SchlossLab/2020-04-12-CodeClub_PlotAnimation/})\strut
+\end{minipage}\tabularnewline
 \bottomrule

### 2020-06-19 submission/manuscript.Rmd
small edits

@@
 A common strategy for keeping up with the literature is by participating in a journal club.
-Regardless of the discipline, a Journal Club involves group discussion of a pre-selected paper.
-These range from informal discussions to PowerPoint presentations and course credit.
+Regardless of the discipline, a Journal Club involves group discussion of a pre-selected paper that range from informal discussions to PowerPoint presentations and course credit.
 In addition to building upon material from traditional coursework and staying current on the literature, journal clubs help strengthen skills in critical thinking, communication, and integrating the literature [@Lonsdale2016].
-Since most journal clubs occur on a regular schedule, they are effective because they incorporate repeated deliberate practices.
+Since most journal clubs occur on a regular schedule, they are effective by virtue of repeated practice.
 With this model in mind, over the past four years we have experimented with creating a Code Club model with the goal of improving reproducible data analysis skills in a community environment.
@@
 For example, at one Code Club participants were given raw data and a finished plot.
-Paired participants were tasked with generating the plot from the data using either R syntax from the base language or the ggplot2 package.
+Paired participants were tasked with generating the plot from the data using R syntax from either the base language or the ggplot2 package.
 For this exercise, base R users had to use ggplot2 and vice versa.
 In either approach, the Code Club ends with a report back to the larger group describing the approach each pair took.
-We generally find that preparing a Code Club session takes as much effort as preparing to present at a journal club.
+We generally find that preparing a Code Club session takes similar effort to preparing a journal club presentation.
 Our Code Clubs typically have 7 to 10 participants, but the inherent "think-pair-share" approach should allow it to be scaled to larger groups [@Kothiyal2014].
@@
 We continue to experiment with approaches for running Code Club.
-For example, during the period of the Covid-19 pandemic in 2020 when many research labs were shuttered, PDS started posting Code Club materials (https://www.riffomonas.org/code_club/) that individuals or research groups can use to practice their coding skills.
-During this time, we continued our group's Code Club sessions by making use of the breakout room and screen sharing features in Zoom.
+For example, during the early period of the Covid-19 pandemic in 2020 when many research labs were shuttered, PDS started posting Code Club materials (https://www.riffomonas.org/code_club/) that individuals or research groups can use to practice their coding skills.
+During this time, we continued our group's Code Club sessions using the breakout room and screen sharing features in Zoom.
 Regardless of our approach to leading a Code Club, we have learned that it is critical for the presenter to clearly articulate their goals and facilitate participant engagement.
@@
 The first 3 rules apply to all participants, Rules 4 through 8 are targeted to presenters, and the last two for non-presenters.
-As the rules describe, this model can easily be adapted to groups of non-biologists and to groups with stronger background in computer science.
+As the rules describe, this model can easily be adapted to groups of non-biologists or those with stronger backgrounds in computer science.
 
@@
 Some tips to help keep it simple are to limit the presented code chunks to less than 50 lines or, conversely, consider the number of lines that might be required to accomplish a solution.
-Remember that learners may need up to three times as long to complete a task that is simple for the presenter, so Code Club is best kept simple.
+Remember that learners may need up to three times as long to complete a task that is straight forward for the presenter, so Code Club is best kept simple.
 
@@
 We have found that using GitHub repositories for each Code Club can help make information, scripts, and data easily available to participants.
-Although this can be convenient, introducing GitHub on top of the session's activities can impose a significant cognitive load and frustration to new group members as well as to the presenter if they are not comfortable with GitHub.
+Although this can be convenient, introducing GitHub on top of the session's activities can impose a significant cognitive load and frustration to those not already comfortable with GitHub.
 Perhaps a first Code Club could introduce using git and GitHub to engage in collaborative coding.
@@
 Beginners will benefit from sessions that cover fundamental concepts and functions.
-The benefits experienced by more experienced participants include the improved understanding of concepts that comes through teaching the concepts and breaking down problems into simpler elements.
-Furthermore, instead of focusing on core functions, the group could balance Code Club sessions that cover basics with those that cover methods and packages that are new to everyone in the group.
+The benefits for more experienced participants include the improved understanding of concepts by teaching and breaking down problems into simpler elements.
+Furthermore, instead of focusing on core functions, the group could balance Code Club sessions that cover basics with those that introduce new methods and packages to the group.
 We have also identified several strategies to overcome the challenges presented by participants at varying skill levels.
@@
 This speaks to both the broader need for Code Club and the likelihood of success when expanded to include a larger group of individuals with broader research interests.
-There is no reason that the Code Club format would not work with a group of people independent of a research group as long as the follow the Rules.
-Finding datasets and applications that are interesting to a critical mass of people is likely critical to starting and sustaining such a group.
+There is no reason that the Code Club format would not work with groups independent of a research group as long as they follow the Rules.
+Finding datasets and applications that are interesting to a critical mass of people will likely be key to starting and sustaining such a group.
 Ultimately, Code Club has improved the overall data analysis skills, community, and research success of our lab by empowering researchers to seek help from their colleagues.

### 2020-06-19 submission/manuscript.tex
Render with Ada's comments

@@
 Regardless of the discipline, a Journal Club involves group discussion
-of a pre-selected paper. These range from informal discussions to
+of a pre-selected paper that range from informal discussions to
 PowerPoint presentations and course credit. In addition to building upon
@@
 communication, and integrating the literature {[}11{]}. Since most
-journal clubs occur on a regular schedule, they are effective because
-they incorporate repeated deliberate practices. With this model in mind,
-over the past four years we have experimented with creating a Code Club
-model with the goal of improving reproducible data analysis skills in a
-community environment.
+journal clubs occur on a regular schedule, they are effective by virtue
+of repeated practice. With this model in mind, over the past four years
+we have experimented with creating a Code Club model with the goal of
+improving reproducible data analysis skills in a community environment.
 
@@
 and a finished plot. Paired participants were tasked with generating the
-plot from the data using either R syntax from the base language or the
+plot from the data using R syntax from either the base language or the
 ggplot2 package. For this exercise, base R users had to use ggplot2 and
@@
 the larger group describing the approach each pair took. We generally
-find that preparing a Code Club session takes as much effort as
-preparing to present at a journal club. Our Code Clubs typically have 7
+find that preparing a Code Club session takes similar effort to
+preparing a journal club presentation. Our Code Clubs typically have 7
 to 10 participants, but the inherent ``think-pair-share'' approach
@@
 We continue to experiment with approaches for running Code Club. For
-example, during the period of the Covid-19 pandemic in 2020 when many
-research labs were shuttered, PDS started posting Code Club materials
-(\url{https://www.riffomonas.org/code_club/}) that individuals or
-research groups can use to practice their coding skills. During this
-time, we continued our group's Code Club sessions by making use of the
-breakout room and screen sharing features in Zoom. Regardless of our
-approach to leading a Code Club, we have learned that it is critical for
-the presenter to clearly articulate their goals and facilitate
-participant engagement. Although some Code Club sessions may be more
-experimental than others, on the whole they are a critical tool to train
-bench scientists in reproducible data analysis practices. We have
-provided some examples of successful topics in Table 1. We have
-summarized the results of our own experimentation as Code Club
-presenters and participants into Ten Rules. The first 3 rules apply to
-all participants, Rules 4 through 8 are targeted to presenters, and the
-last two for non-presenters. As the rules describe, this model can
-easily be adapted to groups of non-biologists and to groups with
-stronger background in computer science.
+example, during the early period of the Covid-19 pandemic in 2020 when
+many research labs were shuttered, PDS started posting Code Club
+materials (\url{https://www.riffomonas.org/code_club/}) that individuals
+or research groups can use to practice their coding skills. During this
+time, we continued our group's Code Club sessions using the breakout
+room and screen sharing features in Zoom. Regardless of our approach to
+leading a Code Club, we have learned that it is critical for the
+presenter to clearly articulate their goals and facilitate participant
+engagement. Although some Code Club sessions may be more experimental
+than others, on the whole they are a critical tool to train bench
+scientists in reproducible data analysis practices. We have provided
+some examples of successful topics in Table 1. We have summarized the
+results of our own experimentation as Code Club presenters and
+participants into Ten Rules. The first 3 rules apply to all
+participants, Rules 4 through 8 are targeted to presenters, and the last
+two for non-presenters. As the rules describe, this model can easily be
+adapted to groups of non-biologists or those with stronger backgrounds
+in computer science.
 
@@
 required to accomplish a solution. Remember that learners may need up to
-three times as long to complete a task that is simple for the presenter,
-so Code Club is best kept simple.
+three times as long to complete a task that is straight forward for the
+presenter, so Code Club is best kept simple.
 
@@
 session's activities can impose a significant cognitive load and
-frustration to new group members as well as to the presenter if they are
-not comfortable with GitHub. Perhaps a first Code Club could introduce
-using git and GitHub to engage in collaborative coding. A lower barrier
-entry point for the presenter is posting their code, data, and
-information in a lab meeting-dedicated Slack channel or via email.
-Whatever method is used, the presenter should be sure to communicate the
-topic and necessary materials with the participants ahead of time.
+frustration to those not already comfortable with GitHub. Perhaps a
+first Code Club could introduce using git and GitHub to engage in
+collaborative coding. A lower barrier entry point for the presenter is
+posting their code, data, and information in a lab meeting-dedicated
+Slack channel or via email. Whatever method is used, the presenter
+should be sure to communicate the topic and necessary materials with the
+participants ahead of time.
 
@@
 membership. Beginners will benefit from sessions that cover fundamental
-concepts and functions. The benefits experienced by more experienced
-participants include the improved understanding of concepts that comes
-through teaching the concepts and breaking down problems into simpler
-elements. Furthermore, instead of focusing on core functions, the group
-could balance Code Club sessions that cover basics with those that cover
-methods and packages that are new to everyone in the group. We have also
-identified several strategies to overcome the challenges presented by
-participants at varying skill levels. Central to the Code Club format is
-the use of paired-programming {[}15{]}. Instead of letting participants
-form their own pairs, the presenter can select pairs of participants
-with either similar or differing skill levels, depending on the
-presenter's goals. Partnerships between those with similar skill levels
-requires the presenter to design appropriate activities for each skill
-level. We have found that commenting code is a good skill for beginners
-since it forces them to dissect and understand a code chunk
-line-by-line. It also reinforces the value of commenting as they develop
-their skills and independence. An advantage of forming partnership
-between people with disparate skill levels is that it is more likely for
-groups to provide the presenter with a diverse range of methods that
-achieve the same result. This approach to pairing also helps to graft
-new members that have emerging programming skills. Regardless of how
-partners are selected, consider asking the pairs to identify a navigator
-and a driver {[}16{]}. The driver types at the computer while the
-navigator tells them what to type, thus ensuring participation of both
-partners. Midway through the activity, the presenter can have the
-partners switch roles. Intentionally forming pairs can also engineer
-group interactions by avoiding potentially disruptive partnerships or
-pairing reliable role models with new group members.
+concepts and functions. The benefits for more experienced participants
+include the improved understanding of concepts by teaching and breaking
+down problems into simpler elements. Furthermore, instead of focusing on
+core functions, the group could balance Code Club sessions that cover
+basics with those that introduce new methods and packages to the group.
+We have also identified several strategies to overcome the challenges
+presented by participants at varying skill levels. Central to the Code
+Club format is the use of paired-programming {[}15{]}. Instead of
+letting participants form their own pairs, the presenter can select
+pairs of participants with either similar or differing skill levels,
+depending on the presenter's goals. Partnerships between those with
+similar skill levels requires the presenter to design appropriate
+activities for each skill level. We have found that commenting code is a
+good skill for beginners since it forces them to dissect and understand
+a code chunk line-by-line. It also reinforces the value of commenting as
+they develop their skills and independence. An advantage of forming
+partnership between people with disparate skill levels is that it is
+more likely for groups to provide the presenter with a diverse range of
+methods that achieve the same result. This approach to pairing also
+helps to graft new members that have emerging programming skills.
+Regardless of how partners are selected, consider asking the pairs to
+identify a navigator and a driver {[}16{]}. The driver types at the
+computer while the navigator tells them what to type, thus ensuring
+participation of both partners. Midway through the activity, the
+presenter can have the partners switch roles. Intentionally forming
+pairs can also engineer group interactions by avoiding potentially
+disruptive partnerships or pairing reliable role models with new group
+members.
 
@@
 larger group of individuals with broader research interests. There is no
-reason that the Code Club format would not work with a group of people
-independent of a research group as long as the follow the Rules. Finding
-datasets and applications that are interesting to a critical mass of
-people is likely critical to starting and sustaining such a group.
-Ultimately, Code Club has improved the overall data analysis skills,
-community, and research success of our lab by empowering researchers to
-seek help from their colleagues.
+reason that the Code Club format would not work with groups independent
+of a research group as long as they follow the Rules. Finding datasets
+and applications that are interesting to a critical mass of people will
+likely be key to starting and sustaining such a group. Ultimately, Code
+Club has improved the overall data analysis skills, community, and
+research success of our lab by empowering researchers to seek help from
+their colleagues.
 

### 2020-06-25 submission/response_to_reviewers.md
Incorporate edits

@@
 Throughout their review, the reviewer encourages us to make more comments about groups where there are more people with CS backgrounds.
-Admittedly, our lab is made of researchers with non-CS backgrounds and so our implementation of Code Clubs is with that audience in mind.
+Admittedly, our lab is mostly made of researchers with non-CS backgrounds and so our implementation of Code Clubs is with that audience in mind.
 We have added text as appropriate to respond to the following suggestions.
+We have expanded our listing of example code clubs in Table 1 along with links to GitHub repositories where they were used.
+Examples of Code Clubs that did not work are highlighted throughout the manuscript (e.g.
+teaching Julia or not clearly stating goals).
 

### 2020-06-25 submission/manuscript.Rmd
Incorporate edits

@@
 These learning formats ask participants to engage in massed practice; an approach that is known to be ineffective [@Carpenter2012; @Rohrer2015b; @Bud2010; @Karpicke2011].
-In contrast, participants need regular opportunities to engage in repeated deliberate practice that encourage them to practice their retrieval and application skills of the material [@Rohrer2015a; @Karpicke2008].
-There is clearly a need to develop a framework that can build upon material introduced in coursework and workshops and that can help researchers to stay up to date on the latest best practices, algorithms, and tools.
+In contrast, participants need regular opportunities to engage in repeated deliberate practice that encourage them to practice their retrieval and application of the material [@Rohrer2015a; @Karpicke2008].
+There is clearly a need to develop a framework that can build upon material introduced in coursework and workshops and that can help researchers stay up to date on the latest best practices, algorithms, and tools.
 
 To address this need, we noted the similarity between the feeling of unease in analyzing the data and the struggles scientists face with engaging the voluminous scientific literature.
-A common strategy for keeping up with the literature is by participating in a journal club.
+A common strategy for keeping up with the literature is to participate in a Journal Club.
 Regardless of the discipline, a Journal Club involves group discussion of a pre-selected paper that range from informal discussions to PowerPoint presentations and course credit.
-In addition to building upon material from traditional coursework and staying current on the literature, journal clubs help strengthen skills in critical thinking, communication, and integrating the literature [@Lonsdale2016].
-Since most journal clubs occur on a regular schedule, they are effective by virtue of repeated practice.
-With this model in mind, over the past four years we have experimented with creating a Code Club model with the goal of improving reproducible data analysis skills in a community environment.
+In addition to building upon material from traditional coursework and staying current on the literature, Journal Clubs help strengthen skills in critical thinking, communication, and integrating the literature [@Lonsdale2016].
+Since most Journal Clubs occur on a regular schedule, they are effective by virtue of repeated practice.
+With this model in mind, over the past four years we have experimented with creating a Code Club model with the goal of improving reproducible data analysis skills in a laboratory environment.
 
@@
 In either approach, the Code Club ends with a report back to the larger group describing the approach each pair took.
-We generally find that preparing a Code Club session takes similar effort to preparing a journal club presentation.
-Our Code Clubs typically have 7 to 10 participants, but the inherent "think-pair-share" approach should allow it to be scaled to larger groups [@Kothiyal2014].
+We generally find that preparing a Code Club session takes similar effort to preparing a Journal Club presentation.
+Our Code Clubs typically have 7 to 10 participants, but the inherent "think-pair-share" approach should allow it to be scaled to groups of variable sizes [@Kothiyal2014].
 
 We continue to experiment with approaches for running Code Club.
-For example, during the early period of the Covid-19 pandemic in 2020 when many research labs were shuttered, PDS started posting Code Club materials (https://www.riffomonas.org/code_club/) that individuals or research groups can use to practice their coding skills.
+For example, during the COVID-19 pandemic in 2020 when many research labs were shuttered, PDS started posting Code Club materials (https://www.riffomonas.org/code_club/) that individuals or research groups can use to practice their coding skills.
 During this time, we continued our group's Code Club sessions using the breakout room and screen sharing features in Zoom.
@@
 The benefits for more experienced participants include the improved understanding of concepts by teaching and breaking down problems into simpler elements.
-Furthermore, instead of focusing on core functions, the group could balance Code Club sessions that cover basics with those that introduce new methods and packages to the group.
+Furthermore, they can benefit by contributing to an environment of learning that they previously benefited from.
+In addition, instead of focusing on core functions, the group could balance Code Club sessions that cover basics with those that introduce new methods and packages to the group.
 We have also identified several strategies to overcome the challenges presented by participants at varying skill levels.
@@
 Aside from expanding our programming skills, we have noticed two other benefits that help create a positive culture.
-First, we have intentionally interviewed postdoc candidates on the days we hold Code Club.
+First, we have intentionally interviewed postdoctoral candidates on the days we hold Code Club.
 We make it clear that they are not being assessed on their coding skills, but instead use it as an opportunity to see how the candidate interacts with other members of the research group.
@@
 This speaks to both the broader need for Code Club and the likelihood of success when expanded to include a larger group of individuals with broader research interests.
-There is no reason that the Code Club format would not work with groups independent of a research group as long as they follow the Rules.
-Finding datasets and applications that are interesting to a critical mass of people will likely be key to starting and sustaining such a group.
+There is no reason that the Code Club format would not work with groups independent of a research group as long as everyone follows the Rules.
+Finding datasets and applications that are interesting to a critical mass of people is essential to starting and sustaining such a group.
 Ultimately, Code Club has improved the overall data analysis skills, community, and research success of our lab by empowering researchers to seek help from their colleagues.

### 2020-06-25 submission/manuscript.tex
Incorporate edits

@@
 engage in repeated deliberate practice that encourage them to practice
-their retrieval and application skills of the material {[}9,10{]}. There
-is clearly a need to develop a framework that can build upon material
-introduced in coursework and workshops and that can help researchers to
+their retrieval and application of the material {[}9,10{]}. There is
+clearly a need to develop a framework that can build upon material
+introduced in coursework and workshops and that can help researchers
 stay up to date on the latest best practices, algorithms, and tools.
@@
 engaging the voluminous scientific literature. A common strategy for
-keeping up with the literature is by participating in a journal club.
+keeping up with the literature is to participate in a Journal Club.
 Regardless of the discipline, a Journal Club involves group discussion
@@
 material from traditional coursework and staying current on the
-literature, journal clubs help strengthen skills in critical thinking,
+literature, Journal Clubs help strengthen skills in critical thinking,
 communication, and integrating the literature {[}11{]}. Since most
-journal clubs occur on a regular schedule, they are effective by virtue
+Journal Clubs occur on a regular schedule, they are effective by virtue
 of repeated practice. With this model in mind, over the past four years
 we have experimented with creating a Code Club model with the goal of
-improving reproducible data analysis skills in a community environment.
+improving reproducible data analysis skills in a laboratory environment.
 
@@
 find that preparing a Code Club session takes similar effort to
-preparing a journal club presentation. Our Code Clubs typically have 7
+preparing a Journal Club presentation. Our Code Clubs typically have 7
 to 10 participants, but the inherent ``think-pair-share'' approach
-should allow it to be scaled to larger groups {[}13{]}.
+should allow it to be scaled to groups of variable sizes {[}13{]}.
 
 We continue to experiment with approaches for running Code Club. For
-example, during the early period of the Covid-19 pandemic in 2020 when
-many research labs were shuttered, PDS started posting Code Club
-materials (\url{https://www.riffomonas.org/code_club/}) that individuals
-or research groups can use to practice their coding skills. During this
+example, during the COVID-19 pandemic in 2020 when many research labs
+were shuttered, PDS started posting Code Club materials
+(\url{https://www.riffomonas.org/code_club/}) that individuals or
+research groups can use to practice their coding skills. During this
 time, we continued our group's Code Club sessions using the breakout
@@
 include the improved understanding of concepts by teaching and breaking
-down problems into simpler elements. Furthermore, instead of focusing on
-core functions, the group could balance Code Club sessions that cover
-basics with those that introduce new methods and packages to the group.
-We have also identified several strategies to overcome the challenges
-presented by participants at varying skill levels. Central to the Code
-Club format is the use of paired-programming {[}15{]}. Instead of
-letting participants form their own pairs, the presenter can select
-pairs of participants with either similar or differing skill levels,
-depending on the presenter's goals. Partnerships between those with
-similar skill levels requires the presenter to design appropriate
-activities for each skill level. We have found that commenting code is a
-good skill for beginners since it forces them to dissect and understand
-a code chunk line-by-line. It also reinforces the value of commenting as
-they develop their skills and independence. An advantage of forming
-partnership between people with disparate skill levels is that it is
-more likely for groups to provide the presenter with a diverse range of
-methods that achieve the same result. This approach to pairing also
-helps to graft new members that have emerging programming skills.
-Regardless of how partners are selected, consider asking the pairs to
-identify a navigator and a driver {[}16{]}. The driver types at the
-computer while the navigator tells them what to type, thus ensuring
-participation of both partners. Midway through the activity, the
-presenter can have the partners switch roles. Intentionally forming
-pairs can also engineer group interactions by avoiding potentially
-disruptive partnerships or pairing reliable role models with new group
-members.
+down problems into simpler elements. Furthermore, they can benefit by
+contributing to an environment of learning that they previously
+benefited from. In addition, instead of focusing on core functions, the
+group could balance Code Club sessions that cover basics with those that
+introduce new methods and packages to the group. We have also identified
+several strategies to overcome the challenges presented by participants
+at varying skill levels. Central to the Code Club format is the use of
+paired-programming {[}15{]}. Instead of letting participants form their
+own pairs, the presenter can select pairs of participants with either
+similar or differing skill levels, depending on the presenter's goals.
+Partnerships between those with similar skill levels requires the
+presenter to design appropriate activities for each skill level. We have
+found that commenting code is a good skill for beginners since it forces
+them to dissect and understand a code chunk line-by-line. It also
+reinforces the value of commenting as they develop their skills and
+independence. An advantage of forming partnership between people with
+disparate skill levels is that it is more likely for groups to provide
+the presenter with a diverse range of methods that achieve the same
+result. This approach to pairing also helps to graft new members that
+have emerging programming skills. Regardless of how partners are
+selected, consider asking the pairs to identify a navigator and a driver
+{[}16{]}. The driver types at the computer while the navigator tells
+them what to type, thus ensuring participation of both partners. Midway
+through the activity, the presenter can have the partners switch roles.
+Intentionally forming pairs can also engineer group interactions by
+avoiding potentially disruptive partnerships or pairing reliable role
+models with new group members.
 
@@
 programming skills, we have noticed two other benefits that help create
-a positive culture. First, we have intentionally interviewed postdoc
-candidates on the days we hold Code Club. We make it clear that they are
-not being assessed on their coding skills, but instead use it as an
-opportunity to see how the candidate interacts with other members of the
-research group. At the same time, the candidate can learn about the
-culture of the research group through active participation. Second,
-members of other research groups have integrated themselves into our
-Code Club to minimize the isolation they feel in growing their skills
-within smaller research groups. This speaks to both the broader need for
-Code Club and the likelihood of success when expanded to include a
-larger group of individuals with broader research interests. There is no
-reason that the Code Club format would not work with groups independent
-of a research group as long as they follow the Rules. Finding datasets
-and applications that are interesting to a critical mass of people will
-likely be key to starting and sustaining such a group. Ultimately, Code
-Club has improved the overall data analysis skills, community, and
-research success of our lab by empowering researchers to seek help from
-their colleagues.
+a positive culture. First, we have intentionally interviewed
+postdoctoral candidates on the days we hold Code Club. We make it clear
+that they are not being assessed on their coding skills, but instead use
+it as an opportunity to see how the candidate interacts with other
+members of the research group. At the same time, the candidate can learn
+about the culture of the research group through active participation.
+Second, members of other research groups have integrated themselves into
+our Code Club to minimize the isolation they feel in growing their
+skills within smaller research groups. This speaks to both the broader
+need for Code Club and the likelihood of success when expanded to
+include a larger group of individuals with broader research interests.
+There is no reason that the Code Club format would not work with groups
+independent of a research group as long as everyone follows the Rules.
+Finding datasets and applications that are interesting to a critical
+mass of people is essential to starting and sustaining such a group.
+Ultimately, Code Club has improved the overall data analysis skills,
+community, and research success of our lab by empowering researchers to
+seek help from their colleagues.
 
